### 2013-12-24 src/docs/papers/stan-language/article.tex
trimmed abstract, added Andrew paragraph on other applications

\documentclass[article]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{{\bf\large Bob Carpenter}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Daniel Lee}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Marcus A.\ Brubaker}
        \\ TTI-Chicago
        %
    \And
        %
        {\bf\large Andrew Gelman}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Ben Goodrich}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Jiqiang Guo}
        \\ Columbia Univesity
        %
     \And
        %
        {\bf\large Matt Hoffman}
        \\ Adobe Research Labs
        %
        \\[9pt]
        %
        {\bf\large Michael Betancourt}
        \\ University College London
        %
        \\[9pt]
        %
        {\bf\large Peter Li}
        \\ Columbia University
}
\title{\proglang{Stan}: A Probabilistic Programming Language}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Bob Carpenter, Andrew Gelman, Matt Hoffman, 
  Daniel Lee, Ben Goodrich, Michael Betancourt, 
  Marcus Brubaker, Jiqiang Guo, Peter Li} %% comma-separated
\Plaintitle{Stan: A Probabilistic Programming Language} %% without formatting
\Shorttitle{Stan: A Probabilistic Programming Language} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{ \proglang{Stan} is a probabilistic programming language for
  specifying statistical models. A \proglang{Stan} program
  imperatively defines a log probability function over parameters
  conditioned on specified data and constants.

  Full Bayesian inference is supported through Markov chain Monte
  Carlo (MCMC) methods such as the No-U-Turn sampler (NUTS), an
  adaptive form of Hamiltonian Monte Carlo (HMC) sampling.  Penalized
  maximum likelihood estimates are calculated using optimization
  methods such as BFGS.

  % This may be contrasted
  % with the \proglang{BUGS} language, in which a program
  % declaratively
  % defines a directed acyclic graphical model.

  % Variables are declared by dimensionality and as to whether they
  % represent data, transformed data, parameters, transformed
  % parameters, or generated quantities; local variables are also
  % supported.  In addition to unconstrained and constrained scalar,
  % discrete, vector, matrix, and array types, there is a library of
  % mathematical functions, probability-related functions, and matrix
  % and linear algebra functions.  Statements, such as assignment,
  % sampling, conditionals, and loops are executed imperatively in the
  % order they are written.

  % \proglang{Stan} programs are translated to \proglang{C++} and
  % compiled.  The target \proglang{C++} code includes forward- and
  % reverse-mode algorithmic differentiation to support samplers and
  % optimizers requiring gradients, Hessians, Hessian-vector products,
  % and other higher-order derivatives.

  % There is an extensive I/O
  % library to deal with constrained variable input for data,
  % initialization for parameters, and output for samples or point
  % estimates.  There are also a range of tools to monitor convergence
  % and mixing, summarize the posterior, perform Bayesian inference,
  % and
  % carry out posterior predictive checks.

  Stan can be viewed as a package for performing full Bayesian
  inference for continuous-variable models, and as a platform for
  computing log-densities and their gradients and Hessians, which can
  be used in alternative algorithms such as variational Bayes,
  expectation propagation, and marginal inference using approximate
  integration.  To this end, Stan is set up so that the densities,
  gradients, and Hessians, along with intermediate quantities of the
  algorithm such as acceptance probabilities, are easily accessible.

  % \proglang{Stan} can be called from the command line, through
  % \proglang{R} using the \pkg{RStan} package, or through
  % \proglang{Python} using the \pkg{PyStan} package.  All three
  % interfaces support sampling or optimization-based inference and
  % analysis.
}

\Keywords{
  probabilistic program,
  Bayesian inference,
  algorithmic differentiation,
  \proglang{Stan}}
%
\Plainkeywords{
  probabilistic programming,
  Bayesian inference,
  algorithmic differentiation,
  Stan}

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Bob Carpenter
  \\
  Department of Statistics
  \\ 
  Columbia University
  \\ 
  1255 Amsterdam Avenue
  \\ 
  New York, NY 10027
  \\
  U.S.A.
  \\
  E-mail: \email{carp@stat.columbia.edu}
  \\ 
  URL: \url{http://mc-stan.org/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\section{TO-DO}

\begin{enumerate}
\item Convergence diagnostics?
\item more model examples?
\item description from C++ perspective?
\item what else do we want to make sure users don't miss?
\item optimization example?
\item future plans section?
\end{enumerate}

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\section[Why Stan?]{Why \proglang{Stan}?}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

We did not set out to build \proglang{Stan} as it currently exists.
Our original goal was to apply full Bayesian inference to the sort of multilevel
generalized linear models discussed in Part~II of
\citep{GelmanHill:2007}, which are structured with grouped and
interacted predictors at multiple levels, hierarchical covariance
priors, nonconjugate coefficient priors, latent effects as in
item-response models, and varying output link functions and
distributions.

These models turned out to be a challenge for
existing general-purpose inference software.  A direct encoding in
\proglang{BUGS} \citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
\proglang{JAGS} \citep{Plummer:2003} can grind these tools to a halt.
We began by adding custom vectorized logistic regressions to \proglang{JAGS}
using \proglang{C++} to overcome the cost of interpretation.  Although
this is much faster than looping in \proglang{JAGS}, it quickly became
clear that the root of the problem was the slow convergence of
conditional sampling when parameters were highly correlated 
in the posterior, as for time-series models and hierarchical models 
with interacted predictors.  We finally realized we needed a better sampler, 
not a more efficient implementation.

We briefly considered trying to tune proposals for a random-walk
Metropolis-Hastings sampler, but that seemed too problem specific and
not even necessarily possible without some kind of adaptation rather
than a global tuning of the proposals.

We were hearing more about HMC (Hamiltonian Monte Carlo), which
appeared promising but was also problematic in that the Hamiltonian
dynamics simulation requires the gradient of the log posterior.
Although possible, computing these gradients by hand on a model-by-model basis
is very tedious and error prone.  That is when we discovered
reverse-mode algorithmic differentiation, which, given a
templated \proglang{C++} function for the log posterior,
automatically computes a proper analytic gradient up to machine
precision accuracy in only a few multiples of the cost to evaluate the
log probability function itself.  We explored existing algorithmic
differentiation packages with open licenses such as \pkg{RAD}
\citep{Gay:2005} and its repackaging in the Sacado module of the
\pkg{Trilinos} toolkit, and the \pkg{CppAD} package of the COIN-OR
toolkit (see \citep{BellBurke:2008}).  Both packages, however, supported
few special functions (e.g., probability functions, log gamma,
inverse logit) or linear algebra operations (e.g., Cholesky
decompositions, matrix division), and neither are easily or modularly
extensible.

Consequently we built our own reverse-mode algorithmic differentiation package.
At that point, we ran into the problem that we could not just plug in
the probability functions from a package like \pkg{Boost} because they
weren't templated generally enough across all the arguments.  Rather
than pay the price of promoting floating point values to algorithmic
differentiation variables, we wrote our own fully templated
probability functions and other special functions.

Next, we integrated the \proglang{C++} package \pkg{Eigen} for matrix
operations and linear algebra functions.  \pkg{Eigen} makes extensive
use of expression templates for lazy evaluation and the CRTP
(curiously recurring template pattern) to implement concepts without
virtual function calls; \cite{VandevoordeJosuttis:2002} provide a 
complete description of template metaprogramming techniques. 
Unfortunately, we
ran into the same problem with Eigen as with the existing probability
libraries---it doesn't support mixed operations of algorithmic
differentiation variables and primitives like \code{double}.  Although
we initially began by promoting floating-point vectors to algorithmic
differentiation variables, we later completely rewrote our matrix
library to efficiently compute derivatives using rules described by
\cite{Giles:2008}.

At this point, we could fit models coded directly in \proglang{C++} on
top of the pre-release versions of the \proglang{Stan} API
(application programming interface).  Seeing how well this all worked,
we set our sights on the generality and ease of use of
\proglang{BUGS} and designed a modeling language in which
statisticians could write their models in familiar notation that could
be transformed to efficient \proglang{C++} code before being compiled into
an efficient executable program.  This paper is primarily about this
language and how it specifies models.

The next problem we ran into when we started implementing richer
models was variables with constrained support, such as positive variables, 
simplexes, and covariance matrices.  
Efficient implementation of these constraints
required the introduction of typed variables which automatically transformed to
unconstrained support with suitable adjustments to the log probability
from the log absolute Jacobian determinant of the inverse transforms.

Even with the prototype compiler generating models, we still faced a
major hurdle for ease of use. The efficiency of HMC is very sensitive
to two tuning parameters, the discretization interval (i.e., step
size) and the total simulation time (i.e., number of steps).  The interval
size parameter could be tuned during warm-up based on Metropolis
rejection rates, but the number of steps proved difficult to tune without
sacrificing the detailed balance of the sampler.  This led to the
development of the NUTS (No U-Turn) sampler
\citep{HoffmanGelman:2011}.  Roughly speaking, NUTS builds a tree
of possible samples by randomly simulating both forwards and backwards
in time until the combined trajectory turns back on itself.  Once the trajectory
has terminated, a new sample is drawn from the tree.

We thought we were home free at this point, but when we measured the
speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
very disappointed.  \proglang{BUGS}'s very first example model, Rats,
ran more than an order of magnitude faster in \proglang{JAGS} than in
\proglang{Stan}.  Rats is a tough test case because the conjugate
priors and lack of posterior correlations make it an ideal candidate
for efficient Gibbs sampling.

Realizing that we were doing redundant calculations, we wrote a
vectorized form of the normal distribution for multiple variates with
the same mean and scale, which sped things up a bit. At the same time,
we introduced some simple template metaprograms to remove the
calculation of constant terms in the log probability.  Performance was
lacking until we finally figured out how to both eliminate redundant 
calculations and partially evaluate the gradients using
a combination of expression templates and metaprogramming.

When we attempted to fit a time-series model, we found that
normalizing the data to unit sample mean and variance sped up the fits
by an order of magnitude.  In hindsight this isn't surprising, as the 
performance of the numerical simulation in HMC scales with the variation 
of the parameter scales; fortunately HMC also provides a choice of metric,
or mass matrix, to compensate for variations in the log posterior.
In \proglang{Stan} 1.0 we introduced an adaptive 
diagonal metric into our HMC implementations that allowed the parameter 
scales to be normalized automatically, with the option of an adaptive dense 
metric and a full automated standardization of the posterior in \proglang{Stan} 2.0.

Because the dense metric performs only a global standardization, the
performance of \proglang{Stan} in models where the relative parameter
scales varies by location, such as hierarchical and latent models, can
suffer.  Riemannian manifold Hamiltonian Monte Carlo (RHMC)
\citep{GirolamiCalderhead:2011} introduces a location-dependent metric
that can overcome these final hurdles, and a full implementation
within Stan is currently under development.

\section{Overview}

This section describes the use of \proglang{Stan} from the command
line for estimating (i.e., sampling from) the posterior of a Bayesian
model.

\subsection{Model for estimating a Bernoulli parameter}

Consider estimating the chance of success parameter for a Bernoulli
distribution based on a sequence of observed binary outcomes.  
Figure~\ref{bernoulli-model.fig} provides an implementation of such a
model in \proglang{Stan}.
%
\begin{figure}
\begin{Code}
data { 
  int<lower=0> N; 
  int<lower=0,upper=1> y[N];
} 
parameters {
  real<lower=0,upper=1> theta;
} 
model {
  theta ~ beta(1,1);  // prior
  for (n in 1:N) 
    y[n] ~ bernoulli(theta);  // likelihood
}
\end{Code}
\caption{Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
\end{figure}
%
The model treats the observed binary data, \code{y[1],...,y[N]}, as
independent and identically distributed, with success probability \code{theta}.  
A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta} , although there is no
special behavior for conjugate priors in \proglang{Stan}.  This model
is available in the \proglang{Stan} source distribution in
\code{src/models/basic\_estimators/bernoulli.stan}.

\subsection{Data format}

Data for running \proglang{Stan} from the command line can be included
in \proglang{R} dump format.%
%
\footnote{A JSON interface for structured data input and output is
  currently under development.}
%
For example, 10 observations for the
model in Figure~\ref{bernoulli-model.fig} could be encoded as
%
\begin{Code}
N <- 10
y <- c(0,1,0,0,0,0,0,0,0,1)
\end{Code}
%
This defines the contents of two variables, an integer \code{N} and a
10-element integer array \code{y}.  In \pkg{RStan} and \pkg{PyStan},
data can also be passed directly through memory without the need to
read or write to a file.  This data file is provided with the
\proglang{Stan} distrbution in file
\code{src/models/basic\_estimators/bernoulli.R.stan}.

\subsection{Compling the model}

After a \proglang{C++} compiler and \code{make} are installed,%
%
\footnote{Appropriate versions are built into Linux. The \pkg{RTools}
  package suffices for Windows; it is available from
  \url{http://cran.r-project.org/bin/windows/Rtools/}.  The
  \pkg{Xcode} package contains everything needed for the Mac; see
  \url{https://developer.apple.com/xcode/} for more information.}
%
the Bernoulli model in Figure~\ref{bernoulli-model.fig} can be
translated to \proglang{C++} and compiled with a single command.
First, the directory must be changed to \code{$stan}, which we use as
a shorthand for the directory in which \proglang{Stan} was unpacked.%
%
\footnote{Before the first model is built, \code{make} must build the
  model translator (target \code{bin/stanc}) and posterior summary tool
  (target \code{bin/print}), along with an optimized version of the
  \proglang{C++} library (target \code{bin/libstan.a}).  Please be patient
  and consider \code{make} option \code{-j2} or \code{-j4} (or higher)
  to run in the specified number of processes if two or four (or more)
  computational cores are available.}
%
\begin{CodeChunk}
\begin{CodeInput}
> cd $stan
> make src/models/basic_estimators/bernoulli 
\end{CodeInput}
\end{CodeChunk}
%
This produces an executable file \code{bernoulli}
(\code{bernoulli.exe} on Windows) on the same path as the model.
Forward slashes can be used with \code{make} on Windows.

\subsection{Running the model}

\subsubsection{Command to run the model}

The executable can be run with default options by specifying a path to
the data file.  The first command in the following example changes the
current directory to that containing the model, which is where the
data resides and where the executable is built.  From there, the path
to the data is just the file name \code{bernoulli.data.R}.
%
\begin{CodeChunk}
{\small
\begin{CodeInput}
> cd $stan/src/models/basic_estimators
> ./bernoulli --data=bernoulli.data.R --samples=samp1.csv --chain_id=1 --seed=7386
\end{CodeInput}
}
\end{CodeChunk}
%
For Windows, the \code{./} before the command should be removed.  This
invocation specifies the executable program, the data file, the file
to which the samples are written, the identity of the chain, and a
seed for the random number generator.

\subsubsection{Terminal output}

The output is as follows, starting with a summary of the command-line
options used, including defaults;  these are also written into the 
samples file as comments.
%
\begin{Code}
STAN SAMPLING COMMAND
data = bernoulli.data.R
init = random initialization
init tries = 1
samples = samp1.csv
append_samples = 0
save_warmup = 0
seed = 7386 (user specified)
chain_id = 1 (user specified)
iter = 2000
warmup = 1000
thin = 1
equal_step_sizes = 0
nondiag_mass = 0
leapfrog_steps = -1
max_treedepth = 10
epsilon = -1
epsilon_pm = 0
delta = 0.5
gamma = 0.05
Iteration:    1 / 2000 [  0%]  (Warmup)
Iteration:   10 / 2000 [  0%]  (Warmup)
...
Iteration: 1990 / 2000 [ 99%]  (Sampling)
Iteration: 2000 / 2000 [100%]  (Sampling)

# Elapsed Time: 0.011269 seconds (Warm-up)
#               0.014105 seconds (Sampling)
#               0.025374 seconds (Total)
\end{Code}
%
A description of all parameters is available with the \code{{-}{-}help}
option; they provide control over input, output, and the configuration
of the sampler.

\subsubsection{Samples file output}

The output CSV file starts with a summary of the parameters for the
run.  
%
\begin{Code}
# Samples Generated by Stan
#
# stan_version_major=1
# stan_version_minor=3
# stan_version_patch=0
# model=bernoulli_model
# data=bernoulli.data.R
# init=random initialization
# append_samples=0
# save_warmup=0
# seed=7386
# chain_id=1
# iter=2000
# warmup=1000
# thin=1
# nondiag_mass=0
# equal_step_sizes=0
# leapfrog_steps=-1
# max_treedepth=10
# epsilon=-1
# epsilon_pm=0
# delta=0.5
# gamma=0.05
# algorithm=NUTS with a diagonal Euclidean metric
\end{Code}
%
It continues with a header indicating the columns of the output.
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,theta
\end{Code}
%
The values for \code{lp\_\_} indicate the log probability (up to an
additive constant).  The column headed by \code{accept\_stat\_\_}
provides the Metropolis acceptance statistic for each iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis-Hastings
  sampling such as HMC.  For NUTS, the acceptance statistic is 
  defined as the average acceptance probabilities of all possible
  samples in the proposed tree.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

Next, the results of adaptation are printed.
%
\begin{Code}
# Adaptation terminated
# Step size = 2.10898
# Diagonal elements of inverse mass matrix:
# 0.465584
\end{Code}
%
By default, the execution above uses a diagonal mass matrix, which
with one parameter is a single element.

The rest of the file contains samples, one per line, matching the
header; here the parameter \code{theta} is the final value printed on
each line, and each line corresponds to a sample.  The warmup samples
are not included by default, but may be included with the appropriate
command-line invocation of the executable. The file ends with comments reporting the elapsed time.
%
\begin{Code}
-7.04574,0.189209,2.10898,0,0.353589
-7.04574,0.00162111,2.10898,0,0.353589
-6.77448,1,2.10898,0,0.222005
...
-7.89094,0.346733,2.10898,0,0.46154
-6.93412,1,2.10898,0,0.330879
-6.93412,0.115548,2.10898,0,0.330879

# Elapsed Time: 0.011269 seconds (Warm-up)
#               0.014105 seconds (Sampling)
#               0.025374 seconds (Total)
\end{Code}
%

\subsection{Output analysis}

Before performing output analysis, we recommend generating multiple
independent chains in order to more effectively monitor convergence.
Three more chains of samples can be created as follows.
%
\begin{CodeChunk}
{\small
\begin{CodeInput}
> ./bernoulli --data=bernoulli.data.R --samples=samp2.csv --chain_id=2 --seed=7386
> ./bernoulli --data=bernoulli.data.R --samples=samp3.csv --chain_id=3 --seed=7386
> ./bernoulli --data=bernoulli.data.R --samples=samp4.csv --chain_id=4 --seed=7386
\end{CodeInput}
}
\end{CodeChunk}
%
The chains can be safely run in parallel under different processes;
details of parallel execution depend on the operating system and the shell or
terminal program. Note that, although the same seed is used for each chain, the
random numbers will in fact be independent as the chain identifier
is used to skip the pseudorandom number generator ahead.  See
Section~\ref{rng.section} for more information.

\proglang{Stan} supplies a command-line program \code{bin/print} to
summarize the output of one or more MCMC chains.  Given a directory
containing output from sampling, it is invoked as
%
\begin{CodeChunk}
\begin{CodeInput}
> ls samp*.csv
\end{CodeInput}
\begin{CodeOutput}
samp1.csv	samp2.csv	samp3.csv	samp4.csv
\end{CodeOutput}
\begin{CodeInput}
> ../../../bin/print samp*.csv
\end{CodeInput}

\end{CodeChunk}
%
The output is shown in Figure~\ref{print-output.fig}.%
%
\footnote{Aligning columns when printing rows of varying scales
  presents a challenge.  For each column, the program calculates the
  the maximum number of digits required to print an entry in that
  column with the specified precision. For example, a precision of 2
  for the number -0.000012 requires nine characters (\code{-0.000012})
  to print without scientific notation versus seven digits with
  (\code{-1.2e-5}).  If the discrepancy is above a fixed threshold,
  scientific notation is used.  Compare the results in the \code{mean}
  column versus the \code{sd} column.}
%

\begin{figure}
\small
\begin{Code}
Inference for Stan model: bernoulli_model
4 chains: each with iter=(1000,1000,1000,1000); warmup=(0,0,0,0); thin=(1,1,1,1); 
4000 iterations saved.

Warmup took (0.012, 0.010, 0.011, 0.011) seconds, 0.046 seconds total
Sampling took (0.016, 0.014, 0.014, 0.015) seconds, 0.060 seconds total

                  mean   se   sd     2.5%      50%    97.5%   n_eff n_eff/s    Rhat
lp__          -7.3e+00  0.0  0.8 -9.3e+00 -7.0e+00 -6.7e+00 1.1e+03 1.9e+04 1.0e+00
accept_stat__  5.3e-01  0.1  0.4  1.1e-06  5.2e-01  1.0e+00 3.7e+03 6.1e+04 1.0e+00
stepsize__     2.1e+00  0.1  0.2  2.0e+00  2.1e+00  2.1e+00 2.0e+00 3.3e+01 1.3e+13
treedepth__    3.6e-02  0.0  0.2  0.0e+00  0.0e+00  1.0e+00 2.9e+03 4.9e+04 1.0e+00
theta          2.5e-01  0.0  0.1  6.0e-02  2.4e-01  4.9e-01 2.1e+03 3.5e+04 1.0e+00

Samples were drawn using NUTS with a diagonal Euclidean metric.
For each parameter, n_eff is a crude measure of effective sample size,
and Rhat is the potential scale reduction factor on split chains (at 
convergence, Rhat=1).
\end{Code}
\caption{Output of \code{bin/print} for the Bernoulli estimation model in
  Figure~\ref{bernoulli-model.fig}.}\label{print-output.fig}
\end{figure}
%
Each row of the output summarizes a different value whose name is
provided in the first column.  The analysis includes the posterior
mean (\code{mean}) and standard deviation (\code{sd}).  The median
(\code{50\%}) and 95\% posterior interval (\code{2.5\%}, \code{97.5\%})
are also displayed; the \code{bin/print} command also prints the 50\%
interval boundaries, but these are ellided for space in the above
output.

The estimated potential scale reduction statistic $\hat{R}$
(\code{Rhat}) value of 1.0 for the parameter of interest \code{theta}
is consistent with convergence.  \proglang{Stan} uses a more
conservative version of $\hat{R}$ than is usual in packages such as
\pkg{Coda} \citep{PlummerEtAl:2006}, first splitting each chain in half to diagnose
montonically increasing or decreasing samples; \citep{GelmanEtAl:2013}
and \citep{Stan:2013} provide detailed definitions.

The estimated number of effective samples per parameter
(\code{n\_eff}) is more than 2000, far more than we actually need for
inference.  As with $\hat{R}$, the number of effective samples is
estimated conservatively using both cross-chain and within-chain
estimates; \citep{GelmanEtAl:2013} and the reference manual provide
details.

The MCMC standard error (\code{se}) is estimated by the posterior
standard deviation estimate divided by the square root of the
estimated number of effective samples (\code{sd / sqrt(n\_eff)}).  In
this example, the MCMC standard error is small relative to the
posterior standard deviation, another sign that a sufficient number
of samples have been drawn.

\section{Models}

In the rest of this paper, we will concentrate on the modeling
language and how compiled models are executed.  These details are the
same whether a \proglang{Stan} model is being used by a built-in or
externally supplied sampler or optimizer.

\subsection{Example: hierarchical model, with inference}

\cite[Section 5.1]{GelmanEtAl:2013} define a hierarchical model of the
incidence of tumors in rats in control groups across trials; a very
similiar model is defined for mortality rates in pediatric surgeries
across hospitals in \citep[Examples, Volume
1]{LunnEtAl:2000,LunnEtAl:2009}.  
%  
\begin{figure}
\begin{Code}
data {
  int<lower=0> J;                         // number of items
  int<lower=0> y[J];                      // number of successes for j
  int<lower=0> n[J];                      // number of trials for j
}
parameters {
  real<lower=0,upper=1> theta[J];        // chance of success for j
  real<lower=0,upper=1> lambda;          // prior mean chance of success
  real<lower=0.1> kappa;                 // prior count
}
transformed parameters {
  real<lower=0> alpha;                   // prior success count
  real<lower=0> beta;                    // prior failure count
  alpha <- lambda * kappa;
  beta <- (1 - lambda) * kappa;
}
model {
  lambda ~ uniform(0,1);                 // hyperprior
  kappa ~ pareto(0.1,1.5);               // hyperprior 
  theta ~ beta(alpha,beta);              // prior
  y ~ binomial(n,theta);                 // likelihood
}
generated quantities {
  real<lower=0,upper=1> avg;             // avg success
  int<lower=0,upper=1> above_avg[J];     // true if j is above avg
  int<lower=1,upper=J> rnk[J];           // rank of j
  int<lower=0,upper=1> highest[J];          // true if j is highest rank
  avg <- mean(theta);
  for (j in 1:J)
    above_avg[j] <- (theta[j] > avg);
  for (j in 1:J) {
    rnk[j] <- rank(theta,j) + 1;
    highest[j] <- rnk[j] == 1;
  }
}
\end{Code}
  \caption{Hierarchical binomial model with posterior inferences,
    coded in \proglang{Stan}.}\label{hier-binom.fig}
\end{figure}
%
A \proglang{Stan} implementation is provided in
Figure~\ref{hier-binom.fig}.  In the rest of this section, we will
walk through what the meaning of the various blocks are for the
execution of the model.

\subsection{Data block}

A \proglang{Stan} program starts with an (optional) data block, which
declares the data required to fit the model.  In the model in
Figure~\ref{hier-binom.fig}, the data block declares an integer
variable \code{J} for the number of groups in the hierarchical model.
The arrays \code{y} and \code{n} have size \code{J}, with \code{y[j]}
being the number of positive outcomes in \code{n[j]} trials.

All of these variables are declared with a lower-bound constraint
restricting their values to be greater than or equal to zero.
\proglang{Stan}'s constraint language is not strong enough to restrict
each \code{y[j]} to be less than or equal to \code{n[j]}.

The data for a \proglang{Stan} model is read in once as the
\proglang{C++} object representing the model is constructed.  After
the data is read in, the constraints are validated.  Failure to
validate data will terminate the program and print an informative
error message.

\subsection{Transformed data block}

The model in Figure~\ref{hier-binom.fig} does not have a transformed
data block.  A transformed data block may be used to define new
variables that can be computed based on the data.  For example,
standardized versions of data can be defined in a transformed
data block or Bernoulli trials can be summed to model as binomial.
Any constant data can also be defined in the transformed data block.

The transformed data block starts with a sequence of variable
declarations and continues with a sequence of statements defining the
variables.  For example, the following transformed data block declares
a vector \code{x\_std}, then defines it to be the standardization of \code{x}.
%
\begin{Code}
transformed data {
  vector[N] x_std;
  x_std <- (x - mean(x)) / sd(x);
}
\end{Code}

The transformed data block is executed during construction, after the
data is read in.  Any data variables declared in the data block may be
used in the variable declarations or statements.  Transformed data
variables may be used after they are declared, although care must be
taken to ensure they are defined before they are used.  Any
constraints declared on transformed data variables are validated after
all of the statements are executed, with execution terminating with an
informative error message at the first variable with an invalid value.

\subsection{Parameter block}

The parameter block in the program in Figure~\ref{hier-binom.fig}
defines three parameters.  The parameter \code{theta[j]} represents
the probability of success in group \code{j}.  The prior on each
\code{theta[j]} is parameterized by a prior mean chance of success
\code{lambda} and prior count \code{kappa}.  Both \code{theta[j]} and
\code{lambda} are constrained to fall between zero and one, whereas
\code{lambda} is constrained to be greater than or equal to 0.1 to
match the support of the Pareto hyperprior it receives in the model block.

The parameter block is executed every time the log probability is
evaluated.  This may be multiple times per iteration of a sampling or
optimization algorithm.  Furthermore, different samplers and
optimizers use different instantiations of the log probability
function depending on the form and order of the derivative information
they require; see Section~\ref{inference-engines.section} for
details.

\subsubsection{Implicit change of variables to unconstrained space}

The probability distribution defined by a \proglang{Stan} program is
intended to have unconstrained support (i.e., no points of zero
probabilty), which greatly simplifies the task of
writing samplers or optimizers.  To achieve unbounded support,
variables declared with constrained support are transformed 
to an unconstrained space.  For
instance, variables declared on $[0,1]$ are log-odds transformed and
non-negative variables declared to fall in $[0,\infty)$ are log
transformed.  More complex transforms are required for simplexes (a
reverse stick-breaking transform) and covariance matrices (Cholesky
factorization).  The dimensionality of the resulting probabilty
function may change as a result of the transform. For example, a $K
\times K$ covariance matrix requires only ${K \choose 2} + K$
unconstrained parameters, and a $K$-simplex requires only $K-1$
unconstrained parameters.

The unconstrained parameters over which the model is defined are
inverse transformed back to their constrained forms before executing
the model code.  To account for the change of variables, the log
absolute Jacobian determinant of the inverse transform is added to the
overall log probabilty function.%
%
\footnote{For optimization, the Jacobian adjustment may be optionally
  suppressed to guarantee the optimizer finds the maximum of the log
  probability function on the constrained parameters.}
%
The gradients of the log probabilty function exposed include the
Jacobian term.

There is no validation required for the parameter block because the
variable transforms are guaranteed to produce values that satisfy the
declared constraints.

\subsection{Transformed parameters block}

The transformed parameters block allows users to define transforms of
parameters within a model.  Following the model in
\citep{GelmanEtAl:2013}, the example in Figure~\ref{hier-binom.fig}
uses the transformed parameter block to define transformed parameters
\code{alpha} and \code{beta} for the prior success and failure counts
to use in the beta prior for \code{theta}.

Following the same convention as the transformed data block, the
(optional) transformed parameter block begins with declarations of the
transformed parameters, followed by a sequence of statements defining
them.  Variables from previous blocks as well as the transformed
parameters block may be used.  In the example, the prior success and
failure counts \code{alpha} and \code{beta} are defined in terms of
the prior mean \code{lambda} and total prior count \code{kappa}.

The transformed parameter block is executed after the parameter
block.  Constraints are validated after all of the statements defining the
transformed parameters have executed.  Failure to validate a
constraint results in an exception being thrown, which halts the
execution of the log probability function.  The log probability
function can be defined to return negative infinity or a special
not-a-number value.

If transformed parameters are used on the left-hand side of a sampling
statement, it is up to the user to add the appropriate log Jacobian
adjustment to the log probability accumulator.  For instance, a
lognormal variate could be generated as follows without the built-in
\code{lognormal} density function  using the normal density as
%
\begin{Code}
parameters {
  real<lower=0> u;
  ...
transformed parameters {
  real v;
  v <- log(u);
  lp__ <- lp__ + u;   // log Jacobian adjustment
}
model {
  v ~ normal(0,1);
}
\end{Code}
%
The transorm is $f(u) = \log u$, the inverse transform is $f^{-1}(v) =
\exp v$, so the absolute log Jacobian is $|\frac{d}{dv} \exp v| =
\exp v = u$.  Whenever a transformed parameter is used on the left
side of a sampling statement, a warning is printed to remind the user
of the need for a Jacobian adjustment for the change of variables.

Values of transformed parameters are saved in the output along
with the parameters.  As an alternative, local variables can be used
to define temporary values that do not need to be saved.

\subsection{Model block}

The model block defines the log probability function.  The example in
Figure~\ref{hier-binom.fig} has a simple model containing four
sampling statements.  The hyperprior on the prior mean \code{lambda}
is uniform, and the hyperprior on the prior count \code{kappa} is a
Pareto distribution with lower-bound of support at 0.1 and shape 1.5,
leading to a probability of $\kappa > 0.1$ proportional to
$\kappa^{-5/2}$.  Note that the hierarchical prior on \code{theta} is
vectorized: each elements of \code{theta} is drawn independently from
a beta distribution with prior success count \code{alpha} and prior
failure count \code{beta}.  Both \code{alpha} and \code{beta} are
transformed parameters, but because they are only used on the
right-hand side of a sampling statement do not require a Jacobian
adjustment of their own.  The likelihood function is also vectorized,
with the effect that each success count \code{y[i]} is drawn from a
binomial distribution with number of trials \code{n[i]} and chance of
success \code{theta[i]}.  In vectorized sampling statements, single
values may be repeated as many times as necessary.

The model block is executed after the transformed parameters block
every time the log probability function is evaluated.

\subsection{Generated quantities block}

The (optional) generated quantities allows values that depend on
parameters and data, but do not affect estimation, to be defined
efficiently.  The generated quantities block is called only once per
sample, not once per log probability function evaluation.  It may be
used to calculate predictive inferences as well as to carry out
forward simulation for predictive posterior checks.

The \proglang{BUGS} surgical example explored the ranking of
institutions in terms of surgical mortality.  This is coded in the
example in Figure~\ref{hier-binom.fig} using the generated quantities
block.  The generated quantity variable \code{rnk[j]} will hold the
rank of institution \code{j} from 1 to \code{J} in terms of mortality
rate \code{theta[j]}.  The ranks are extracted using the \code{rank}
function. The posterior summary will print average rank and deviation,
but the values can also be extracted in order to plot posterior rank
histograms by institution as done in \citep[Examples, Volume
1]{LunnEtAl:2000}.

Posterior comparisons can be carried out directly or using rankings.
For instance, the model in Figure~\ref{hier-binom.fig} sets
\code{highest[j]} to 1 if hospital \code{j} has the highest estimated
mortality rate.  Applying a hierarchical model then considering
posterior inference appropriately adjusts for mulitple comparisons and
allows; see \citep{GelmanEtAl:2012, Efron:2010} for discussion.

As a second illustration, the generated quantities block calculate the
(posterior) probability that a given institution is above-average in
terms of mortality rate.  This is done for each institution \code{j}
with the usual plug-in estimate of \code{theta[j] > mean(theta)},
which returns a binary (0 or 1) value.  The posterior mean of
\code{above\_avg[j]} calculates the posterior probability
$\mbox{Pr}[\theta_j > \bar{\theta}|y,n]$ according to the model.

\subsection{Initialization}

\proglang{Stan}'s samplers and optimizers all start from either random
or user-supplied values for each parameter.  User
supplied initial values are validated and transformed to
the underlying unconstrained space; if a parameter value does not satisfy its
declared constraints, the program exits and an informative error
message is printed.  If random initialization is specified, the
built-in pseudorandom number generator is called once per
unconstrained variable dimension.  The default initialization is to
randomly generate values uniformly on $[-2,2]$, which supplies fairly
diffuse starting points when transformed back to the constrained
scale.

\subsection{Variable definition and block execution summary}

A table summarizing the point at which variables are read, written,
or defined is provided in Figure~\ref{block-actions.fig}. 
%
\begin{figure}
\begin{center}
\begin{tabular}{l|c|l|l}
{ Block} & { Statements?} & { Action} & { Evaluated}
\\\hline\hline
\code{user initialization} & n/a & transform & chain
\\[3pt]
\code{random initialization} & n/a & randomize & chain 
\\\hline\hline
\code{data} & no & read & chain  
\\
\code{transformed data} & yes & evaluate & chain  
\\ \hline
\code{parameters} & no & inv.\ transform, Jacobian & leapfrog  \\
& & inv.\ transform, write & sample 
\\[3pt]
\code{transformed parameters} & yes & evaluate & leapfrog \\
& & write & sample 
\\\hline
\code{model} & yes & evaluate & leapfrog
\\\hline
\code{generated quantities} & yes & evaluate & sample \\
& & write & sample
\end{tabular}
\end{center}
\caption{Each \proglang{Stan} program block admits certain actions that are evaluated
  at specific times during sampling.  For example, the parameter initialization in
  the \code{data} block requires a read operation once per chain.}
\label{block-actions.fig}
\end{figure}
%
This table is defined assuming HMC or NUTS samplers, which require a
log probability and gradient calculation for one or more leapfrog
steps per iteration.

%
\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Variable Categorization} & { Declaration Block}
\\ \hline\hline
% constants & \code{transformed data}
% \\ \hline
unmodeled data & \code{data}, \code{transformed data}
\\ 
modeled data & \code{data}, \code{transformed data}
\\ \hline
missing data & \code{parameters}, \code{transformed parameters}
\\
modeled parameters & \code{parameters}, \code{transformed parameters}
\\
unmodeled parameters & \code{data}, \code{transformed data}
\\[2pt] \hline
generated quantities & \code{transformed data}, \code{transformed parameters}, 
\\ 
& \code{generated quantities}
\\ \hline\hline
loop indices & any loop statement
\\ \hline
local variables & any statement block
\\ 
\end{tabular}
\end{center}
\caption{Variables of each categorization must be
  declared in specific blocks.  Data
  may also be expressed using numeric
  literals.}\label{variable-kinds.fig}
\end{figure}
%
\cite[p.~366]{GelmanHill:2007} provide a taxonomy of the kinds of
variables used in Bayesian models.  Figure~\ref{variable-kinds.fig}
contains Gelman and Hill's taxonomy aligned with the corresponding
locations of declarations and definitions in \proglang{Stan}.
Unmodeled data variables includes size constants and regression
predictors.  Modeled data variables include known outcomes or
measurements.  A data variable or constant literal is modeled in a
\proglang{Stan} program if it (or a variable that depends on it)
occurs on the left-hand side of a sampling statement.  Unmodeled
parameters are provided as data, and are either known or fixed to some
value for convenience, a typical use being the parameters of a weakly
informative prior for a parameter.

A modeled parameter is given a distribution by the model (usually
dependent on data and correlated with other parameters).  This can be
done either by placing it (or a variable that depends on it) on the
left-hand side of a sampling statement or by declaring the variable
with a constraint (see Section~\ref{implicit-prior.section}).

Any variable that occurs in a (transformed) data block can also be
provided instead as a constant.  So a user may decide to write \code{y
  ~ normal(0,1)} or to declare \code{mu} and \code{sigma} as data, and
write \code{y ~ normal(mu,sigma)}.  The latter choice allows the
parameters to be changed without recompiling the model, but requires
them to be specified as part of the input data.

Missing data is a new variable type included here.  In order to
perform inference on missing data, it must be declared as a parameter
and modeled; see \citep{GelmanEtAl:2013} for a discussion of
statistical models of missing data.  Unlike \proglang{BUGS}, data may
not contain a mixture of missing (i.e., \code{NA}) values and numeric
values.  \proglang{Stan} uses the built-in \proglang{C++}
floating-point (and integer) arithmetic, for which there is no
equivalent of \code{NA}.  See \citep{Stan:2013} for strategies for
coding missing data problems in \proglang{Stan}.

Generated quantities include simple transforms of data required for
printing.  For example, the variable of interest might be a standard
deviation, resulting from transforming a precision parameter $\tau$ to
a scale parameter $\sigma = \tau^{-1/2}$.  A non-linearly transformed
variable will have different effective sample sizes and $\hat{R}$
statistics than the variable it is derived from, so it is convenient
to define variables of interest in the generated quantities block to
calculate these statistics automatically.  As seen in the example in
Figure~\ref{hier-binom.fig}, generated quantities may also be used for
event probability estimates.

The generated quantities block may also be used for forward
simulations, generating values to make predictions or to perform
posterior predictive checks; see \citep{GelmanEtAl:2013} for more
information.  The generated quantities block is the only location in
\proglang{Stan} in which random-number generators may be applied
explicitly (they are implicit in parameters).

Calculations in the generated quantities block do not impact the
estimates of the parameters.  In this way, it can be used like the cut
feature of \proglang{BUGS}.  Nevertheless, we recommend full Bayesian
inference in general, and do not use it for cut-like behavior unless
it does not effect inference (e.g., for generated predictions for
unseen data in many models).

The list is completed with two types of local variables, loop indices
and traditional local variables.  Unlike \proglang{BUGS},
\proglang{Stan} allows local variables to be declared and assigned.
For example, it is possible to compute the sum of the squares of
entries in an array or vector \code{y} as follows.%
%
\footnote{The built-in squared norm function is the right way to
  implement sums of squares in \proglang{Stan}.}
%
\begin{Code}
{ 
  real sum_of_squares;
  sum <- 0;
  for (n in 1:N)
    sum <- sum + y[n] * y[n];
}  
\end{Code}
%
Local variables may not be declared with constraints, because there is
no location at which it makes sense to test that they satisfy the
constraints.

\subsection{Implicit Uniform Priors}\label{implicit-prior.section}

The default distribution for a variable is uniform over its support.
For instance, a variable declared with a lower bound of 0 and an upper
bound of 1 implicitly receives a $\mbox{\sf Uniform}(0,1)$
distribution.  These implicit uniform priors are improper if the
variable has unbounded support.  For instance, the uniform
distribution over simplexes and correlation matrices is proper, but
the uniform distribution over ordered vectors or covariance matrices
is not.

\subsection{The ``Matt trick'' for hierarchical models}

Consider the following hierarchical logistic regression model fragment.
%
\begin{Code}
data {
  int<lower=1,upper=K> group[N];  // group of data item n
}  
parameters {
  ...
  vector mu;             // mean coeff
  real<lower=0> sigma;   // coeff scale
  vector[K] beta;        // coeff for group k
}
model {
  beta ~ normal(mu,sigma);
  for (n in 1:N)
    y[n] ~ bernoulli_logit(beta[group[n]] * x[n]);
  ...
}
\end{Code}
%
With the direct parameterization, a change in \code{sigma} is
amplified by the lower-level parameters and introduces a large
change in density across the posterior.  Unfortunately the 
expected density variation of an HMC transition is limited%
%
\footnote{The limited density variation is a ultimately consequence
of a constant metric.  An additional benefit of RMHMC are transitions
that can cover much larger variations in density, making it uniquely
suited to these models.}
%
and many transitions are required to explore the full posterior.
The resulting sampler devolves into a random walk with high autocorrelations.

The following reparameterization employs the ``Matt trick'',
%
\footnote{The reparameterization is named after Matt Hoffman, who
  introduced it during modeling discussions in the \proglang{Stan}
  group meetings.  A similar reparameterization by \citep{Choo:2000}
  is discussed in \citep[Section~5.4.5]{Neal:2011}.}
%
\begin{Code}
parameters {
  vector[K] beta_raw;    
}
transformed parameters {
  vector[K] beta;
  beta <- mu + sigma * beta_raw;
}
model {
  beta ~ normal(0,1);
}
\end{Code}
%
This reparameterization removes the particular correlations amongst the
hierarchical variables that would otherwise limit the effectiveness
of the samplers.

\section{Data types}

Expressions in \proglang{Stan}, including variables are statically
typed.  This means their type is declared at compile time as part of
the model, and does not change throughout the execution of the
program.  This is the same behavior as is found in compiled
programming languages such as \proglang{C(++)}, \proglang{Fortran},
and \proglang{Java}, but is unlike the behavior of interpreted
languages such as \proglang{BUGS}, \proglang{R}, and
\proglang{Python}.  Statically typing the variables (as well as
declaring them in appropriate blocks based on usage) makes
\proglang{Stan} programs easier to read and easier to debug by making
explicit the modeling decisions and expression types.

\subsection{Primitive Types}

The primitive types of \proglang{Stan} are \code{real} and \code{int},
which are used to represent continuous and integral values.  These
values are represented directly in \proglang{C++}.  Integer
expressions can be used anywhere a real value is required, but not
{\it vice-versa}.

\subsection{Vector and Matrix Types}

\proglang{Stan} supports vectors, row vectors, and matrices with the
usual access operations.  Indexing for vector, matrix, and array types
starts from one.

Vectors are declared with their sizes and matrices with their number
of rows and columns.

All vector and matrix types contain real values and may not be
declared to contain integers.  Collections of integers are represented
using arrays.

\subsection{Array Types}

An array may have entries of any other type.  For example, arrays of
integers and reals are allowed, as are arrays of vectors or arrays of
matrices.

Higher-dimensional arrays are intrinsically arrays of arrays.  An
entry two-dimensional array \code{y} may be accessed as \code{y[1,2]}
or \code{y[1][2]}, with the expression \code{y[1]} denoting an array.%
%
\footnote{ Arrays are stored in row-major order and matrices in
  column-major order.}

\subsection{Constrained Variable Types}

Variables may be declared with constraints.  The constraints have
different effects depending on the block in which the variable is
declared.

Integer and real types may be provided with lower bounds, upper
bounds, or both.  This includes the types used in arrays, and the real
types used in vectors and matrices.

Vector types may be constrained to be unit simplexes (all entries
non-negative and summing to 1), unit length vectors (sum of squares is 1), or
ordered (entries are in ascending order), positive ordered (entries in
ascending order, all non-negative), using the types \code{simplex[K]},
\code{unit_vector[K]}, \code{ordered[K]}, or
\code{positive_ordered[K]}, where \code{K} is the size of the vector.

Matrices may be constrained to be covariance matrices (symmetric,
positive definite) or correlation matrices (symmetric, positive
definite, unit diagonal), using the types \code{cov_matrix[K]} and
\code{corr_matrix[K]}.

\section{Expressions and type inference}

The syntax of \code{Stan} is defined in terms of expressions and
statements.  Expressions denote values of a particular type.  
Statements represent operations such as assignment and sampling as
well as control structures such as for loops and conditionals.

\subsection{Expressions}

\proglang{Stan} provides the usual kinds of expressions found in
programming languages.  This includes variables, literals denoting
integers, real values or strings, binary and unary operators over
expressions, and function application.

\subsubsection{Type inference}

The type of each variable is declared statically and cannot change.

The type of a numeric literal is determined by whether or not it
contains a period or scientific notation; for example, \code{20} has
type \code{int} whereas \code{20.0} and \code{2e+1} have type
\code{real}.

The type of applying an operator or a function to one or more
expressions is determined by the available signatures for the
function.  For example, the multiplication operator (\code{*}) has a
signature that maps two \code{int} arguments to an \code{int} and two
\code{real} arguments to a \code{real} result.  Another signature for
the same operator maps a \code{row\_vector} and a \code{vector} to a
\code{real} result.

\subsubsection{Type promotion}

If necessary, an integer type will be promoted to a \code{real} value.
For example, multiplying an \code{int} by a \code{real} produces a
\code{real} result by promoting the \code{int} argument to a
\code{real}.  \proglang{Stan} also attempts to treat arrays as
covariant, meaning that arrays of of \code{int} values can often be
used where arrays of \code{real} values are specified.

\section{Statements}

\subsection{Assignment and sampling}

\proglang{Stan} supports the same two basic statements as
\proglang{BUGS}, assignment and sampling, examples of which were
introduced earlier.  In \proglang{BUGS}, these two kinds of statment
define a directed acyclic graphical model; in \proglang{Stan}, they
define a log probability function.

\subsubsection{Log probability accumulator}

There is an implicitly defined variable \code{lp\_\_} (available in
the transformed parameters and model blocks) denoting the log
probability that will be returned by the log probability function.  A
sampling statement is nothing more than shorthand for incrementing the
log probability accumulator \code{lp\_\_}.  For example, if \code{beta}
is a parameter of type \code{real}, the sampling statement
%
\begin{Code}
beta ~ normal(0,1);
\end{Code}
%
has the exact same effect as the assignment statement
%
\begin{Code}
lp__ <- lp__ + normal_log(beta,0,1);
\end{Code}

Because computation is only up to a proportion, this sampling
statement in turn has the same effect as the direct implementation in
terms of basic arithmetic,
%
\begin{Code}
lp__ <- lp__ - 0.5 * beta * beta;
\end{Code}
%
If \code{beta} is of type \code{vector}, replace \code{beta * beta}
with \code{beta' * beta}.  Distributions whose probability functions
are not built directly into \proglang{Stan} can be implemented
directly in this fashion.

\subsection{Sequences of statements and execution order}

\proglang{Stan} allows sequences of statements wherever statements may
occur. Unlike \proglang{BUGS}, in which statements define a directed
acyclic graph, in \proglang{Stan}, statements are executed
imperatively in the order in which they occur in a program.

\subsubsection{Blocks and variable scope}

Sequences of statements surrounded by curly braces (\code{\{} and
\code{\}}) form blocks.  Blocks may start with local variable
declarations.  The scope of a local variable (i.e., where it is
available to be used) is that of the block in which it is declared.

Other variables, such as those declared as data or parameters, may
only be assigned to in the block in which they are declared.  They may
be used in the block in which they are declared and may also be used
in any block after the block in which they are declared.

\subsection{Whitespace, semicolons, and comments}

Following the convention of \proglang{C++}, statements are separated
with semicolons in \proglang{Stan} so that the content of whitespace
(outside of comments) is irrelevant.  This is in contrast to
\proglang{BUGS} and \proglang{JAGS} in which carriage returns are
special and may indicate the end of a statement.

\proglang{Stan} supports the line comment style of \proglang{C++},
using two forward slashes (\code{//}) to comment out the rest of a
line; this is the one location where the content of whitespace
matters.  \proglang{Stan} also supports the line comment style of \proglang{R} and
\proglang{BUGS}, treating a pound sign (\code{\#}) as commenting out
everything until the end of the line.  \proglang{Stan} also supports
\proglang{C++}-style block comments, with everything between the
start-comment (\code{/*}) and end-comment (\code{*/}) markers being
ignored.

The preferred style follows that of \proglang{C++}, with line comment
used for everything but multiline comments.

\subsection{Control structures}

\proglang{Stan} supports the same kind of explicitly bounded for loops
as found in \code{BUGS} and \code{R}.  Like \code{R}, but unlike
\proglang{BUGS}, \proglang{Stan} supports while loops and conditional
(if-then-else) statements.%
%
\footnote{\proglang{BUGS} omits these control structures because they
  would introduce ambiguities into the directed, acyclic graph defined
  by model.}
%
\proglang{Stan} provides the usual comparison operators and boolean
operators to help define conditionals and condition-controlled while
loops.

\subsection{Print statements and debugging}

\proglang{Stan} provides print statements which take arbitrarily many
arguments consisting of expressions or string literals consisting of
sequences of characters surrounded by double quotes (\code{"}).
These statements may be used for debugging purposes to report on
intermediate states of variables or to indicate how far execution has
proceeded before an error.

\section{Function and distribution library}

In order to support the algorithmic differentiation required to
calculate gradients, Hessians, and higher-order derivatives in
\proglang{Stan}, we require \proglang{C++} functions that are 
templated separately on all of their arguments.  In order for these
functions to be efficient in computing both values and derivatives,
they need to be vectorized so that shared computations can be reused.

\subsection{Basic operators}

\proglang{Stan} supports all of the basic \proglang{C++} arithmetic
operators, boolean operators, comparison operators  In addition, it
extends the arithmetic operators to matrices and includes pointwise
matrix operators.%
%
\footnote{This is in contrast to \proglang{R} and \proglang{BUGS}, who
  treat the basic multiplication and division operators pointwise and
  use special symbols for matrix operations.}
%
The full set of operators is listed in Figure~\ref{operators.fig}.

\subsection{Special functions}

\proglang{Stan} provides an especially rich set of special functions.
This includes all of the \proglang{C++} math library functions, as
well as numerous more specialized functions such as Bessel functions,
gamma and digamma functions, and generalized linear model link
functions and their inverses.  There are also many compound functions,
such as \code{log1m(x)}, which is more stable arithmetically for
values of x near 0 than \code{log(1 - x)}.  \proglang{Stan}'s special
functions are listed in Figures~\ref{special-functions.fig} and
\ref{special-functions-cont.fig}.

\subsection{Matrix and linear algebra functions}

Following the usual convention in mathematics, matrix and array
indexing uses the usual square brackets (\code{[ ]}) operator, and
begins from 1.  \proglang{R}, \proglang{MATLAB}, and \proglang{BUGS},
but not of \proglang{C++}, \proglang{Python}, or \proglang{Java}.
If \code{Sigma} is a matrix, then \code{Sigma[n]} is the \code{n}-th
row of \code{Sigma}.

Various reductions are provided for arrays and matrices, such as sums,
means, standard deviations and norms.  Replications are also available
to copy a value into every cell of a matrix.  Slices of matrices and
vectors may be accessed by row, column, or general sub-block
operations.

Matrix operators use the types of their operands to determine the type
of the result.  For instance, multiplying a vector by a (column) row
vector returns a matrix, whereas multiplying a row vector by a
(column) vector returns a real.  A postfix apostrophe (\code{'}) is
used for matrix and vector transposition.  For example, if \code{y}
and \code{mu} are vectors and \code{Sigma} is a square matrix, all of
the same dimensionality, then \code{y - mu} is a vector, \code{(y -
  mu)'} is a row vector, \code{(y - mu)' * Sigma} is a row vector, and
\code{(y - mu)' * Sigma * (y - mu)} will be a real value.  Matrix
division is provided, which is much more arithmetically stable than
inversion, e.g., \code{(y - mu)' / Sigma} computes the same function
as \code{(y - mu)' * inverse(Sigma)}.  \proglang{Stan} also supports
elementwise multiplication (\code{.*}) and division (\code{./}).

Linear algebra functions are provided for trace, left and right
division, Cholesky factorization, determinants and log determinants,
inverses, eigenvalues and eigenvectors, and singular value
decomposition.  All of these operations may be applied to matrices of
parameters or constants.  Various functions are specialized for speed,
such as quadratic products, diagonal specializations, multiply by self
transposed, e.g., the previous example could be written as
\code{quad\_form(Sigma, y - mu)}.

The full set of matrix and linear-algebra functions is listed in
Figure~\ref{matrix-functions.fig}; operators, which also apply to
matrices and vectors, are listed in Figure~\ref{operators.fig}.

\subsection{Probability functions}

Stan supports a growing collection of built-in univariate and
multivariate probability density and mass functions.  These
probability functions share various features of their declarations and
behavior.

All probability functions are defined on the log scale to avoid
underflow.  They are all named with the suffix \code{\_log}, e.g.
\code{normal\_log()}, is the log-scale normal distribution density
function.

All probability functions check that their arguments are within the
appropriate constrained support and may be configured to throw
exceptions or return $-\infty$ or a special not-a-number value
(\code{NaN}) for out-of-domain arguments (the behavior of positive and
negative infinity and not-a-number values are built into
floating-point arithmetic).  For example,
\code{normal\_log(y,} \code{mu,} \code{sigma)} requires the scale parameter
\code{sigma} to be non-negative.

The list of probability functions is provided in Figures
\ref{prob-functions.fig}, \ref{prob-functions-cont.fig}, and
\ref{prob-functions-cont-2.fig}.

\subsubsection{Up to a proportion calculations}

All probability functions support calculating results up to a constant
proportion, which becomes an additive constant on the log scale.
Constancy here refers to being a numeric literal such as \code{1}, a
constant function such as \code{pi()}, a variable declared as data or
transformed data, or an expression involving only other constants.
Non-constants include parameters, transformed parameters, local
variables declared in the transformed parameters or model statements,
as well as any expression involving a non-constant.

Constant terms are dropped from probability function calculations at
the time the model is compiled, so there is no run-time overhead to
decide which expressions denote constants.%
%
\footnote{Both vectorization and dropping constant terms are
  implemented in \proglang{C++} through template metaprograms that
  infer traits of template arguments to the probability functions.
  Whether to drop constants is configurable through a boolean template
  parameter on the log probability and derivative functions generated
  in \proglang{C++} for a model.}
%
For example, calling \code{normal\_log(y,0,1)}, where \code{y} is a
parameter and \code{0} and \code{1} are constants, does not bother
computing the $\log \sigma$ term in the log density because $\sigma=1$
is a constant.  On the other hand, the $y - 0$ term is calculated and
is not a constant because $y$ is not a constant.

\subsubsection{Vectorization}

All of the univariate%
%
\footnote{We are in the process of vectorizing the multivariate
  probability functions, but they are not all available in
  \proglang{Stan} 2.0.  We are also in the process of vectorizing
  all of the special functions.}
%
probability functions in \proglang{Stan} are vectorized so that they
accept arrays or vectors of arguments.  For example, although the
basic signature of the probability function
\code{normal\_log(y,mu,sigma)} involves real \code{y}, \code{mu} and
\code{sigma}, it supports calls in which any any or all of \code{y},
\code{mu} and \code{sigma} contain more than one element. A typical
use case would be for linear regression, such as \code{normal\_log(y,X
  * beta,sigma)}, where \code{y} is a vector of observed data,
\code{X} is a predictor matrix, \code{beta} is a coefficient vector,
and \code{sigma} is a real value for the noise scale.

The advantage of using vectorization is twofold.  First, the models
are more concise and closer to mathematical notation.  Second, the
vectorized versions are much faster.  They reduce the number of times
expensive operations need to be evaluated and also reduce the number
of virtual function calls required in the compiled \proglang{C++}
executable for calculating gradients and higher-order derivatives.
In the example above, \code{normal\_log(y,X *
  beta,sigma)}, the logarithm of \code{sigma} need only be computed
once; if \code{y} is an $N$-vector, it also reduces the number of
virtual function calls from $N$ to 1.

\section{Built-in inference engines}\label{inference-engines.section}

\proglang{Stan} includes several Markov chain Monte Carlo (MCMC)
samplers and several optimizers.  Others may be straightforwardly
implemented within \proglang{Stan}'s \proglang{C++} framework for
sampling and optimization using the log probability and derivative
information supplied by a model.

\subsection{Markov chain Monte Carlo samplers}

\subsubsection{Hamiltonian Monte Carlo}

The MCMC samplers provided include Euclidean Manifold Hamiltonian
Monte Carlo (EM-HMC) \citep{DuaneEtAl:1987, Neal:1994, Neal:2011} and
the No-U-Turn sampler (NUTS) \citep{HoffmanGelman:2011}.  Both
versions of HMC allow estimation or specification of unit, diagonal,
or full mass matrices.  NUTS, the default sampler for \proglang{Stan},
automatically adapts the number of leapfrog steps, eliminating the 
need for user-specified tuning parameters.  Both algorithms take advantage
of curvature information, namely the gradient, to generate coherent motion
through the posterior that dramatically reduces the autocorrelation of
the resulting transitions.

\proglang{Stan} will soon have an implementation of Riemannian Manifold
Hamiltonian Monte Carlo (RM-HMC) \citep{GirolamiCalderhead:2011},
using the SoftAbs metric to robustly incorporate posterior curvature
into the Hamiltonian simulation \cite{Betancourt:2012}, although
\proglang{Stan}'s MCMC framework makes it
straightforward to incorporate other metrics.  This
implementation will also generalize NUTS to Riemannian manifolds
\citep{Betancourt:2013}.  Both support adapting step sizes during
warmup.  RM-HMC uses first, second, and third order derivatives 
to adapt to the local curvature of the posterior.

\subsubsection{Metropolis-Hastings}

\proglang{Stan} supports random-walk Metropolis-Hastings samplers with
multivariate normal proposals, the primary purpose of which is to
provide baselines against which to measure more efficient approaches
to sampling.  The covariance matrix used can be adapted during warmup,
assuming either equal diagonal covariance, diagonal covariance, or a
full covariance matrix.  Metropolis-Hastings does not require any
derivatives of the log probability function, but the lack of gradient
information induces random walk behavior.

\subsubsection{Ensemble Samplers}

Two flavors of ensemble samplers are also in the works: an affine invariant
ensemble sampler \citep{GoodmanWeare:2010} and a differential
evolution sampler \cite{TerBraak:2006}.  Ensemble samplers do not
require derivative information, but typically require at least as many
chains to be run as the number of dimensions in order to have full
rank search of the posterior.

\subsection{Optimizers}

\subsubsection{BFGS}

The default optimizer in \proglang{Stan} is the
Broyden-Fletcher-Goldfarb-Shannon-Boyd (BFGS) optimizer.  BFGS is a
quasi-Newton optimizer that evaluates gradients directly, then uses
the gradients to update an approximation to the Hessian; see
\citep{NocedalWright:2006} for details.  Plans are in the works to
also include the more involved, but more scalable limited-memory BFGS
(L-BFGS) scheme.

\subsubsection{Conjugate gradient}

\proglang{Stan} provides a standard form of conjugate gradient
optimization; see \citep{NocedalWright:2006}.  As its name implies,
conjugate gradient optimization requires gradient evaluations.

\subsubsection{Accelerated gradient}

Additionally, \proglang{Stan} implements a crude version of Nesterov's
accelerated gradient optimizer \cite{Nesterov:1983} which combines
gradient updates with a momentum-like update to speed up convergence.

\section{Random number generation}\label{rng.section}

Random number generation for \proglang{Stan} is done on a per-chain
basis; ensemble samplers form a single joint chain over the ensemble.  By
specifying the chain being used, the random number generator can be
skipped ahead sufficiently to avoid replication of subsequences of
random numbers across chains.

The generated model code and underlying \proglang{C++} algorithms
provide template parameters for a class implementing the \pkg{Boost}
random number generator concept.

By default, \proglang{Stan} uses linear congruential generators
\citep{LEcuyer:1988}.  This generator supports efficient skip-ahead.

\subsection{Replicability}

The \proglang{Stan} interfaces all allow random-number generator seeds
to be specified explicitly.  Execution uses a single base random
number generator instance.  Therefore, by specifying a seed,
\proglang{Stan}'s behavior is deterministic.  This is very useful for
debugging purposes.  Seeds can be generated randomly based on
properties of the system time, but when they are, the seed used is
printed as part of the output to allow it to be fully replicated.

\section{Library dependencies}

\proglang{Stan}'s modeling language is only dependent on two external
libraries.

\subsection{Boost}

\proglang{Stan} depends on several of the \pkg{Boost} \proglang{C++}
libraries \citep{Boost:2011}.  \proglang{Stan} makes extensive use of
\pkg{Boost}'s template metaprogramming facilities including the
\pkg{Enable if} package, the \pkg{Type Traits} library, and the
\pkg{Lexical Cast} library.  The \proglang{Stan} language is parsed
using \pkg{Boost}'s \pkg{Spirit} parser, which itself depends on
the binding libraries \pkg{Phoenix}, \pkg{Bind}, and \pkg{Lambda},
the variant type library \pkg{Variant}, and the container library
\pkg{Fusion}.  Exceptions are largely handled and configured through
the error checking facilities in the \pkg{Math} and
\pkg{Exception} packages.  Output formatting and ad-hoc input parsing
for various formats is facilitated with the \pkg{Format} library.
\proglang{Stan} relies heavily on the special functions defined in the
\pkg{Math} subpackages \pkg{Special Functions} and \pkg{Statistical
  Distributions}.  Random number generation is carried out using the
\pkg{Random} package.  The posterior analysis framework and some
built-in functions depend on the \pkg{Accumulators} package.

\subsection{Eigen}

\proglang{Stan}'s handling of matrices and linear algebra is
implemented through the \pkg{Eigen} \proglang{C++} template library
\citep{Eigen:2012}.  Eigen uses template metaprogramming to achieve
state-of-the-art performance for matrix and linear algebra operations
with a great deal of flexiblity with respect to input types.
Unfortunately, many of the expression templates that Eigen uses for
efficient static anaysis and lazy evaluation are short-circuited
because of \proglang{Stan}'s need to have mixed type operations (i.e.,
multiplying a constant predictor matrix of double values by a
parameter vector of algorithmic differentiation values).  To make up for
this in some important cases, \proglang{Stan} has provided compound
functions such as the quadratic form, which allow speedups of both the
matrix operations and their derivatives compared to a direct
implementation using \proglang{Stan}'s built-in operators.

\section{Developer process}

\subsection{Version control and source repository}

\proglang{Stan}'s source code is hosted on GitHub and managed using
the \pkg{Git} version control system \citep{Chacon:2009}.  To manage
the workflow with so many developers working at any given time, the
project follows the GitFlow process \citep{Driessen:2010}.  All
developer submissions are managed through pull requests and we have
gratefully received patches from numerous sources outside the core
development team.

\subsection{Continuous integration}

\proglang{Stan} uses continuous integration, meaning that the entire
program and set of tests are run automatically as code is pushed to
the \pkg{Git} repository.  Each pull request is tested for
compatibility with the development branch, and the development branch
itself is tested for stability.  \proglang{Stan} uses Jenkins
\citep{Smart:2011}, an open-source continuous integration server.

\subsection{Testing framework}

This includes extensive unit tests for low-level \proglang{C++} code.
Unit tests are implemented using the \pkg{googletest} framework
\citep{GoogleTest:2011}.  The probability functions and command-line
invocations are complex enough that programs are used to automatically
generate test code for \pkg{googletest}.

These unit tests evaluate every function for both appropriate values
and appropriate derivatives.  This requires an extensive meta-testing
framework for the probability distributions due to their high degree
of configurability as to argument types.  The testing portion of the
makefile also runs tests of all of the built-in models, including
almost all of the \proglang{BUGS} sample models.  Models are tested
for both convergence and posterior mean estimation to within MCMC
standard error.

\subsection{Builds}

The build process for \proglang{Stan} is highly automated through a
cross-platform series of \code{make} files.  The top-level makefile
builds the \proglang{Stan}-to-\proglang{C++} translator command
\code{bin/stanc} and posterior analysis command \code{bin/print}.  It
also builds the library archive \code{bin/libstan.a}.  Great care was
taken to avoid complicated platform-dependent configuration
requirements that place a high burden on user system knowledge for
installation.  All that is needed is a relatively recent
\proglang{C++} compiler and version of \code{make}.

As exemplified in the introduction, the makefile is automated enough
to build an executable form of a \proglang{Stan} model in a single
command.  All libraries and other executables will be built as a side
effect.

The top-level makefile also supplies targets to build all of the
documentation \proglang{C++} API documentation is generated using the
\pkg{doxygen} package \citep{Doxygen:2011}.  The \proglang{Stan}
manual \citep{Stan:2013} is typeset using the \LaTeX package
\citep{MittelbachEtAl:2004}.

The makefile also has targets for all of the unit and functional
testing, for building the source-level distribution, and for cleaning
any temporary files that it creates.

\section*{Acknowledgments}

First and foremost, we would like to thank all of the users of
\proglang{Stan} for taking a chance on a new package and sticking with
it as we ironed out the details in the first release.  We'd like to
particularly single out the students in Andrew Gelman's Bayesian data
analysis courses at Columbia Univesity and Harvard University, who
served as trial subjects for both \proglang{Stan} and \cite{GelmanEtAl:2013}.

Stan was and continues to be supported largely through grants from the
U.~S. government.  Grants which indirectly supported the initial
research and development included grants from the Department of Energy
(DE-SC0002099), the National Science Foundation (ATM-0934516), and the
Department of Education Institute of Education Sciences
(ED-GRANTS-032309-005 and R305D090006-09A).  The high-performance
computing facility on which we ran evaluations was made possible
through a grant from the National Institutes of Health
(1G20RR030893-01).  Stan is currently supported in part by a grant
from the National Science Foundation (CNS-1205516).

Finally, we would like to thank all those who contributed
documentation corrections and code patches, Jeffrey Arnold, David
R. Blair, Eric C. Brown, Eric N. Brown, Devin Caughey, Wayne Folta,
Andrew Hunter, Marco Inacio, Louis Luangkesorn, Jeffrey Oldham, Mike
Ross, Terrance Savitsky, Yajuan Si, Dan Stowell, Zhenming Su, and
Dougal Sutherland.

And a special thanks to those who have contributed code for new
features, Jeffrey Arnold, Yuanjun Gao, and Marco Inacio.

\begin{figure}
\begin{center}
\begin{tabular}{c|ccl|l}
{ Operation} & { Precedence} & { Associativity} & {\it
  Placement} & { Description}
\\ \hline \hline
\code{||} & 9 & left & binary infix & logical or
\\ \hline
\Verb|&&| & 8 & left & binary infix & logical and
\\ \hline
\Verb|==| & 7 & left & binary infix & equality
\\
\Verb|!=| & 7 & left & binary infix & inequality
\\ \hline
\Verb|<| & 6 & left & binary infix & less than
\\
\Verb|<=| & 6 & left & binary infix & less than or equal
\\
\Verb|>| & 6 & left & binary infix & greater than 
\\
\Verb|>=| & 6 & left & binary infix & greater than or equal
\\ \hline
\code{+} & 5 & left & binary infix & addition
\\
\code{-} & 5 & left & binary infix & subtraction
\\ \hline
\code{*} & 4 & left & binary infix & multiplication
\\
\code{/} & 4 & left & binary infix & (right) division
\\ \hline
\Verb|\| & 3 & left & binary infix & left division
\\ \hline
\code{.*} & 2 & left & binary infix & elementwise multiplication
\\
\code{./} & 2 & left & binary infix & elementwise division
\\ \hline
\code{!} & 1 & n/a & unary prefix & logical negation
\\
\code{-} & 1 & n/a & unary prefix & negation
\\ 
\code{+} & 1 & n/a & unary prefix & promotion (no-op in \proglang{Stan})
\\ \hline
\code{'} & 0 & n/a & unary postfix & transposition
\\ \hline \hline
\code{()} & 0 & n/a & prefix, wrap & function application
\\
\code{[]} & 0 & left & prefix, wrap & array, matrix indexing
\end{tabular}
\end{center}
\caption{Each of \proglang{Stan}'s unary and binary operators follow
  strict precedences, associativities, placement within an expression.  
  The operators are listed in order of precedence, from least tightly
  binded to most tightly binding.}\label{operators.fig}
\end{figure}
%

\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Function} & { Description} \\ \hline \hline
\code{e} &  base of natural logarithm \\ 
\code{epsilon} &  smallest positive number \\ 
\code{negative\_epsilon} &  largest negative value \\ 
\code{negative\_infinity} &  negative infinity \\ 
\code{not\_a\_number} &  not-a-number \\ 
\code{pi} &  $\pi$ \\
\code{positive\_infinity} &  positive infinity \\  
\code{sqrt2} &  square root of two \\ 
\end{tabular}
\end{center}
\caption{\proglang{Stan} implements a variety of useful constants.}\label{constants.fig}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Function} & { Description} \\ \hline \hline
\code{acos} &  arc cosine \\ 
\code{acosh} &  arc hyperbolic cosine \\ 
\code{asin} &  arc sine \\ 
\code{asinh} &  arc hyperbolic sine \\ 
\code{atan} &  arc tangent \\ 
\code{atan2} &  arc ratio tangent \\ 
\code{atanh} &  arc hyperbolic tangent \\ 
\code{cos} &  cosine \\ 
\code{cosh} &  hyperbolic cosine\\ 
\code{hypot} &  hypoteneuse \\ 
\code{sin} &  sine \\ 
\code{sinh} &  hyperbolic sine \\ 
\code{tan} &  tangent \\ 
\code{tanh} &  hyperbolic tangent \\ 
\end{tabular}
\end{center}
\caption{\proglang{Stan} implements both circular and 
  hyperbolic trigonometric functions, as well as their inverses.}
  \label{trig-functions-cont.fig}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Function} & { Description}  \\ \hline \hline
\code{abs} &  double absolute value \\ 
\code{abs} &  integer absolute value \\ 
\code{binary\_log\_loss} &  log loss \\ 
\code{bessel\_first\_kind} & Bessel function of the first kind \\
\code{bessel\_second\_kind} & Bessel function of the second kind \\
\code{binomial\_coefficient\_log} & log binomial coefficient \\ 
\code{cbrt} &  cube root \\ 
\code{ceil} &  ceiling \\ 
\code{cumulative\_sum} &  cumulative sum \\ 
\code{erf} &  error function \\ 
\code{erfc} &  complementary error function \\ 
\code{exp} &  base-$e$ exponential \\ 
\code{exp2} &  base-2 exponential \\ 
\code{expm1} &  exponential of quantity minus one \\
\code{fabs} &  real absolute value \\ 
\code{fdim} &  positive difference \\ 
\code{floor} &  floor \\ 
\code{fma} &  fused multiply-add \\ 
\code{fmax} &  floating-point maximum \\ 
\code{fmin} &  floating-point minimum \\ 
\code{fmod} &  floating-point modulus \\ 
\code{if\_else} &  conditional \\ 
\code{int\_step} &  Heaviside step function \\ 
\code{inv} &  inverse (one over argument) \\ 
\code{inv\_cloglog} &  inverse of complementary log-log \\ 
\code{inv\_logit} &  logistic sigmoid \\ 
\code{inv\_sqrt} &  inverse square root \\ 
\code{inv\_square} &  inverse square \\ 
\code{lbeta} & log beta function \\ 
\code{lgamma} &  log $\Gamma$ function \\ 
\code{lmgamma} &  log multi-$\Gamma$ function \\ 
\code{log} &  natural (base-$e$) logarithm \\ 
\code{log10} &  base-10 logarithm \\ 
\code{log1m} &  natural logarithm of one minus argument \\ 
\code{log1m\_exp} &  natural logarithm of one minus natural exponential \\ 
\code{log1m\_inv\_logit} &  natural logarithm of logistic sigmoid \\ 
\code{log1p} &  natural logarithm of one plus argument \\ 
\code{log1p\_exp} &  natural logarithm of one plus natural exponential \\ 
\code{log2} &  base-2 logarithm \\ 
\end{tabular}
\end{center}
\caption{\proglang{Stan} implements many special and transcendental functions.}\label{special-functions.fig}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Function} & { Description} \\ \hline \hline
\code{log\_diff\_exp} & natural logarithm of difference of
exponentials \\ 
\code{log\_falling\_factorial} & falling factorial (Pochhammer) \\
\code{log\_inv\_logit} &  natural logarithm of the logistic sigmoid \\ 
\code{log\_rising\_factorial} & falling factorial (Pochhammer) \\
\code{log\_sum\_exp} &  logarithm of the sum of exponentials of arguments \\ 
\code{logit} &  log-odds \\ 
\code{max} &  integer maximum \\ 
\code{max} &  real maximum \\ 
\code{mean} &  sample average \\ 
\code{min} &  integer minimum \\ 
\code{min} &  real minimum \\ 
\code{modified\_bessel\_first\_kind} & modified Bessel function of the first kind \\
\code{modified\_bessel\_second\_kind} & modified Bessel function of the second kind \\
\code{multiply\_log} &  multiply linear by log \\ 
\code{owens\_t} &  Owens-t \\ 
\code{phi} &  $\Phi$ function (cumulative unit normal) \\ 
\code{phi\_approx} &  efficient, approximate $\Phi$ \\ 
\code{pow} &  power (i.e., exponentiatiation) \\
\code{prod} &  product of sequence \\
\code{rank} &  rank of element in array or vector \\ 
\code{rep\_array} &  fill array with value \\
\code{round} &  round to nearest integer \\ 
\code{sd} &  sample standard deviation \\ 
\code{softmax} &  softmax (multi-logit link) \\ 
\code{sort\_asc} &  sort in ascending order \\ 
\code{sort\_desc} &  sort in descending order \\ 
\code{sqrt} &  square root \\ 
\code{square} &  square \\ 
\code{step} &  Heaviside step function \\ 
\code{sum} &  sum of sequence \\ 
\code{tgamma} &  $\Gamma$ function \\ 
\code{trunc} & truncate real to integer \\ 
\code{variance} &  sample variance \\ 
\end{tabular}
\end{center}
\caption{Special functions (continued).}\label{special-functions-cont.fig}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Function} & { Description} \\ \hline \hline
\code{block} &  sub-block of matrix \\
\code{cholesky\_decompose} &  Cholesky decomposition \\ 
\code{col} &  column of matrix \\ 
\code{cols} &  number of columns in matrix \\ 
\code{columns\_dot\_product} &  dot product of matrix columns \\ 
\code{columns\_dot\_self} &  dot product of matrix columns with self \\ 
\code{crossprod} &  cross-product \\ 
\code{determinant} &  matrix determinant \\ 
\code{diag\_matrix} &  vector to diagonal matrix \\ 
\code{diag\_post\_multiply} &  post-multiply matrix by diagonal matrix \\ 
\code{diag\_post\_multiply} &  pre-multiply matrix by diagonal matrix \\ 
\code{diagonal} &  diagonal of matrix as vector \\ 
\code{dims} &  dimensions of matrix \\ 
\code{dot\_product} &  dot product \\ 
\code{dot\_self} &  dot product with self \\ 
\code{eigenvalues\_sym} &  eigenvalues of symmetric matrix \\ 
\code{eigenvectors\_sym} &  eigenvectors of symmetric matrix \\ 
\code{head} &  head of vector \\
\code{inverse} &  matrix inverse \\ 
\code{inverse\_spd} & symmetric, positive-definite matrix inverse \\ 
\code{log\_determinant} &  natural logarithm of determinant \\ 
\code{mdivide\_left\_tri\_low} &  lower-triangular matrix left division \\ 
\code{mdivide\_right\_tri\_low} &  lower-triangular matrix right division \\
\code{multiply\_lower\_tri\_self\_transpose} &  multiply
lower-triangular by transpose \\ 
\code{quad\_form} &  quadratic form vector-matrix multiplication \\ 
\code{rep\_matrix} &  replicate scalar, row vector or vector to matrix \\ 
\code{rep\_row\_vector} &  replicate scalar to row vector \\ 
\code{rep\_vector} &  replicate scalar to vector \\ 
\code{row} &  row of matrix \\ 
\code{rows} &  number of rows in matrix \\ 
\code{rows\_dot\_product} &  dot-product of rows of matrices \\ 
\code{rows\_dot\_self} &  dot-product of matrix with itself \\ 
\code{segment} &  sub-vector \\ 
\code{singular\_values} &  singular values of matrix \\ 
\code{size} &  number of entries in array or vector \\ 
\code{sub\_col} &  sub-column of matrix \\ 
\code{sub\_row} &  sub-row of matrix \\ 
\code{tail} &  tail of vector \\ 
\code{tcrossprod} &  matrix post-multiply by own transpose \\ 
\code{trace} &  trace of matrix \\ 
\code{trace\_gen\_quad\_form} & trace of generalized quadratic form \\ 
\code{trace\_quad\_form} &  trace of quadratic form
\end{tabular}
\end{center}
\caption{A large suite of matrix functions admits efficient 
  multivariate model implementation in \proglang{Stan}.}
  \label{matrix-functions.fig}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Function} & { Description} \\ \hline \hline
\code{bernoulli\_cdf} &  Bernoulli cdf \\ 
\code{bernoulli\_log} &  log Bernoulli pmf \\ 
\code{bernoulli\_logit\_log} &  logit-scale log Bernoulli pmf \\ 
\code{bernoulli\_rng} &  Bernoulli RNG \\ 
\code{beta\_binomial\_cdf} &  beta-binomial cdf \\ 
\code{beta\_binomial\_log} &  log beta-binomial pmf \\ 
\code{beta\_binomial\_rng} &  beta-binomial rng \\ 
\code{beta\_cdf} &  beta cdf \\ 
\code{beta\_log} &  log beta pdf \\ 
\code{beta\_rng} &  beta RNG \\ 
\code{binomial\_cdf} &  binomial cdf n \\ 
\code{binomial\_log} &  log binomial pmf \\ 
\code{binomial\_logit\_log} &  log logit-scaled binomial pmf \\ 
\code{binomial\_rng} &  binomail RNG \\ 
\code{categorical\_log} &  log categorical pmf \\ 
\code{categorical\_rng} &  categorical RNG \\ 
\code{cauchy\_cdf} &  Cauchy cdf \\ 
\code{cauchy\_log} &  log Cauchy pdf \\ 
\code{cauchy\_rng} &  Cauchy RNG \\ 
\code{chi\_square\_log} &  log chi-square pdf \\ 
\code{chi\_square\_rng} &  chi-square RNG \\ 
\code{dirichlet\_log} &  log Dirichlet pdf \\ 
\code{dirichlet\_rng} &  Dirichlet RNG \\ 
\code{double\_exponential\_log} &  log double-exponential (Laplace) pdf \\ 
\code{double\_exponential\_rng} &  double-exponential (Laplace) RNG \\ 
\code{exp\_mod\_normal\_cdf} &  exponentially modified normal cdf \\
\code{exp\_mod\_normal\_log} &  log exponentially modified normal pdf \\ 
\code{exp\_mod\_normal\_rng} &  exponentially modified normal RNG \\ 
\code{exponential\_cdf} &  exponentia cdf \\ 
\code{exponential\_log} &  log of exponential pdf \\ 
\code{exponential\_rng} &  exponential RNG \\ 
\code{gamma\_log} &  log gamma pdf \\ 
\code{gamma\_rng} &  gamma RNG \\ 
\code{gumbel\_cdf} &  Gumbel cdf \\ 
\code{gumbel\_log} &  log Gumbel pdf \\ 
\code{gumbel\_rng} &  Gumbel RNG\\ 
\code{hypergeometric\_log} &  log hypergeometric pmf \\ 
\code{hypergeometric\_rng} &  hypergeometric RNG \\ 
\code{inv\_chi\_square\_cdf} &  inverse chi-square cdf \\ 
\code{inv\_chi\_square\_log} &  log inverse chi-square pdf \\ 
\code{inv\_chi\_square\_rng} &  inverse chi-square RNG \\ 
\end{tabular}
\end{center}
\caption{Most common probability distributions have
  explicitly implemented in \proglang{Stan}.}\label{prob-functions.fig}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Function} & { Description} \\ \hline \hline
\code{inv\_gamma\_cdf} & inverse gamma cdf \\ 
\code{inv\_gamma\_log} &  log inverse gamma pdf \\ 
\code{inv\_gamma\_rng} &  inverse gamma RNG \\ 
\code{inv\_wishart\_log} &  log inverse Wishart pdf \\ 
\code{inv\_wishart\_rng} &  inverse Wishart RNG \\ 
\code{lkj\_corr\_cholesky\_log} &  log LKJ correlation, Cholesky-variate pdf \\ 
\code{lkj\_corr\_cholesky\_rng} &  LKJ correlation, Cholesky-variate RNG \\
\code{lkj\_corr\_log} &  log of LKJ correlation pdf \\ 
\code{lkj\_corr\_rng} &  LKJ correlation RNG \\ 
% \code{lkj\_cov\_log} &  {\it deprecated} \\ 
\code{logistic\_cdf} &  logistic cdf \\ 
\code{logistic\_log} &  log logistic pdf \\ 
\code{logistic\_rng} &  logistic RNG \\ 
\code{lognormal\_cdf} &  lognormal cdf \\ 
\code{lognormal\_log} &  log of lognormal pdf \\ 
\code{lognormal\_rng} &  lognormal RNG \\ 
\code{multi\_normal\_cholesky\_log} & log multi-normal Cholesky-parameterized pdf  \\ 
\code{multi\_normal\_log} &  log multi-normal pdf \\ 
\code{multi\_normal\_prec\_log} & log multi-normal precision-parameterized pdf \\ 
\code{multi\_normal\_rng} &  multi-normal RNG \\ 
\code{multi\_student\_t\_log} & log multi student-t pdf \\
\code{multi\_student\_t\_rng} &  multi student-t RNG \\ 
\code{multinomial\_log} &  log multinomial pmf \\ 
\code{multinomial\_rng} &  multinomial RNG \\ 
\code{neg\_binomial\_cdf} &  negative binomial cdf \\ 
\code{neg\_binomial\_log} &  log negative binomial pmf \\ 
\code{neg\_binomial\_rng} &  negative biomial RNG \\ 
\code{normal\_cdf} &  normal cdf \\ 
\code{normal\_log} &  log normal pdf (c.f. log lognormal pdf) \\ 
\code{normal\_rng} &  normal RNG \\ 
\code{ordered\_logistic\_log} &  log ordinal logistic pmf \\ 
\code{ordered\_logistic\_rng} &  ordinal logistic RNG \\ 
\code{pareto\_cdf} &  Pareto cdf \\ 
\code{pareto\_log} &  log Pareto pdf \\ 
\code{pareto\_rng} &  Pareto RNG \\ 
\code{poisson\_cdf} &  Poisson cdf \\ 
\code{poisson\_log} &  log Poisson pmf \\ 
\code{poisson\_log\_log} &  log Poisson log-parameter pdf \\ 
\code{poisson\_rng} &  Poisson RNG \\ 
\end{tabular}
\end{center}
\caption{Probability functions (continued)}\label{prob-functions-cont.fig}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Function} & { Description} \\ \hline \hline
\code{rayleigh\_log} & log Rayleigh pdf \\
\code{rayleigh\_rng} & Rayleigh RNG \\
\code{rayleigh\_cdf} & Rayleigh cdf \\
\code{scaled\_inv\_chi\_square\_cdf} &  scaled inverse-chi-square cdf \\ 
\code{scaled\_inv\_chi\_square\_log} &  log scaled inverse-chi-square pdf \\ 
\code{scaled\_inv\_chi\_square\_rng} &  scaled inverse-chi-square RNG \\ 
\code{skew\_normal\_cdf} &  skew-normal cdf \\
\code{skew\_normal\_log} &  log of skew-normal pdf \\
\code{skew\_normal\_rng} &  skew-normal RNG \\ 
\code{student\_t\_cdf} &  Student-t cdf \\
\code{student\_t\_log} &  log of Student-t pdf \\
\code{student\_t\_rng} &  Student-t RNG \\ 
\code{uniform\_log} &  log of uniform pdf \\ 
\code{uniform\_rng} &  uniform RNG \\ 
\code{weibull\_cdf} &  Weibull cdf \\ 
\code{weibull\_log} &  log of Weibull pdf \\ 
\code{weibull\_rng} &  Weibull RNG \\ 
\code{wishart\_log} &  log of Wishart pdf \\ 
\code{wishart\_rng} &  Wishart RNG \\ 
\end{tabular}
\end{center}
\caption{Probability functions (continued 2)}\label{prob-functions-cont-2.fig}
\end{figure}

% \nocite{WainwrightJordan:2008}
\nocite{R:2013}
\clearpage
%\bibliographystyle{jss}
\bibliography{stan-paper}

\end{document}

### 2013-12-24 src/docs/papers/stan-language/article.tex
updated with Andrew's and Marcus's comments

@@
 
-We did not set out to build \proglang{Stan} as it currently exists.
+We did not set out to build \proglang{Stan} as it currently exists.%
+%
+\footnote{\proglang{Stan}'s home page, \url{http://mc-stan.org/},
+  links to the Stan manual, example models including translations of
+  the BUGS examples and the models from \citep{GelmanHill:2007},
+  getting started instructions and full documentation for Stan's
+  interfaces for the command line (CmdStan), Python (PyStan), and
+  R (Rstan), and the source code repositories and issue trackers.}
+
 Our original goal was to apply full Bayesian inference to the sort of multilevel
@@
 by an order of magnitude.  In hindsight this isn't surprising, as the
-performance of the numerical simulation in HMC scales with the variation 
-of the parameter scales; fortunately HMC also provides a choice of metric,
-or mass matrix, to compensate for variations in the log posterior.
-In \proglang{Stan} 1.0 we introduced an adaptive 
-diagonal metric into our HMC implementations that allowed the parameter 
-scales to be normalized automatically, with the option of an adaptive dense 
-metric and a full automated standardization of the posterior in \proglang{Stan} 2.0.
+performance of the numerical simulation in HMC scales with the
+variation of the parameter scales; fortunately HMC also provides a
+choice of metric, or mass matrix, to compensate for variations in the
+log posterior.  In \proglang{Stan} 1.0 we introduced an adaptive
+diagonal metric into our HMC implementations that allowed the
+parameter scales to be normalized automatically, with the option of an
+adaptive dense metric and a full automated standardization of the
+posterior in \proglang{Stan} 2.0.
 
@@
 scales varies by location, such as hierarchical and latent models, can
-suffer.  Riemannian manifold Hamiltonian Monte Carlo (RHMC)
-\citep{GirolamiCalderhead:2011} introduces a location-dependent metric
-that can overcome these final hurdles, and a full implementation
-within Stan is currently under development.
+suffer.  Riemannian manifold Hamiltonian Monte Carlo (RHMC) introduces
+a location-dependent metric that can overcome these final hurdles
+\citep{GirolamiCalderhead:2011}; Stan has a prototype implementation
+of RHMC based on the metric of \citep{Betancourt:2012}, using a
+generalization of NUTS to Riemann manifolds \citep{Betancourt:2013}.
 
@@
   theta ~ beta(1,1);     // prior
-  for (n in 1:N) 
-    y[n] ~ bernoulli(theta);  // likelihood
+  y ~ bernoulli(theta);  // likelihood
 }
@@
 The model treats the observed binary data, \code{y[1],...,y[N]}, as
-independent and identically distributed, with success probability \code{theta}.  
-A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta} , although there is no
-special behavior for conjugate priors in \proglang{Stan}.  This model
-is available in the \proglang{Stan} source distribution in
+independent and identically distributed, with success probability
+\code{theta}.  The vectorized likelihood statement can also be coded
+using a loop as in BUGS, although it will run more slowly than the
+vectorized form:
+%
+\begin{Code}
+for (n in 1:N)
+  y[n] ~ bernoulli(theta);
+\end{Code}
+%
+A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta},
+although there is no special behavior for conjugate priors in
+\proglang{Stan}.  The prior could be dropped from the model altogether
+because parameters start with uniform distributions on their support,
+here constrained to be between 0 and 1 in the parameter declaration
+for \code{theta}.  This model is available in the \proglang{Stan}
+source distribution in
 \code{src/models/basic\_estimators/bernoulli.stan}.
@@
 \pkg{Coda} \citep{PlummerEtAl:2006}, first splitting each chain in half to diagnose
-montonically increasing or decreasing samples; \citep{GelmanEtAl:2013}
+nonstationary chains; \citep{GelmanEtAl:2013}
 and \citep{Stan:2013} provide detailed definitions.
@@
 
-\subsection{The ``Matt trick'' for hierarchical models}
+\subsection{The non-centered parameterization for hierarchical models}
 
@@
 %
-With the direct parameterization, a change in \code{sigma} is
-amplified by the lower-level parameters and introduces a large
-change in density across the posterior.  Unfortunately the 
-expected density variation of an HMC transition is limited%
-%
-\footnote{The limited density variation is a ultimately consequence
-of a constant metric.  An additional benefit of RMHMC are transitions
+With this parameterization, the \code{beta[k]} are centered around
+\code{mu}.  As a result, a change in \code{sigma} is amplified by the
+lower-level parameters and introduces a large change in density across
+the posterior.  Unfortunately the expected density variation of an HMC
+transition is limited%
+%
+\footnote{The limited density variation is a ultimately consequence of
+  a constant metric.  An additional benefit of RHMC are transitions
   that can cover much larger variations in density, making it uniquely
-suited to these models.}
-%
-and many transitions are required to explore the full posterior.
-The resulting sampler devolves into a random walk with high autocorrelations.
-
-The following reparameterization employs the ``Matt trick'',
-%
-\footnote{The reparameterization is named after Matt Hoffman, who
-  introduced it during modeling discussions in the \proglang{Stan}
-  group meetings.  A similar reparameterization by \citep{Choo:2000}
-  is discussed in \citep[Section~5.4.5]{Neal:2011}.}
+  suited to these models; see \citep{Neal:2011}.}
+%
+and many transitions are required to explore the full posterior.  The
+resulting sampler devolves into a random walk with high
+autocorrelations.
+
+\cite{BetancourtGirolami:2013} discuss for more
+analysis of centered vs.\ non-centered parameterizations of
+hierarchical models.
+
+The following reparameterization employs a non-centered
+parameterization (see \citep{BetancourtGirolami:2013} for an analysis
+of the benefits of non-centered parameterizations).
 %
@@
 
+Stan follows the \proglang{C++} convention of separating words in
+variable names using underbars (\code{\_}), rather than dots
+(\code{.}), as used in \proglang{R} and \proglang{BUGS}, or camel case
+as used in Java.
+
 \subsection{Control structures}
@@
 
+In addition to special functions, Stan includes distributions with
+alternative parameterizations, such as \code{bernoulli_logit}, which
+takes a parameter on the log odds (i.e., logit) scale.  This allows a
+more concise notation for generalized linear models as well as more
+efficient and arithmetically stable execution.
+
 \subsection{Matrix and linear algebra functions}
@@
 indexing uses the usual square brackets (\code{[ ]}) operator, and
-begins from 1.  \proglang{R}, \proglang{MATLAB}, and \proglang{BUGS},
-but not of \proglang{C++}, \proglang{Python}, or \proglang{Java}.
-If \code{Sigma} is a matrix, then \code{Sigma[n]} is the \code{n}-th
-row of \code{Sigma}.
+begins indexing from 1.  If \code{Sigma} is a matrix, then
+\code{Sigma[n]} is the \code{n}-th row of \code{Sigma}.
 
@@
 proportion, which becomes an additive constant on the log scale.
-Constancy here refers to being a numeric literal such as \code{1}, a
-constant function such as \code{pi()}, a variable declared as data or
-transformed data, or an expression involving only other constants.
+Constancy here refers to being a numeric literal such as \code{1} or
+\code{0.5}, a constant function such as \code{pi()}, data and
+transformed data variables, or a function that only depends on
+literals, constant functions or data variables.
+
 Non-constants include parameters, transformed parameters, local
@@
 %
-For example, calling \code{normal\_log(y,0,1)}, where \code{y} is a
-parameter and \code{0} and \code{1} are constants, does not bother
-computing the $\log \sigma$ term in the log density because $\sigma=1$
-is a constant.  On the other hand, the $y - 0$ term is calculated and
-is not a constant because $y$ is not a constant.
+For example, executing \code{normal\_log(y,0,sigma)} only evaluates
+\code{log(sigma)} if \code{sigma} is a parameter, transformed
+parameter, or a local variable in the transformed parameters or model
+block; that is, \code{log(sigma)} is not evaluated if \code{sigma} is
+constant as defined above.
 
@@
 \subsection{Optimizers}
+
+In addition to performing full Bayesian inference via posterior
+sampling, Stan also can perform optimization (i.e., computation of the
+posterior mode).  We are currently working on implementing other
+optimization-based inference approaches including variational Bayes,
+expectation propagation, and and marginal inference using approximate
+integration.  All these algorithms require optimization steps.
 

### 2013-12-25 src/docs/papers/stan-language/article.tex
updated first half of paper

@@
 \documentclass[article]{jss}
+
+\renewcommand{\textfraction}{0.07}
+\renewcommand{\topfraction}{0.9}	% max fraction of floats at top
 
@@
   imperatively defines a log probability function over parameters
-  conditioned on specified data and constants.
-
-  Full Bayesian inference is supported through Markov chain Monte
-  Carlo (MCMC) methods such as the No-U-Turn sampler (NUTS), an
-  adaptive form of Hamiltonian Monte Carlo (HMC) sampling.  Penalized
-  maximum likelihood estimates are calculated using optimization
-  methods such as BFGS.
+  conditioned on specified data and constants.  As of version 2.1,
+  \proglang{Stan} provides full Bayesian inference for continuous-variable
+  models through Markov chain Monte Carlo (MCMC) methods such as
+  the No-U-Turn sampler (NUTS), an adaptive form of Hamiltonian Monte
+  Carlo (HMC) sampling.  Penalized maximum likelihood estimates are
+  calculated using optimization methods such as BFGS.
 
@@
 
-  Stan can be viewed as a package for performing full Bayesian
-  inference for continuous-variable models, and as a platform for
-  computing log-densities and their gradients and Hessians, which can
-  be used in alternative algorithms such as variational Bayes,
-  expectation propagation, and marginal inference using approximate
-  integration.  To this end, Stan is set up so that the densities,
-  gradients, and Hessians, along with intermediate quantities of the
-  algorithm such as acceptance probabilities, are easily accessible.
-
-  % \proglang{Stan} can be called from the command line, through
-  % \proglang{R} using the \pkg{RStan} package, or through
-  % \proglang{Python} using the \pkg{PyStan} package.  All three
-  % interfaces support sampling or optimization-based inference and
-  % analysis.
+  \proglang{Stan} is also a platform for computing log-densities and
+  their gradients and Hessians, which can be used in alternative
+  algorithms such as variational Bayes, expectation propagation, and
+  marginal inference using approximate integration.  To this end, Stan
+  is set up so that the densities, gradients, and Hessians, along with
+  intermediate quantities of the algorithm such as acceptance
+  probabilities, are easily accessible.
+
+  \proglang{Stan} can be called from the command line, through
+  \proglang{R} using the \pkg{RStan} package, or through
+  \proglang{Python} using the \pkg{PyStan} package.  All three
+  interfaces support sampling or optimization-based inference and
+  analysis, and \pkg{RStan} and \pkg{PyStan} also provide access to
+  log probabilities, gradients, Hessians, and I/O transforms.
 }
@@
   getting started instructions and full documentation for Stan's
-  interfaces for the command line (CmdStan), Python (PyStan), and
-  R (Rstan), and the source code repositories and issue trackers.}
-
+  interfaces for the command line shell (\pkg{CmdStan}), \proglang{Python}
+  (\pkg{PyStan}), and \proglang{R} (\pkg{RStan}), and the source code repositories
+  and issue trackers.}
+%
 Our original goal was to apply full Bayesian inference to the sort of multilevel
@@
 with interacted predictors.  We finally realized we needed a better sampler, 
-not a more efficient implementation.
+not a more efficient implementation of Gibbs sampling.
 
@@
 
-We were hearing more about HMC (Hamiltonian Monte Carlo), which
-appeared promising but was also problematic in that the Hamiltonian
-dynamics simulation requires the gradient of the log posterior.
-Although possible, computing these gradients by hand on a model-by-model basis
-is very tedious and error prone.  That is when we discovered
-reverse-mode algorithmic differentiation, which, given a
-templated \proglang{C++} function for the log posterior,
-automatically computes a proper analytic gradient up to machine
-precision accuracy in only a few multiples of the cost to evaluate the
-log probability function itself.  We explored existing algorithmic
-differentiation packages with open licenses such as \pkg{RAD}
-\citep{Gay:2005} and its repackaging in the Sacado module of the
-\pkg{Trilinos} toolkit, and the \pkg{CppAD} package of the COIN-OR
-toolkit (see \citep{BellBurke:2008}).  Both packages, however, supported
-few special functions (e.g., probability functions, log gamma,
-inverse logit) or linear algebra operations (e.g., Cholesky
-decompositions, matrix division), and neither are easily or modularly
-extensible.
+We were hearing more about Hamiltonian Monte Carlo (HMC) sampling,
+which appeared promising but was also problematic in that the
+Hamiltonian dynamics simulation requires the gradient of the log
+posterior.  Although not difficult in theory for most functions,
+computing these gradients by hand on a model-by-model basis is very
+tedious and error prone.  That is when we discovered reverse-mode
+algorithmic differentiation, which, given a templated \proglang{C++}
+function for the log posterior, automatically computes a proper
+analytic gradient up to machine precision accuracy in only a few
+multiples of the cost to evaluate the log probability function itself.
+We explored existing algorithmic differentiation packages with open
+licenses such as \pkg{RAD} \citep{Gay:2005} and its repackaging in the
+Sacado module of the \pkg{Trilinos} toolkit, and the \pkg{CppAD}
+package of the COIN-OR toolkit \citep{BellBurke:2008}.  Both
+packages, however, supported few special functions (e.g., probability
+functions, log gamma, inverse logit) or linear algebra operations
+(e.g., Cholesky decompositions, matrix division), and neither are
+easily or modularly extensible.
 
@@
 the probability functions from a package like \pkg{Boost} because they
-weren't templated generally enough across all the arguments.  Rather
+weren't templated generally enough across all arguments.  Rather
 than pay the price of promoting floating point values to algorithmic
@@
 operations and linear algebra functions.  \pkg{Eigen} makes extensive
-use of expression templates for lazy evaluation and the CRTP
-(curiously recurring template pattern) to implement concepts without
+use of expression templates for lazy evaluation and the curiously
+recurring template pattern (CRTP) to implement concepts without
 virtual function calls; \cite{VandevoordeJosuttis:2002} provide a
 complete description of template metaprogramming techniques.
-Unfortunately, we
-ran into the same problem with Eigen as with the existing probability
-libraries---it doesn't support mixed operations of algorithmic
-differentiation variables and primitives like \code{double}.  Although
-we initially began by promoting floating-point vectors to algorithmic
-differentiation variables, we later completely rewrote our matrix
-library to efficiently compute derivatives using rules described by
-\cite{Giles:2008}.
+Unfortunately, we ran into the same problem with Eigen as with the
+existing probability libraries---it doesn't support mixed operations
+of algorithmic differentiation variables and primitives like
+\code{double}.  Although we initially began by promoting
+floating-point vectors to algorithmic differentiation variables, we
+later completely rewrote mixed-mode operations like multiplying a data
+matrix by a parameter vector.  To compute derivatives of matrix
+operations such as division and inverse, we implemented specialized
+derivatives using the rules described by \cite{Giles:2008}.
 
@@
 (application programming interface).  Seeing how well this all worked,
-we set our sights on the generality and ease of use of
-\proglang{BUGS} and designed a modeling language in which
-statisticians could write their models in familiar notation that could
-be transformed to efficient \proglang{C++} code before being compiled into
-an efficient executable program.  This paper is primarily about this
-language and how it specifies models.
+we set our sights on the generality and ease of use of \proglang{BUGS}
+and designed a modeling language in which statisticians could write
+their models in familiar notation that could be transformed to
+efficient \proglang{C++} code before being compiled into an efficient
+executable program.  Although we started with the target of specifying
+directed graphical models, our first implementation turned out to be a
+more general imperative form of specifying log probability functions;
+\proglang{BUGS} models could be translated line for line, but the
+\proglang{Stan} language also supports much richer imperative
+constructs like conditionals, while loops, and local variables.
+This paper is primarily about the \proglang{Stan} language and how
+it can be used to specify statistical models.
 
@@
 to two tuning parameters, the discretization interval (i.e., step
-size) and the total simulation time (i.e., number of steps).  The interval
-size parameter could be tuned during warm-up based on Metropolis
-rejection rates, but the number of steps proved difficult to tune without
-sacrificing the detailed balance of the sampler.  This led to the
-development of the NUTS (No U-Turn) sampler
-\citep{HoffmanGelman:2011}.  Roughly speaking, NUTS builds a tree
-of possible samples by randomly simulating both forwards and backwards
-in time until the combined trajectory turns back on itself.  Once the trajectory
-has terminated, a new sample is drawn from the tree.
+size) and the total simulation time (i.e., number of steps).  The
+interval size parameter could be tuned during warm-up based on
+Metropolis rejection rates, but the number of steps proved difficult
+to tune without sacrificing the detailed balance of the sampler.  This
+led to the development of the No U-Turn (NUTS) sampler
+\citep{HoffmanGelman:2011}.  Roughly speaking, NUTS builds a tree of
+possible samples by randomly simulating Hamiltonian dynamics both
+forwards and backwards in time until the combined trajectory turns
+back on itself.  Once the trajectory has terminated, a new sample is
+drawn from the tree.
 
@@
 priors and lack of posterior correlations make it an ideal candidate
-for efficient Gibbs sampling.
-
-Realizing that we were doing redundant calculations, we wrote a
-vectorized form of the normal distribution for multiple variates with
-the same mean and scale, which sped things up a bit. At the same time,
-we introduced some simple template metaprograms to remove the
-calculation of constant terms in the log probability.  Performance was
-lacking until we finally figured out how to both eliminate redundant 
-calculations and partially evaluate the gradients using
-a combination of expression templates and metaprogramming.
+for efficient Gibbs sampling.  Realizing that we were doing redundant
+calculations, we wrote a vectorized form of the normal distribution
+for multiple variates with the same mean and scale, which sped things
+up a bit. At the same time, we introduced some simple template
+metaprograms to remove the calculation of constant terms in the log
+probability.  Performance was lacking until we finally figured out how
+to both eliminate redundant calculations and partially evaluate the
+gradients using a combination of expression templates and
+metaprogramming.
 
@@
 variation of the parameter scales; fortunately HMC also provides a
-choice of metric, or mass matrix, to compensate for variations in the
-log posterior.  In \proglang{Stan} 1.0 we introduced an adaptive
-diagonal metric into our HMC implementations that allowed the
-parameter scales to be normalized automatically, with the option of an
-adaptive dense metric and a full automated standardization of the
-posterior in \proglang{Stan} 2.0.
+choice of metric, in our case a Euclidean metric with a constant mass
+matrix, to compensate for variations in the log posterior.  In
+\proglang{Stan} 1.0 we introduced an adaptive diagonal metric into our
+HMC implementations that allowed the parameter scales to be normalized
+automatically, with the option of an adaptive dense metric and a full
+automated standardization of the posterior in \proglang{Stan} 2.0.
 
@@
 \citep{GirolamiCalderhead:2011}; Stan has a prototype implementation
-of RHMC based on the metric of \citep{Betancourt:2012}, using a
-generalization of NUTS to Riemann manifolds \citep{Betancourt:2013}.
+of RHMC based on the SoftAbs metric of \citep{Betancourt:2012}, using
+a generalization of NUTS to Riemannian manifolds \citep{Betancourt:2013}.
 
@@
 This section describes the use of \proglang{Stan} from the command
-line for estimating (i.e., sampling from) the posterior of a Bayesian
-model.
+line for estimating a Bayesian model using both MCMC sampling for full
+Bayesian inference and optimization to provide a point estimate at the
+posterior mode.
 
@@
 Figure~\ref{bernoulli-model.fig} provides an implementation of such a
-model in \proglang{Stan}.
+model in \proglang{Stan}.%
+%
+\footnote{This model is available in the \proglang{Stan} source
+  distribution in \nolinkurl{src/models/basic\_estimators/bernoulli.stan}.}
 %
@@
 data {
-  int<lower=0> N; 
-  int<lower=0,upper=1> y[N];
+  int<lower=0> N;                  // N >= 0
+  int<lower=0,upper=1> y[N];       // y[n] in { 0, 1 }
 }
 parameters {
-  real<lower=0,upper=1> theta;
+  real<lower=0,upper=1> theta;     // theta in [0, 1]
 }
@@
 \end{Code}
-\caption{Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
+\caption{\it Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
 \end{figure}
@@
 here constrained to be between 0 and 1 in the parameter declaration
-for \code{theta}.  This model is available in the \proglang{Stan}
-source distribution in
-\code{src/models/basic\_estimators/bernoulli.stan}.
+for \code{theta}.
 
@@
 %
-\footnote{A JSON interface for structured data input and output is
+\footnote{A \pkg{JSON} interface for structured data input and output is
   currently under development.}
@@
 For example, 10 observations for the
-model in Figure~\ref{bernoulli-model.fig} could be encoded as
+model in Figure~\ref{bernoulli-model.fig} could be encoded as%
+%
+\footnote{This data
+file is provided with the \proglang{Stan} distrbution in file
+\nolinkurl{src/models/basic\_estimators/bernoulli.R.stan}.}
 %
@@
 This defines the contents of two variables, an integer \code{N} and a
-10-element integer array \code{y}.  In \pkg{RStan} and \pkg{PyStan},
-data can also be passed directly through memory without the need to
-read or write to a file.  This data file is provided with the
-\proglang{Stan} distrbution in file
-\code{src/models/basic\_estimators/bernoulli.R.stan}.
+10-element integer array \code{y}.  The variable \code{N} is declared
+in the data block of the program as being an integer greater than or
+equal to zero;  the variable \code{y} is declared as an integer array of size
+\code{N} with entries between 0 and 1 inclusive.
+
+In \pkg{RStan} and \pkg{PyStan}, data can also be passed directly
+through memory without the need to read or write to a file.
 
@@
 
-\subsection{Running the model}
-
-\subsubsection{Command to run the model}
+\subsection{Running the sampler}
+
+\subsubsection{Command to sample from the model}
 
@@
 \begin{CodeChunk}
-{\small
 \begin{CodeInput}
 > cd $stan/src/models/basic_estimators
-> ./bernoulli --data=bernoulli.data.R --samples=samp1.csv --chain_id=1 --seed=7386
+> ./bernoulli sample data file=bernoulli.data.R
 \end{CodeInput}
-}
 \end{CodeChunk}
@@
 
-\subsubsection{Terminal output}
+\subsubsection{Terminal output from sampler}
 
@@
 \begin{Code}
-STAN SAMPLING COMMAND
-data = bernoulli.data.R
-init = random initialization
-init tries = 1
-samples = samp1.csv
-append_samples = 0
-save_warmup = 0
-seed = 7386 (user specified)
-chain_id = 1 (user specified)
-iter = 2000
-warmup = 1000
-thin = 1
-equal_step_sizes = 0
-nondiag_mass = 0
-leapfrog_steps = -1
-max_treedepth = 10
-epsilon = -1
-epsilon_pm = 0
-delta = 0.5
-gamma = 0.05
+ method = sample (Default)
+   sample
+     num_samples = 1000 (Default)
+     num_warmup = 1000 (Default)
+     save_warmup = 0 (Default)
+     thin = 1 (Default)
+     adapt
+       engaged = 1 (Default)
+       gamma = 0.050000000000000003 (Default)
+       delta = 0.80000000000000004 (Default)
+       kappa = 0.75 (Default)
+       t0 = 10 (Default)
+       init_buffer = 75 (Default)
+       term_buffer = 50 (Default)
+       window = 25 (Default)
+     algorithm = hmc (Default)
+       hmc
+         engine = nuts (Default)
+           nuts
+             max_depth = 10 (Default)
+         metric = diag_e (Default)
+         stepsize = 1 (Default)
+         stepsize_jitter = 0 (Default)
+ id = 0 (Default)
+ data
+   file = bernoulli.data.R
+ init = 2 (Default)
+ random
+   seed = 4294967295 (Default)
+ output
+   file = output.csv (Default)
+   diagnostic_file =  (Default)
+   refresh = 100 (Default)
+
+Gradient evaluation took 4e-06 seconds
+1000 transitions using 10 leapfrog steps per transition would take 0.04 seconds.
+Adjust your expectations accordingly!
+
 Iteration:    1 / 2000 [  0%]  (Warmup)
-Iteration:   10 / 2000 [  0%]  (Warmup)
+Iteration:  100 / 2000 [  5%]  (Warmup)
 ...
-Iteration: 1990 / 2000 [ 99%]  (Sampling)
+Iteration: 1000 / 2000 [ 50%]  (Warmup)
+Iteration: 1001 / 2000 [ 50%]  (Sampling)
+...
 Iteration: 2000 / 2000 [100%]  (Sampling)
 
-# Elapsed Time: 0.011269 seconds (Warm-up)
-#               0.014105 seconds (Sampling)
-#               0.025374 seconds (Total)
+  Elapsed Time: 0.00932 seconds (Warm-up)
+                0.016889 seconds (Sampling)
+                0.026209 seconds (Total)
 \end{Code}
 %
-A description of all parameters is available with the \code{{-}{-}help}
-option; they provide control over input, output, and the configuration
-of the sampler.
+The sampler configuration parameters are echoed, here they are all
+default values other than the data file.  These parameters may be set
+on the command line.
+
+\subsubsection{Help}
+
+A description of all configuration parameters including default values
+and constraints is available by executing
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+> ./bernoulli help-all
+\end{CodeInput}
+\end{CodeChunk}
+%
+The sampler and its configuration are described at greater length in
+the manual.
 
@@
 
-The output CSV file starts with a summary of the parameters for the
-run.  
+The output CSV file, written by default to \code{output.csv}, starts
+with a summary of the configuration parameters for the run.
 %
 \begin{Code}
-# Samples Generated by Stan
-#
-# stan_version_major=1
-# stan_version_minor=3
-# stan_version_patch=0
-# model=bernoulli_model
-# data=bernoulli.data.R
-# init=random initialization
-# append_samples=0
-# save_warmup=0
-# seed=7386
-# chain_id=1
-# iter=2000
-# warmup=1000
-# thin=1
-# nondiag_mass=0
-# equal_step_sizes=0
-# leapfrog_steps=-1
-# max_treedepth=10
-# epsilon=-1
-# epsilon_pm=0
-# delta=0.5
-# gamma=0.05
-# algorithm=NUTS with a diagonal Euclidean metric
+# stan_version_major = 2
+# stan_version_minor = 1
+# stan_version_patch = 0
+# model = bernoulli_model
+# method = sample (Default)
+#   sample
+#     num_samples = 1000 (Default)
+#     num_warmup = 1000 (Default)
+#     save_warmup = 0 (Default)
+#     thin = 1 (Default)
+#     adapt
+#       engaged = 1 (Default)
+#       gamma = 0.050000000000000003 (Default)
+#       delta = 0.80000000000000004 (Default)
+#       kappa = 0.75 (Default)
+#       t0 = 10 (Default)
+#       init_buffer = 75 (Default)
+#       term_buffer = 50 (Default)
+#       window = 25 (Default)
+#     algorithm = hmc (Default)
+#       hmc
+#         engine = nuts (Default)
+#           nuts
+#             max_depth = 10 (Default)
+#         metric = diag_e (Default)
+#         stepsize = 1 (Default)
+#         stepsize_jitter = 0 (Default)
+# id = 0 (Default)
+# data
+#   file = bernoulli.data.R
+# init = 2 (Default)
+# random
+#   seed = 847896134
+# output
+#   file = output.csv (Default)
+#   diagnostic_file =  (Default)
+#   refresh = 100 (Default)
 \end{Code}
 %
-It continues with a header indicating the columns of the output.
+\proglang{Stan}'s behavior is fully specified by these configuration
+parameters.  By using the same version of \proglang{Stan} and these
+configuration parameters, exactly the same output file can be
+reproduced.  The pseudo-random numbers generated by the sampler are
+fully determined by the seed (here randomly generated based on the
+time of the run, with value \code{847896134}) and the identifier (here
+\code{0}).  The identifier is used to advance the underlying
+pseudo-random number generator a sufficient number of values that
+using multiple chains with the same seed and different identifiers
+will draw from different subsequences of the pseudo-random number
+stream determined by the seed.
+
+The output contiues with a CSV header naming the columns of the
+output. For the default NUTS sampler in Stan 2.1.0, these are
 %
 \begin{Code}
-lp__,accept_stat__,stepsize__,treedepth__,theta
+lp__,accept_stat__,stepsize__,treedepth__,n_divergent__,theta
 \end{Code}
@@
 additive constant).  The column headed by \code{accept\_stat\_\_}
-provides the Metropolis acceptance statistic for each iteration.%
-%
-\footnote{Acceptance is the usual notion for a Metropolis-Hastings
-  sampling such as HMC.  For NUTS, the acceptance statistic is 
-  defined as the average acceptance probabilities of all possible
-  samples in the proposed tree.}
+provides the Metropolis/slice acceptance statistic for each
+iteration.%
+%
+\footnote{Acceptance is the usual notion for a Metropolis sampler such
+  as HMC \citep{MetropolisEtAl:1953}.  For NUTS, the acceptance
+  statistic is defined as the average acceptance probabilities of all
+  possible samples in the proposed tree; NUTS itself uses a slice
+  sampling algorithm for rejection \citep{Neal:2003}.}
 %
@@
 
-Next, the results of adaptation are printed.
+Next, the results of adaptation are printed as comments.
 %
@@
 # Adaptation terminated
-# Step size = 2.10898
+# Step size = 0.783667
 # Diagonal elements of inverse mass matrix:
-# 0.465584
+# 0.517727
 \end{Code}
 %
-By default, the execution above uses a diagonal mass matrix, which
-with one parameter is a single element.
+By default, \proglang{Stan} uses the NUTS sampler with a diagonal mass
+matrix.  The mass matrix is estimated, roughly speaking, by
+regularizing the sample covariance of the latter half of the warmup
+samplers; see \citep{Stan:2013} for full details.  A dense mass matrix
+may also be estimated, or the mass matrix may be set to the unit matrix.
 
@@
 \begin{Code}
--7.04574,0.189209,2.10898,0,0.353589
--7.04574,0.00162111,2.10898,0,0.353589
--6.77448,1,2.10898,0,0.222005
+-7.19297,1,0.783667,1,0,0.145989
+-8.2236,0.927238,0.783667,1,0,0.0838792
 ...
--7.89094,0.346733,2.10898,0,0.46154
--6.93412,1,2.10898,0,0.330879
--6.93412,0.115548,2.10898,0,0.330879
-
-# Elapsed Time: 0.011269 seconds (Warm-up)
-#               0.014105 seconds (Sampling)
-#               0.025374 seconds (Total)
+-7.48489,0.738509,0.783667,0,0,0.121812
+-7.40361,0.995299,0.783667,1,0,0.407478
+-9.49745,0.771026,0.783667,2,0,0.0490488
+-9.11119,1,0.783667,0,0,0.0572588
+-7.20021,0.979883,0.783667,1,0,0.14527
+
+#  Elapsed Time: 0.010849 seconds (Warm-up)
+#                0.01873 seconds (Sampling)
+#                0.029579 seconds (Total)
 \end{Code}
 %
-
-\subsection{Output analysis}
+It is evident from the values sampled for \code{theta} in the last
+column that there is a high degree of posterior uncertainty in the
+estimate of \code{theta} from the ten data points in the data file.
+
+The log probabilities reported in the first column include not only
+the model log probabilities but also the Jacobian adjustment resulting
+from the transformation of the variables to unconstrained space.
+Here, that is the absolute derivative of the inverse logistic
+function;  see \citep{Stan:2013} for full details on all of the
+transforms and their Jacobians.
+
+\subsection{Sampler output analysis}
 
 Before performing output analysis, we recommend generating multiple
-independent chains in order to more effectively monitor convergence.
-Three more chains of samples can be created as follows.
+independent chains in order to more effectively monitor convergence;
+see \citep{GelmanRubin:1992} for more analysis.  Three more chains of
+samples can be created as follows.
 %
 \begin{CodeChunk}
-{\small
 \begin{CodeInput}
-> ./bernoulli --data=bernoulli.data.R --samples=samp2.csv --chain_id=2 --seed=7386
-> ./bernoulli --data=bernoulli.data.R --samples=samp3.csv --chain_id=3 --seed=7386
-> ./bernoulli --data=bernoulli.data.R --samples=samp4.csv --chain_id=4 --seed=7386
+./bernoulli sample data file=bernoulli.data.R random seed=847896134 \
+            id=1 output file=output1.csv
+./bernoulli sample data file=bernoulli.data.R random seed=847896134 \
+            id=2 output file=output2.csv
+./bernoulli sample data file=bernoulli.data.R random seed=847896134 \
+            id=3 output file=output3.csv
 \end{CodeInput}
-}
 \end{CodeChunk}
 %
+These calls illustrate how additional parameters are specified
+directly on the command line following the hierarchy given in the
+output.  The backslash (\code{\textbackslash}) at the end of each line indicates
+that the command continues on the last line;  a caret (\code{\textasciicircum})
+should be used in Windows.
+
 The chains can be safely run in parallel under different processes;
@@
 \begin{CodeInput}
-> ls samp*.csv
+> ls output*.csv
 \end{CodeInput}
 \begin{CodeOutput}
-samp1.csv	samp2.csv	samp3.csv	samp4.csv
+output.csv	output1.csv	output2.csv	output3.csv
 \end{CodeOutput}
 \begin{CodeInput}
-> ../../../bin/print samp*.csv
+> $stan/bin/print output*.csv
 \end{CodeInput}
@@
 \begin{figure}
-\small
 \begin{Code}
 Inference for Stan model: bernoulli_model
-4 chains: each with iter=(1000,1000,1000,1000); warmup=(0,0,0,0); thin=(1,1,1,1); 
-4000 iterations saved.
-
-Warmup took (0.012, 0.010, 0.011, 0.011) seconds, 0.046 seconds total
-Sampling took (0.016, 0.014, 0.014, 0.015) seconds, 0.060 seconds total
-
-                  mean   se   sd     2.5%      50%    97.5%   n_eff n_eff/s    Rhat
-lp__          -7.3e+00  0.0  0.8 -9.3e+00 -7.0e+00 -6.7e+00 1.1e+03 1.9e+04 1.0e+00
-accept_stat__  5.3e-01  0.1  0.4  1.1e-06  5.2e-01  1.0e+00 3.7e+03 6.1e+04 1.0e+00
-stepsize__     2.1e+00  0.1  0.2  2.0e+00  2.1e+00  2.1e+00 2.0e+00 3.3e+01 1.3e+13
-treedepth__    3.6e-02  0.0  0.2  0.0e+00  0.0e+00  1.0e+00 2.9e+03 4.9e+04 1.0e+00
-theta          2.5e-01  0.0  0.1  6.0e-02  2.4e-01  4.9e-01 2.1e+03 3.5e+04 1.0e+00
-
-Samples were drawn using NUTS with a diagonal Euclidean metric.
-For each parameter, n_eff is a crude measure of effective sample size,
-and Rhat is the potential scale reduction factor on split chains (at 
-convergence, Rhat=1).
+4 chains: each with iter=(1000,1000,1000,1000); warmup=(0,0,0,0); 
+                    thin=(1,1,1,1); 4000 iterations saved.
+
+Warmup took (0.0108, 0.0130, 0.0110, 0.0110) seconds, 0.0459 seconds total
+Sampling took (0.0187, 0.0190, 0.0168, 0.0176) seconds, 0.0722 seconds total
+
+                 Mean      MCSE  StdDev         5%    50%    95%
+lp__            -7.28  1.98e-02   0.742  -8.85e+00  -6.99  -6.75
+accept_stat__   0.909  4.98e-03   0.148   5.70e-01  0.971   1.00
+stepsize__      0.927  7.45e-02   0.105   7.84e-01   1.00   1.05
+treedepth__     0.437  1.03e-02   0.551   0.00e+00  0.000   1.00
+n_divergent__   0.000  0.00e+00   0.000   0.00e+00  0.000  0.000
+theta           0.254  3.25e-03   0.122   7.58e-02  0.238  0.479
+
+                N_Eff  N_Eff/s     R_hat
+lp__             1404    19447  1.00e+00
+accept_stat__     887    12297  1.02e+00
+stepsize__       2.00     27.7  5.56e+13
+treedepth__      2856    39572  1.01e+00
+n_divergent__    4000    55424       nan
+theta            1399    19382  1.00e+00
 \end{Code}
-\caption{Output of \code{bin/print} for the Bernoulli estimation model in
+\caption{\it Output of \code{bin/print} for the Bernoulli estimation model in
   Figure~\ref{bernoulli-model.fig}.}\label{print-output.fig}
@@
 Each row of the output summarizes a different value whose name is
-provided in the first column.  The analysis includes the posterior
-mean (\code{mean}) and standard deviation (\code{sd}).  The median
-(\code{50\%}) and 95\% posterior interval (\code{2.5\%}, \code{97.5\%})
-are also displayed; the \code{bin/print} command also prints the 50\%
-interval boundaries, but these are ellided for space in the above
-output.
-
-The estimated potential scale reduction statistic $\hat{R}$
-(\code{Rhat}) value of 1.0 for the parameter of interest \code{theta}
-is consistent with convergence.  \proglang{Stan} uses a more
-conservative version of $\hat{R}$ than is usual in packages such as
-\pkg{Coda} \citep{PlummerEtAl:2006}, first splitting each chain in half to diagnose
-nonstationary chains; \citep{GelmanEtAl:2013}
-and \citep{Stan:2013} provide detailed definitions.
-
-The estimated number of effective samples per parameter
-(\code{n\_eff}) is more than 2000, far more than we actually need for
-inference.  As with $\hat{R}$, the number of effective samples is
-estimated conservatively using both cross-chain and within-chain
-estimates; \citep{GelmanEtAl:2013} and the reference manual provide
-details.
-
-The MCMC standard error (\code{se}) is estimated by the posterior
-standard deviation estimate divided by the square root of the
-estimated number of effective samples (\code{sd / sqrt(n\_eff)}).  In
-this example, the MCMC standard error is small relative to the
-posterior standard deviation, another sign that a sufficient number
-of samples have been drawn.
+provided in the first column.  These correspond to the columns in the
+output CSV files. The analysis includes estimates of the posterior
+mean (\code{Mean}) and standard deviation (\code{StdDev}).  The median
+(\code{50\%}) and 90\% posterior interval (\code{5\%}, \code{95\%})
+are also displayed.
+
+The remaining columns in the output provide an analysis of the
+sampling and its efficiency.  The convergence diagnostic that is built
+into the \code{bin/print} command is the estimated potential scale
+reduction statistic $\hat{R}$ (\code{Rhat}); its value should be close
+to 1.0 when the chains have all converged to the same stationary
+distribution.  \proglang{Stan} uses a more conservative version of
+$\hat{R}$ than is usual in packages such as \pkg{Coda}
+\citep{PlummerEtAl:2006}, first splitting each chain in half to
+diagnose nonstationary chains; see \citep{GelmanEtAl:2013} and
+\citep{Stan:2013} for detailed definitions.
+
+The column \code{N\_eff} is the number of effective samples in a
+chain.  Because MCMC methods produce correlated samples in each chain,
+estimates such as posterior means are not as accurate as they would be
+with truly independent samples.  To number of effective samples is an
+estimate of the number of independent samples that would lead to the
+same accuracy.  The Monte Carlo standard error (\code{MCSE}) is an
+estimate of the error in estimating the posterior mean based on
+dividing the posterior standard deviation estimate by the square root
+of the number of effective samples (\code{sd / sqrt(n\_eff)}).
+\cite{Geyer:2011} provides a thorough introduction to effective sample
+size and MCSE estimation.  \proglang{Stan} uses the more conservative
+estimates based on both within-chain and cross-chain convergence; see
+\citep{GelmanEtAl:2013} and \citep{Stan:2013} for motivation and
+definitions.
+
+Because estimation accuracy is governed by the square root of the
+number of effective samples, effective samples per second (or seconds
+per effective sample) is the most relevant statistic for comparing the
+efficiency of sampler implementations.  Compared to \proglang{BUGS}
+and \proglang{JAGS}, \proglang{Stan} is often relatively slow per
+iteration but relatively fast per effective sample.
+
+In this example, the estimated number of effective samples per
+parameter (\code{n\_eff}) is 1399, which far more than we typically
+need for inference.  The posterior mean here is estimated to be 0.254
+with an MCSE of 0.00325.  Because the model is conjugate, the exact
+posterior is known to be $p(\theta|y) = \mbox{\sf Beta}(3,9)$.  Thus
+the posterior mean of $\theta$ is $3/(3+9) = 0.25$ and the posterior
+mode of $\theta$ is $(3-1)/(3 + 9 - 2) = 0.2$.
+
+\subsection{Posterior Mode Estimates}
+
+\subsubsection{Posterior modes with optimization}
+
+The posterior mode of a model can be found by using one of
+\proglang{Stan}'s built-in optimizers.  The following command invokes
+optimization for the Bernoulli model using all default configuration
+parameters.
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+> ./bernoulli optimize data file=bernoulli.data.R 
+\end{CodeInput}
+\begin{CodeOutput}
+ method = optimize
+   optimize
+     algorithm = bfgs (Default)
+       bfgs
+         init_alpha = 0.001 (Default)
+         tol_obj = 1e-08 (Default)
+         tol_grad = 1e-08 (Default)
+         tol_param = 1e-08 (Default)
+     iter = 2000 (Default)
+     save_iterations = 0 (Default)
+ id = 0 (Default)
+ data
+   file = bernoulli.data.R
+ init = 2 (Default)
+ random
+   seed = 4294967295 (Default)
+ output
+   file = output.csv (Default)
+   diagnostic_file =  (Default)
+   refresh = 100 (Default)
+
+initial log joint probability = -12.4873
+    Iter      log prob        ||dx||      ||grad||       alpha  # evals  Notes 
+       7      -5.00402   8.61455e-07   1.25715e-10           1       10   
+Optimization terminated normally: 
+  Convergence detected: change in objective function was below
+  tolerance
+\end{CodeOutput}
+\end{CodeChunk}
+%
+The final lines of the output indicate normal termination after seven
+iterations by convergence of the objective function (here the log
+probability) to the default tolerance of \code{1e-08}. The final log
+probability (\code{log prob}), length of the difference between the
+current iteration's value of the parameter vector and the previous
+value (\code{||dx||}), and the length of the gradient vector
+(\code{||grad||}).
+
+ The optimizer
+terminates when any of the log probability, gradient, or parameter
+values are within their specified tolerance.  The default optimizer is
+the Broyden-Fletcher-Goldfarb-Shanno (BFGS), a quasi-Newton method
+which employs exactly computed gradients and an efficient
+approximation to the Hessian; see \citep{NocedalWright:2006} for a
+textbook exposition of the BFGS algorithm.
+
+\subsubsection{Optimizer output file}
+
+By default, optimizations results are written into \code{output.csv},
+which is a valid CSV file.
+%
+\begin{CodeChunk}
+\begin{CodeOutput}
+# stan_version_major = 2
+# stan_version_minor = 1
+# stan_version_patch = 0
+# model = bernoulli_model
+# method = optimize
+#   optimize
+#     algorithm = bfgs (Default)
+#       bfgs
+#         init_alpha = 0.001 (Default)
+#         tol_obj = 1e-08 (Default)
+#         tol_grad = 1e-08 (Default)
+#         tol_param = 1e-08 (Default)
+#     iter = 2000 (Default)
+#     save_iterations = 0 (Default)
+# id = 0 (Default)
+# data
+#   file = bernoulli.data.R
+# init = 2 (Default)
+# random
+#   seed = 777510854
+# output
+#   file = output.csv (Default)
+#   diagnostic_file =  (Default)
+#   refresh = 100 (Default)
+lp__,theta
+-5.00402,0.2000000000125715
+\end{CodeOutput}
+\end{CodeChunk}
+%
+As with the sampler output, the configuration of the optimizer is
+dumped as CSV comments (lines beginning with \code{\#}).  Then there
+is a header, listing the log probability, \code{lp\_\_}, and the
+single parameter name, \code{theta}.  The next line shows that the
+posterior mode for \code{theta} is 0.2000000000125715, matching the
+true posterior mode of 0.20 very closely.
+
+Optimization is carried out on the unconstrained parameter space, but 
+without the Jacobian adjustment to the log probability.
+
+\subsection{Diagnostic mode}
+
+\proglang{Stan} provides a diagnostic mode that evaluates the log
+probability and gradient calculations at the initial parameter values
+(either user supplied or generated randomly based on the specified
+or default seed).
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+> ./bernoulli diagnose data file=bernoulli.data.R 
+\end{CodeInput}
+\begin{CodeOutput}
+ method = diagnose
+   diagnose
+     test = gradient (Default)
+       gradient
+         epsilon = 9.9999999999999995e-07 (Default)
+         error = 9.9999999999999995e-07 (Default)
+ id = 0 (Default)
+ data
+   file = bernoulli.data.R
+ init = 2 (Default)
+ random
+   seed = 4294967295 (Default)
+ output
+   file = output.csv (Default)
+   diagnostic_file =  (Default)
+   refresh = 100 (Default)
+
+TEST GRADIENT MODE
+ Log probability=-6.74818
+ param idx           value           model     finite diff           error
+         0         -1.1103       0.0262302       0.0262302    -3.81445e-10
+\end{CodeOutput}
+\end{CodeChunk}
+%
+Here, a random initialization is used and the initial log probability
+is \code{-6.74818} and the single parameter \code{theta}, here
+represented by index 0, has a value of \code{-1.1103} on the
+unconstrained scale.  The derivative supplied by the model and by a
+finite differences calculation are the same to within
+\code{-3.81445e-10}.  Non-finite log probability values or derivatives
+indicate a problem with the model in terms of constraints on parameter
+values or function inputs being violated, boundary conditions in
+functions, and sometimes overflow or underflow issues with
+floating-point calculations.  Errors between the model's gradient
+calculation and finite differences can indicate a bug in
+\proglang{Stan}'s algorithmic differentiation for a function in the model.
 
@@
 language and how compiled models are executed.  These details are the
-same whether a \proglang{Stan} model is being used by a built-in or
-externally supplied sampler or optimizer.
+same whether a \proglang{Stan} model is being used by one of the
+built-in samplers or optimizers or being used externally by a
+user-defined sampler or optimizer.
 
@@
 \end{Code}
-  \caption{Hierarchical binomial model with posterior inferences,
+  \caption{\it Hierarchical binomial model with posterior inferences,
     coded in \proglang{Stan}.}\label{hier-binom.fig}
@@
 A \proglang{Stan} program starts with an (optional) data block, which
-declares the data required to fit the model.  In the model in
-Figure~\ref{hier-binom.fig}, the data block declares an integer
-variable \code{J} for the number of groups in the hierarchical model.
-The arrays \code{y} and \code{n} have size \code{J}, with \code{y[j]}
-being the number of positive outcomes in \code{n[j]} trials.
+declares the data required to fit the model.  This is a very different
+approach to modeling and declarations than in \proglang{BUGS} and
+\proglang{JAGS}, which determine which variables are data and which
+are parameters at run time based on the shape of the data input to
+them.  These declarations make it possible to compile \proglang{Stan}
+to much more efficient code.%
+%
+\footnote{The speedup is because coding data variables as
+  \code{double} types in \proglang{C++} is much faster than promoting
+  all values to algorithmic differentiation class variables.}
+%
+Missing data models may still be coded in \proglang{Stan}, but the
+missing values must be declared as parameters; see \citep{Stan:2013}
+for examples of missing data, censored data, and truncated data
+models.
+
+In the model in Figure~\ref{hier-binom.fig}, the data block declares
+an integer variable \code{J} for the number of groups in the
+hierarchical model.  The arrays \code{y} and \code{n} have size
+\code{J}, with \code{y[j]} being the number of positive outcomes in
+\code{n[j]} trials.
 
@@
 \proglang{C++} object representing the model is constructed.  After
-the data is read in, the constraints are validated.  Failure to
-validate data will terminate the program and print an informative
-error message.
+the data is read in, the constraints are validated.  If the data does
+not satisfy the declared constraints, the model will throw an
+exception with an informative error message, which is displayed to the
+user in the command-line, R, and Python interfaces.
 
@@
 intended to have unconstrained support (i.e., no points of zero
-probabilty), which greatly simplifies the task of
-writing samplers or optimizers.  To achieve unbounded support,
-variables declared with constrained support are transformed 
-to an unconstrained space.  For
+probabilty), which greatly simplifies the task of writing samplers or
+optimizers.  To achieve unbounded support, variables declared with
+constrained support are transformed to an unconstrained space.  For
 instance, variables declared on $[0,1]$ are log-odds transformed and
@@
 transformed.  More complex transforms are required for simplexes (a
-reverse stick-breaking transform) and covariance matrices (Cholesky
-factorization).  The dimensionality of the resulting probabilty
-function may change as a result of the transform. For example, a $K
-\times K$ covariance matrix requires only ${K \choose 2} + K$
-unconstrained parameters, and a $K$-simplex requires only $K-1$
-unconstrained parameters.
+reverse stick-breaking transform) and covariance and correlation
+matrices (Cholesky factorization).  The dimensionality of the
+resulting probabilty function may change as a result of the
+transform. For example, a $K \times K$ covariance matrix requires only
+${K \choose 2} + K$ unconstrained parameters, and a $K$-simplex
+requires only $K-1$ unconstrained parameters.
 
@@
 %
-\footnote{For optimization, the Jacobian adjustment may be optionally
-  suppressed to guarantee the optimizer finds the maximum of the log
-  probability function on the constrained parameters.}
+\footnote{For optimization, the Jacobian adjustment is suppressed to
+  guarantee the optimizer finds the maximum of the log probability
+  function on the constrained parameters.  The calculation of the
+  Jacobian is controlled by a template parameter in the \proglang{C++}
+  code generated for a model.}
 %
@@
   v <- log(u);
-  lp__ <- lp__ + u;   // log Jacobian adjustment
+  increment_log_prob(u);      // log absolute Jacobian adjustment
 }
@@
 
-The model block defines the log probability function.  The example in
+The model block defines the log probability function on the
+constrained parameter space.  The example in
 Figure~\ref{hier-binom.fig} has a simple model containing four
@@
 $\kappa^{-5/2}$.  Note that the hierarchical prior on \code{theta} is
-vectorized: each elements of \code{theta} is drawn independently from
+vectorized: each element of \code{theta} is drawn independently from
 a beta distribution with prior success count \code{alpha} and prior
@@
 used to calculate predictive inferences as well as to carry out
-forward simulation for predictive posterior checks.
+forward simulation for predictive posterior checks; see
+\citep{GelmanEtAl:2013} for examples.
 
@@
 mortality rate.  Applying a hierarchical model then considering
-posterior inference appropriately adjusts for mulitple comparisons and
-allows; see \citep{GelmanEtAl:2012, Efron:2010} for discussion.
-
-As a second illustration, the generated quantities block calculate the
-(posterior) probability that a given institution is above-average in
-terms of mortality rate.  This is done for each institution \code{j}
-with the usual plug-in estimate of \code{theta[j] > mean(theta)},
-which returns a binary (0 or 1) value.  The posterior mean of
-\code{above\_avg[j]} calculates the posterior probability
-$\mbox{Pr}[\theta_j > \bar{\theta}|y,n]$ according to the model.
+posterior inference appropriately adjusts for mulitple comparisons;
+see \citep{GelmanEtAl:2012, Efron:2010} for discussion.
+
+As a second illustration, the generated quantities block n
+Figure~\ref{hier-binom.fig} calculates the (posterior) probability
+that a given institution is above-average in terms of mortality rate.
+This is done for each institution \code{j} with the usual plug-in
+estimate of \code{theta[j] > mean(theta)}, which returns a binary (0
+or 1) value.  The posterior mean of \code{above\_avg[j]} calculates
+the posterior probability $\mbox{Pr}[\theta_j > \bar{\theta}|y,n]$
+according to the model.
 
@@
 \proglang{Stan}'s samplers and optimizers all start from either random
-or user-supplied values for each parameter.  User
-supplied initial values are validated and transformed to
-the underlying unconstrained space; if a parameter value does not satisfy its
-declared constraints, the program exits and an informative error
-message is printed.  If random initialization is specified, the
-built-in pseudorandom number generator is called once per
-unconstrained variable dimension.  The default initialization is to
-randomly generate values uniformly on $[-2,2]$, which supplies fairly
-diffuse starting points when transformed back to the constrained
-scale.
+or user-supplied values for each parameter.  User supplied initial
+values are validated and transformed to the underlying unconstrained
+space; if a parameter value does not satisfy its declared constraints,
+the program exits and an informative error message is printed.  If
+random initialization is specified, the built-in pseudorandom number
+generator is called once per unconstrained variable dimension.  The
+default initialization is to randomly generate values uniformly on
+$[-2,2]$ by default or another interval by specification.  This
+supplies fairly diffuse starting points when transformed back to the
+constrained scale, and thus help with convergence diagnostics as
+discussed in \citep{GelmanEtAl:2013}.  Models with more data or more
+elaborate structure require narrower intervals for initialization to
+ensure the sampler is able to quickly converge to a stationary
+distribution in the high mass region of the posterior.
 
@@
 \end{center}
-\caption{Each \proglang{Stan} program block admits certain actions that are evaluated
+\caption{\it Each \proglang{Stan} program block admits certain actions that are evaluated
   at specific times during sampling.  For example, the parameter initialization in
@@
 log probability and gradient calculation for one or more leapfrog
-steps per iteration.
-
+steps per iteration; see \citep{HoffmanGelman:2011}.
 %
@@
 \end{center}
-\caption{Variables of each categorization must be
-  declared in specific blocks.  Data
-  may also be expressed using numeric
+\caption{\it Variables of each categorization must be declared in
+  specific blocks.  Data may also be expressed using numeric
   literals.}\label{variable-kinds.fig}
@@
 Calculations in the generated quantities block do not impact the
-estimates of the parameters.  In this way, it can be used like the cut
-feature of \proglang{BUGS}.  Nevertheless, we recommend full Bayesian
-inference in general, and do not use it for cut-like behavior unless
-it does not effect inference (e.g., for generated predictions for
-unseen data in many models).
-
-The list is completed with two types of local variables, loop indices
-and traditional local variables.  Unlike \proglang{BUGS},
-\proglang{Stan} allows local variables to be declared and assigned.
-For example, it is possible to compute the sum of the squares of
-entries in an array or vector \code{y} as follows.%
+estimates of the parameters.  In this way, generated quantities can be
+used for predictions with a behavior not unlike that of cut in
+\proglang{BUGS} \citep{LunnEtAl:2012}.  Nevertheless, we recommend
+full Bayesian inference in general, and only use the generated
+quantities block for efficiency and clarity of code.  For example, in
+many models, generating predictions for unseen data or replicating
+data for posterior predictive checks does not affect the estimation of
+the parameters of interest.
+
+The list in Figure~\ref{variable-kinds.fig} is completed with two
+types of local variables, loop indices and traditional local
+variables.  Unlike \proglang{BUGS}, \proglang{Stan} allows local
+variables to be declared and assigned.  For example, it is possible to
+compute the sum of the squares of entries in an array or vector
+\code{y} as follows.%
 %
@@
 variable has unbounded support.  For instance, the uniform
-distribution over simplexes and correlation matrices is proper, but
-the uniform distribution over ordered vectors or covariance matrices
-is not.
+distribution over real values with upper and lower bounds, simplexes
+and correlation matrices is proper, but the uniform distribution over
+unconstrained or one-side constrained reals, ordered vectors or
+covariance matrices is not proper.
+
+\proglang{Stan} does not require proper priors, but if the posterior
+is improper, \proglang{Stan} will halt with an error message.%
+%
+\footnote{Improper posteriors are diagnosed automatically when
+  parameters overflow to infinity during simulation.}
 
@@
 resulting sampler devolves into a random walk with high
-autocorrelations.
-
-\cite{BetancourtGirolami:2013} discuss for more
-analysis of centered vs.\ non-centered parameterizations of
-hierarchical models.
+autocorrelations.  \cite{BetancourtGirolami:2013} provide a detailed
+analysis of the benefits of centered vs.\ non-centered
+parameterizations of hierarchical models in Euclidean and Riemannian
+HMC, with the conclusion that non-centered are best when data are
+sparse and centered when the data strongly identifies coefficients.
 
 The following reparameterization employs a non-centered
-parameterization (see \citep{BetancourtGirolami:2013} for an analysis
-of the benefits of non-centered parameterizations).
+parameterization for comparison.
 %
@@
 hierarchical variables that would otherwise limit the effectiveness
-of the samplers.
+of the samplers.  We recommend starting with the more natural centered
+parameterization and moving to the non-centered parameterization if
+sampling of the coefficients does not mix well.%
+%
+\footnote{As the saying goes in computer science, it is easier to
+  optimize a correct program than debug an optimized program.}
 
@@
 
-Expressions in \proglang{Stan}, including variables are statically
-typed.  This means their type is declared at compile time as part of
-the model, and does not change throughout the execution of the
+All expressions in \proglang{Stan} are statically typed, including
+variables.  This means their type is declared at compile time as part
+of the model, and does not change throughout the execution of the
 program.  This is the same behavior as is found in compiled
@@
 automatically adapts the number of leapfrog steps, eliminating the
-need for user-specified tuning parameters.  Both algorithms take advantage
-of curvature information, namely the gradient, to generate coherent motion
-through the posterior that dramatically reduces the autocorrelation of
-the resulting transitions.
+need for user-specified tuning parameters.  Both algorithms take
+advantage of curvature information, namely the gradient, to generate
+coherent motion through the posterior that dramatically reduces the
+autocorrelation of the resulting transitions.
 
@@
 \end{center}
-\caption{Each of \proglang{Stan}'s unary and binary operators follow
+\caption{\it Each of \proglang{Stan}'s unary and binary operators follow
   strict precedences, associativities, placement within an expression.  
@@
 \end{center}
-\caption{\proglang{Stan} implements a variety of useful constants.}\label{constants.fig}
+\caption{\it \proglang{Stan} implements a variety of useful constants.}\label{constants.fig}
 \end{figure}
@@
 \end{center}
-\caption{\proglang{Stan} implements both circular and 
+\caption{\it \proglang{Stan} implements both circular and 
   hyperbolic trigonometric functions, as well as their inverses.}
@@
 \end{center}
-\caption{\proglang{Stan} implements many special and transcendental functions.}\label{special-functions.fig}
+\caption{\it \proglang{Stan} implements many special and transcendental functions.}\label{special-functions.fig}
 \end{figure}
@@
 \end{center}
-\caption{Special functions (continued).}\label{special-functions-cont.fig}
+\caption{\it Special functions (continued).}\label{special-functions-cont.fig}
 \end{figure}
@@
 \end{center}
-\caption{A large suite of matrix functions admits efficient 
+\caption{\it A large suite of matrix functions admits efficient 
   multivariate model implementation in \proglang{Stan}.}
@@
 \end{center}
-\caption{Most common probability distributions have
+\caption{\it Most common probability distributions have
   explicitly implemented in \proglang{Stan}.}\label{prob-functions.fig}
@@
 \end{center}
-\caption{Probability functions (continued)}\label{prob-functions-cont.fig}
+\caption{\it Probability functions (continued)}\label{prob-functions-cont.fig}
 \end{figure}
@@
 \end{center}
-\caption{Probability functions (continued 2)}\label{prob-functions-cont-2.fig}
+\caption{\it Probability functions (continued 2)}\label{prob-functions-cont-2.fig}
 \end{figure}

### 2013-12-26 src/docs/papers/stan-language/article.tex
finished revisions of Stan paper

@@
   \\
-  E-mail: \email{carp@stat.columbia.edu}
+  E-mail: \email{carp@alias-i.com}
   \\ 
@@
 \begin{document}
-
-\section{TO-DO}
-
-\begin{enumerate}
-\item Convergence diagnostics?
-\item more model examples?
-\item description from C++ perspective?
-\item what else do we want to make sure users don't miss?
-\item optimization example?
-\item future plans section?
-\end{enumerate}
 
@@
 The primitive types of \proglang{Stan} are \code{real} and \code{int},
-which are used to represent continuous and integral values.  These
-values are represented directly in \proglang{C++}.  Integer
-expressions can be used anywhere a real value is required, but not
-{\it vice-versa}.
+which are used to represent continuous and integer values.  These
+values are represented directly in \proglang{C++} as types
+\code{double} and \code{int}.  Integer expressions can be used
+anywhere a real value is required, but not {\it vice-versa}.
 
@@
 %
-\footnote{ Arrays are stored in row-major order and matrices in
+\footnote{Arrays are stored in row-major order and matrices in
   column-major order.}
+
+Unlike integers, which may be used where real values are required,
+arrays of integers may not be used where real arrays are required.%
+%
+\footnote{In the language of type theory, \proglang{Stan} arrays are
+  not covariant.  This follows the behavior of both arrays and
+  standard library containers in \proglang{C++}.}
 
@@
 \code{real} result by promoting the \code{int} argument to a
-\code{real}.  \proglang{Stan} also attempts to treat arrays as
-covariant, meaning that arrays of of \code{int} values can often be
-used where arrays of \code{real} values are specified.
+\code{real}.
 
@@
 \begin{Code}
-lp__ <- lp__ + normal_log(beta,0,1);
+increment_log_prob(normal_log(beta,0,1));
 \end{Code}
+%
+The sampling statement distribution \code{normal} is shorthand for the
+function \code{normal\_log}, and both the left-hand side variate and
+right-hand side parameters become arguments.
 
@@
 \begin{Code}
-lp__ <- lp__ - 0.5 * beta * beta;
+increment_log_prob(-0.5 * beta * beta);
 \end{Code}
@@
 (outside of comments) is irrelevant.  This is in contrast to
-\proglang{BUGS} and \proglang{JAGS} in which carriage returns are
+\proglang{BUGS} and \proglang{R}, in which carriage returns are
 special and may indicate the end of a statement.
@@
 indexing uses the usual square brackets (\code{[ ]}) operator, and
-begins indexing from 1.  If \code{Sigma} is a matrix, then
-\code{Sigma[n]} is the \code{n}-th row of \code{Sigma}.
+begins indexing from 1.  For example, if \code{Sigma} is a matrix,
+then \code{Sigma[m,n]} is the entry at row \code{m} and column
+\code{n}.  \proglang{Stan} also allows slices of data structures to be
+returned, so that \code{Sigma[m]} is row \code{m} of the matrix
+\code{Sigma}.
 
@@
 \code{mu} and \code{sigma} contain more than one element. A typical
-use case would be for linear regression, such as \code{normal\_log(y,X
+use case would be for linear regression, such as \code{y ~ normal(X
   * beta,sigma)}, where \code{y} is a vector of observed data,
@@
 executable for calculating gradients and higher-order derivatives.
-In the example above, \code{normal\_log(y,X *
+In the example above, \code{y ~ normal(X *
   beta,sigma)}, the logarithm of \code{sigma} need only be computed
@@
 The MCMC samplers provided include Euclidean Manifold Hamiltonian
-Monte Carlo (EM-HMC) \citep{DuaneEtAl:1987, Neal:1994, Neal:2011} and
-the No-U-Turn sampler (NUTS) \citep{HoffmanGelman:2011}.  Both
-versions of HMC allow estimation or specification of unit, diagonal,
-or full mass matrices.  NUTS, the default sampler for \proglang{Stan},
-automatically adapts the number of leapfrog steps, eliminating the
-need for user-specified tuning parameters.  Both algorithms take
-advantage of curvature information, namely the gradient, to generate
-coherent motion through the posterior that dramatically reduces the
-autocorrelation of the resulting transitions.
-
-\proglang{Stan} will soon have an implementation of Riemannian Manifold
-Hamiltonian Monte Carlo (RM-HMC) \citep{GirolamiCalderhead:2011},
-using the SoftAbs metric to robustly incorporate posterior curvature
-into the Hamiltonian simulation \cite{Betancourt:2012}, although
-\proglang{Stan}'s MCMC framework makes it
-straightforward to incorporate other metrics.  This
-implementation will also generalize NUTS to Riemannian manifolds
+Monte Carlo (EHMC, or just HMC) \citep{DuaneEtAl:1987, Neal:1994,
+  Neal:2011} and the No-U-Turn sampler (NUTS)
+\citep{HoffmanGelman:2011}.  Both versions of HMC allow estimation or
+specification of unit, diagonal, or full mass matrices.  NUTS, the
+default sampler for \proglang{Stan}, automatically adapts the number
+of leapfrog steps, eliminating the need for user-specified tuning
+parameters.  Both algorithms take advantage of gradient information in
+the log probability function to generate coherent motion through the
+posterior that dramatically reduces the autocorrelation of the
+resulting transitions.
+
+\proglang{Stan} will soon have an implementation of Riemannian
+Manifold Hamiltonian Monte Carlo (RHMC)
+\citep{GirolamiCalderhead:2011}, using the SoftAbs metric to robustly
+incorporate posterior curvature into the Hamiltonian simulation
+\cite{Betancourt:2012}.  \proglang{Stan}'s MCMC framework also makes
+it straightforward to incorporate other metrics.  This implementation
+will also generalize NUTS to Riemannian manifolds
 \citep{Betancourt:2013}.  Both support adapting step sizes during
-warmup.  RM-HMC uses first, second, and third order derivatives 
-to adapt to the local curvature of the posterior.
+warmup.  RHMC with SoftAbs uses first, second, and third order
+derivatives to adapt to the local curvature of the posterior.
 
@@
 
-\proglang{Stan} supports random-walk Metropolis-Hastings samplers with
-multivariate normal proposals, the primary purpose of which is to
-provide baselines against which to measure more efficient approaches
-to sampling.  The covariance matrix used can be adapted during warmup,
-assuming either equal diagonal covariance, diagonal covariance, or a
-full covariance matrix.  Metropolis-Hastings does not require any
-derivatives of the log probability function, but the lack of gradient
-information induces random walk behavior.
+Random-walk Metropolis-Hastings samplers with multivariate normal
+proposals have been prototyped, but not yet integrated into the
+released versions of \proglang{Stan}.  The primary purpose of
+implementing Metropolis-Hastings is to provide baselines against which
+to measure more efficient approaches to sampling.  The covariance
+matrix used can be adapted during warmup, assuming either equal
+diagonal covariance, diagonal covariance, or a full covariance matrix.
+Metropolis-Hastings does not require any derivatives of the log
+probability function, but the lack of gradient information induces
+random walk behavior, leading to slow mixing.
 
@@
 chains to be run as the number of dimensions in order to have full
-rank search of the posterior.
+rank search of the posterior.  They also require slightly different
+management of output due to a single sampler instance producing
+multiple correlated chains of draws.
 
@@
 quasi-Newton optimizer that evaluates gradients directly, then uses
-the gradients to update an approximation to the Hessian; see
-\citep{NocedalWright:2006} for details.  Plans are in the works to
-also include the more involved, but more scalable limited-memory BFGS
-(L-BFGS) scheme.
+the gradients to update an approximation to the Hessian.  Plans are in
+the works to also include the more involved, but more scalable
+limited-memory BFGS (L-BFGS) scheme.  \citep{NocedalWright:2006} cover
+both BFGS and L-BFGS samplers.
 
@@
 Additionally, \proglang{Stan} implements a crude version of Nesterov's
-accelerated gradient optimizer \cite{Nesterov:1983} which combines
-gradient updates with a momentum-like update to speed up convergence.
+accelerated gradient optimizer \cite{Nesterov:1983}, which combines
+gradient updates with a momentum-like update to hasten convergence.
 
@@
 
-This includes extensive unit tests for low-level \proglang{C++} code.
+\proglang{Stan} includes extensive unit tests for low-level \proglang{C++} code.
 Unit tests are implemented using the \pkg{googletest} framework

### 2013-12-29 src/docs/papers/stan-language/article.tex
submitted version with pdf of paper and tarball of source

@@
 
-\renewcommand{\textfraction}{0.07}
-\renewcommand{\topfraction}{0.9}	% max fraction of floats at top
+\renewcommand{\textfraction}{0.02}
+\renewcommand{\topfraction}{0.95}	% max fraction of floats at top
 
@@
         %
+        \\[9pt]
+        %
+        {\bf\large Allen Riddell}
+        \\ Dartmouth College
     \And
@@
         \\ Columbia University
+        %
+        \\[9pt]
 }
@@
 
-  \proglang{Stan} is also a platform for computing log-densities and
+  \proglang{Stan} is also a platform for computing log densities and
   their gradients and Hessians, which can be used in alternative
@@
 These models turned out to be a challenge for
-existing general-purpose inference software.  A direct encoding in
+existing general purpose inference software.  A direct encoding in
 \proglang{BUGS} \citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
@@
 algorithmic differentiation, which, given a templated \proglang{C++}
-function for the log posterior, automatically computes a proper
-analytic gradient up to machine precision accuracy in only a few
-multiples of the cost to evaluate the log probability function itself.
+function for the log posterior, automatically computes its analytic
+gradient up to machine precision accuracy in only a few multiples of
+the time to evaluate the log probability function itself.%
+%
+\footnote{Hessian matrices (i.e., all second order derivatives) are
+  more expensive to calculate; each row corresponds to a parameter and
+  is filled by the gradient of the derivative of the log probability
+  function with respect to the parameter.}
+%
 We explored existing algorithmic differentiation packages with open
@@
 Sacado module of the \pkg{Trilinos} toolkit, and the \pkg{CppAD}
-package of the COIN-OR toolkit \citep{BellBurke:2008}.  Both
-packages, however, supported few special functions (e.g., probability
-functions, log gamma, inverse logit) or linear algebra operations
-(e.g., Cholesky decompositions, matrix division), and neither are
-easily or modularly extensible.
+package of the COIN-OR toolkit \citep{BellBurke:2008}.  Both packages,
+however, supported few special functions (e.g., probability functions,
+log gamma, inverse logit) or linear algebra operations (e.g., Cholesky
+decompositions, matrix division), and neither are easily or modularly
+extensible.
 
@@
 At this point, we could fit models coded directly in \proglang{C++} on
-top of the pre-release versions of the \proglang{Stan} API
-(application programming interface).  Seeing how well this all worked,
-we set our sights on the generality and ease of use of \proglang{BUGS}
-and designed a modeling language in which statisticians could write
-their models in familiar notation that could be transformed to
-efficient \proglang{C++} code before being compiled into an efficient
-executable program.  Although we started with the target of specifying
-directed graphical models, our first implementation turned out to be a
-more general imperative form of specifying log probability functions;
+top of the pre-release versions of the \proglang{Stan} application
+programming interface (API).  Seeing how well this all worked, we set
+our sights on the generality and ease of use of \proglang{BUGS} and
+designed a modeling language in which statisticians could write their
+models in familiar notation that could be transformed to efficient
+\proglang{C++} code before being compiled into an efficient executable
+program.  Although we started with the target of specifying directed
+graphical models, our first implementation turned out to be a more
+general imperative form of specifying log probability functions;
 \proglang{BUGS} models could be translated line for line, but the
 \proglang{Stan} language also supports much richer imperative
-constructs like conditionals, while loops, and local variables.
-This paper is primarily about the \proglang{Stan} language and how
-it can be used to specify statistical models.
+constructs like conditionals, while loops, and local variables.  This
+paper is primarily about the \proglang{Stan} language and how it can
+be used to specify statistical models.
 
@@
 back on itself.  Once the trajectory has terminated, a new sample is
-drawn from the tree.
+drawn from the tree.  In its original version, NUTS also estimates a
+step size during warmup based on a target acceptance probability.
 
@@
 for multiple variates with the same mean and scale, which sped things
-up a bit. At the same time, we introduced some simple template
-metaprograms to remove the calculation of constant terms in the log
-probability.  Performance was lacking until we finally figured out how
+up a bit.  Performance was lacking until we finally figured out how
 to both eliminate redundant calculations and partially evaluate the
@@
 performance of the numerical simulation in HMC scales with the
-variation of the parameter scales; fortunately HMC also provides a
-choice of metric, in our case a Euclidean metric with a constant mass
-matrix, to compensate for variations in the log posterior.  In
-\proglang{Stan} 1.0 we introduced an adaptive diagonal metric into our
-HMC implementations that allowed the parameter scales to be normalized
-automatically, with the option of an adaptive dense metric and a full
-automated standardization of the posterior in \proglang{Stan} 2.0.
-
-Because the dense metric performs only a global standardization, the
-performance of \proglang{Stan} in models where the relative parameter
-scales varies by location, such as hierarchical and latent models, can
-suffer.  Riemannian manifold Hamiltonian Monte Carlo (RHMC) introduces
-a location-dependent metric that can overcome these final hurdles
-\citep{GirolamiCalderhead:2011}; Stan has a prototype implementation
-of RHMC based on the SoftAbs metric of \citep{Betancourt:2012}, using
-a generalization of NUTS to Riemannian manifolds \citep{Betancourt:2013}.
+variation of the parameter scales.  In \proglang{Stan} 1.0 we
+introduced an adaptive diagonal metric (mass matrix) into our HMC
+implementations that allowed the parameter scales to be standardized
+automatically; Stan 2.0 added an option to estimate a dense metric
+(mass matrix) and a full covariance standardization of the posterior.
+
+Because the diagonal and dense metrics perform only a global
+standardization, the performance of \proglang{Stan} in models where
+the relative parameter scales varies by location, such as hierarchical
+and latent models, can suffer (hierarchical modeling is covered in
+Sections~\ref{hierarchical-modeling.section} and reparameterizations to speed it
+up in \ref{centering.section}).  Riemannian manifold Hamiltonian Monte
+Carlo (RHMC) introduces a location-dependent metric that can overcome
+these final hurdles \citep{GirolamiCalderhead:2011}; Stan has a
+prototype implementation of RHMC based on the SoftAbs metric of
+\citep{Betancourt:2012}, using a generalization of NUTS to Riemannian
+manifolds \citep{Betancourt:2013}.
 
@@
 %
-For Windows, the \code{./} before the command should be removed.  This
-invocation specifies the executable program, the data file, the file
-to which the samples are written, the identity of the chain, and a
-seed for the random number generator.
+For Windows, the \code{./} before the command should be removed.
 
@@
 The sampler and its configuration are described at greater length in
-the manual.
+the manual \citep{Stan:2013}.
 
@@
 configuration parameters, exactly the same output file can be
-reproduced.  The pseudo-random numbers generated by the sampler are
+reproduced.  The pseudorandom numbers generated by the sampler are
 fully determined by the seed (here randomly generated based on the
@@
 \code{0}).  The identifier is used to advance the underlying
-pseudo-random number generator a sufficient number of values that
+pseudorandom number generator a sufficient number of values that
 using multiple chains with the same seed and different identifiers
-will draw from different subsequences of the pseudo-random number
+will draw from different subsequences of the pseudorandom number
 stream determined by the seed.
@@
   possible samples in the proposed tree; NUTS itself uses a slice
-  sampling algorithm for rejection \citep{Neal:2003}.}
+  sampling algorithm for rejection \citep{Neal:2003, HoffmanGelman:2011}.}
 %
@@
 regularizing the sample covariance of the latter half of the warmup
-samplers; see \citep{Stan:2013} for full details.  A dense mass matrix
+samples; see \citep{Stan:2013} for full details.  A dense mass matrix
 may also be estimated, or the mass matrix may be set to the unit matrix.
@@
 Section~\ref{rng.section} for more information.
-
-\proglang{Stan} supplies a command-line program \code{bin/print} to
-summarize the output of one or more MCMC chains.  Given a directory
-containing output from sampling, it is invoked as
-%
-\begin{CodeChunk}
-\begin{CodeInput}
-> ls output*.csv
-\end{CodeInput}
-\begin{CodeOutput}
-output.csv	output1.csv	output2.csv	output3.csv
-\end{CodeOutput}
-\begin{CodeInput}
-> $stan/bin/print output*.csv
-\end{CodeInput}
-
-\end{CodeChunk}
-%
-The output is shown in Figure~\ref{print-output.fig}.%
-%
-\footnote{Aligning columns when printing rows of varying scales
-  presents a challenge.  For each column, the program calculates the
-  the maximum number of digits required to print an entry in that
-  column with the specified precision. For example, a precision of 2
-  for the number -0.000012 requires nine characters (\code{-0.000012})
-  to print without scientific notation versus seven digits with
-  (\code{-1.2e-5}).  If the discrepancy is above a fixed threshold,
-  scientific notation is used.  Compare the results in the \code{mean}
-  column versus the \code{sd} column.}
-%
 
@@
 \end{figure}
+
+\proglang{Stan} supplies a command-line program \code{bin/print} to
+summarize the output of one or more MCMC chains.  Given a directory
+containing output from sampling,
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+> ls output*.csv
+\end{CodeInput}
+\begin{CodeOutput}
+output.csv	output1.csv	output2.csv	output3.csv
+\end{CodeOutput}
+\end{CodeChunk}
+%
+posterior summaries are printed using
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+> $stan/bin/print output*.csv
+\end{CodeInput}
+\end{CodeChunk}
+%
+The output is shown in Figure~\ref{print-output.fig}.%
+%
+\footnote{Aligning columns when printing rows of varying scales
+  presents a challenge.  For each column, the program calculates the
+  the maximum number of digits required to print an entry in that
+  column with the specified precision. For example, a precision of 2
+  for the number -0.000012 requires nine characters (\code{-0.000012})
+  to print without scientific notation versus seven digits with
+  (\code{-1.2e-5}).  If the discrepancy is above a fixed threshold,
+  scientific notation is used.  Compare the results in the \code{mean}
+  column versus the \code{sd} column.}
 %
@@
 estimates such as posterior means are not as accurate as they would be
-with truly independent samples.  To number of effective samples is an
+with truly independent samples.  The number of effective samples is an
 estimate of the number of independent samples that would lead to the
@@
 
-\subsection{Posterior Mode Estimates}
+\subsection{Posterior mode estimates}
 
@@
 
- The optimizer
-terminates when any of the log probability, gradient, or parameter
-values are within their specified tolerance.  The default optimizer is
-the Broyden-Fletcher-Goldfarb-Shanno (BFGS), a quasi-Newton method
-which employs exactly computed gradients and an efficient
-approximation to the Hessian; see \citep{NocedalWright:2006} for a
-textbook exposition of the BFGS algorithm.
+The optimizer terminates when any of the log probability, gradient, or
+parameter values are within their specified tolerance.  The default
+optimizer uses the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm,
+a quasi-Newton method which employs exactly computed gradients and an
+efficient approximation to the Hessian; see \citep{NocedalWright:2006}
+for a textbook exposition of the BFGS algorithm.
 
@@
 Optimization is carried out on the unconstrained parameter space, but
-without the Jacobian adjustment to the log probability.
+without the Jacobian adjustment to the log probability.  This ensures
+modes are defined with respect to the constrained parameter space as
+declared in the parameters block and used in the model specification.
+The need to suppress the Jacobian to match the scaling of the declared
+parameters highlights the sensitivity of posterior modes to parameter
+transforms.
 
@@
 
-\subsection{Example: hierarchical model, with inference}
+\subsection{Example: hierarchical model, with inference}\label{hierarchical-modeling.section}
 
@@
 The \proglang{BUGS} surgical example explored the ranking of
-institutions in terms of surgical mortality.  This is coded in the
-example in Figure~\ref{hier-binom.fig} using the generated quantities
-block.  The generated quantity variable \code{rnk[j]} will hold the
-rank of institution \code{j} from 1 to \code{J} in terms of mortality
-rate \code{theta[j]}.  The ranks are extracted using the \code{rank}
-function. The posterior summary will print average rank and deviation,
-but the values can also be extracted in order to plot posterior rank
-histograms by institution as done in \citep[Examples, Volume
-1]{LunnEtAl:2000}.
+institutions in terms of surgical mortality \citep[Examples, Volume
+1]{LunnEtAl:2000}.  This is coded in the example in
+Figure~\ref{hier-binom.fig} using the generated quantities block.  The
+generated quantity variable \code{rnk[j]} will hold the rank of
+institution \code{j} from 1 to \code{J} in terms of mortality rate
+\code{theta[j]}.  The ranks are extracted using the \code{rank}
+function. The posterior summary will print average rank and deviation.
+\citep{LunnEtAl:2000} illustrated posterior inference by plotting
+posterior rank histograms.
 
@@
 %
-\cite[p.~366]{GelmanHill:2007} provide a taxonomy of the kinds of
+\citet[p.~366]{GelmanHill:2007} provide a taxonomy of the kinds of
 variables used in Bayesian models.  Figure~\ref{variable-kinds.fig}
@@
 
-\subsection{Implicit Uniform Priors}\label{implicit-prior.section}
+\subsection{Implicit uniform priors}\label{implicit-prior.section}
 
@@
 variable has unbounded support.  For instance, the uniform
-distribution over real values with upper and lower bounds, simplexes
+distributions over real values with upper and lower bounds, simplexes
 and correlation matrices is proper, but the uniform distribution over
 unconstrained or one-side constrained reals, ordered vectors or
-covariance matrices is not proper.
+covariance matrices are not proper.
 
@@
 
-\subsection{The non-centered parameterization for hierarchical models}
+\subsection{The non-centered parameterization for hierarchical models}\label{centering.section}
 
@@
 %
-\footnote{The limited density variation is a ultimately consequence of
+\footnote{The limited density variation is ultimately a consequence of
   a constant metric.  An additional benefit of RHMC are transitions
@@
 
-\subsection{Primitive Types}
+\subsection{Primitive types}
 
@@
 
-\subsection{Vector and Matrix Types}
+\subsection{Vector and matrix types}
 
@@
 
-\subsection{Array Types}
+\subsection{Array types}
 
@@
 Higher-dimensional arrays are intrinsically arrays of arrays.  An
-entry two-dimensional array \code{y} may be accessed as \code{y[1,2]}
-or \code{y[1][2]}, with the expression \code{y[1]} denoting an array.%
+entry in a two-dimensional array \code{y} may be accessed as
+\code{y[1,2]}.  The expression \code{y[1]} by itself denotes the
+one-dimensional array whose values correspond to the first row of
+\code{y}.  Thus \code{y[1][2]} has the same value as \code{y[1,2]}.%
 %
@@
 
-\subsection{Constrained Variable Types}
+\subsection{Constrained variable types}
 
@@
 the transformed parameters and model blocks) denoting the log
-probability that will be returned by the log probability function.  A
-sampling statement is nothing more than shorthand for incrementing the
-log probability accumulator \code{lp\_\_}.  For example, if \code{beta}
-is a parameter of type \code{real}, the sampling statement
+probability that will be returned by the log probability function.
+
+\subsubsection{Sampling statements}
+
+A sampling statement is nothing more than shorthand for incrementing
+the log probability accumulator \code{lp\_\_}.  For example, if
+\code{beta} is a parameter of type \code{real}, the sampling statement
 %
@@
 %
-has the exact same effect as the assignment statement
+has the exact same effect (up to dropping constant terms) as the
+special log probability increment statement
 %
@@
 \end{Code}
-%
-The sampling statement distribution \code{normal} is shorthand for the
-function \code{normal\_log}, and both the left-hand side variate and
-right-hand side parameters become arguments.
+
+\subsubsection{Define variables before sampling statements}
+
+The translation of sampling statements to log probability function
+evaluations explains why variables must be defined {\it before} they
+are used.  In particular, a sampling statement does \emph{not} sample
+the left-hand side variable from the right-hand side distribution.
+
+Parameters are all defined externally by the sampler; all other
+variables must be explicitly defined with an assignment statement
+before being used.
+
+\subsubsection{Direct definition of probability functions}
 
@@
 Various reductions are provided for arrays and matrices, such as sums,
-means, standard deviations and norms.  Replications are also available
+means, standard deviations, and norms.  Replications are also available
 to copy a value into every cell of a matrix.  Slices of matrices and
@@
 All probability functions are defined on the log scale to avoid
-underflow.  They are all named with the suffix \code{\_log}, e.g.
+underflow.  They are all named with the suffix \code{\_log}, e.g.,
 \code{normal\_log()}, is the log-scale normal distribution density
@@
 %
-For example, executing \code{normal\_log(y,0,sigma)} only evaluates
+For example, executing \code{y ~ normal(0,sigma)} only evaluates
 \code{log(sigma)} if \code{sigma} is a parameter, transformed
@@
 constant as defined above.
+
+Constant terms are {\it not} dropped in explicit function evaluations,
+such as \code{normal\_log(y,0,sigma)}.
 
@@
 
-The MCMC samplers provided include Euclidean Manifold Hamiltonian
+The MCMC samplers provided include Euclidean manifold Hamiltonian
 Monte Carlo (EHMC, or just HMC) \citep{DuaneEtAl:1987, Neal:1994,
   Neal:2011} and the No-U-Turn sampler (NUTS)
-\citep{HoffmanGelman:2011}.  Both versions of HMC allow estimation or
-specification of unit, diagonal, or full mass matrices.  NUTS, the
-default sampler for \proglang{Stan}, automatically adapts the number
-of leapfrog steps, eliminating the need for user-specified tuning
-parameters.  Both algorithms take advantage of gradient information in
-the log probability function to generate coherent motion through the
-posterior that dramatically reduces the autocorrelation of the
-resulting transitions.
+\citep{HoffmanGelman:2011}.  Both the basic and NUTS versions of HMC
+allow estimation or specification of unit, diagonal, or full mass
+matrices.  NUTS, the default sampler for \proglang{Stan},
+automatically adapts the number of leapfrog steps, eliminating the
+need for user-specified tuning parameters.  Both algorithms take
+advantage of gradient information in the log probability function to
+generate coherent motion through the posterior that dramatically
+reduces the autocorrelation of the resulting transitions.
 
 \proglang{Stan} will soon have an implementation of Riemannian
-Manifold Hamiltonian Monte Carlo (RHMC)
+manifold Hamiltonian Monte Carlo (RHMC)
 \citep{GirolamiCalderhead:2011}, using the SoftAbs metric to robustly
@@
 
-\subsubsection{Ensemble Samplers}
+\subsubsection{Ensemble samplers}
 
@@
 ensemble sampler \citep{GoodmanWeare:2010} and a differential
-evolution sampler \cite{TerBraak:2006}.  Ensemble samplers do not
+evolution sampler \citep{TerBraak:2006}.  Ensemble samplers do not
 require derivative information, but typically require at least as many
@@
 the works to also include the more involved, but more scalable
-limited-memory BFGS (L-BFGS) scheme.  \citep{NocedalWright:2006} cover
+limited-memory BFGS (L-BFGS) scheme.  \citet{NocedalWright:2006} cover
 both BFGS and L-BFGS samplers.
@@
 \pkg{doxygen} package \citep{Doxygen:2011}.  The \proglang{Stan}
-manual \citep{Stan:2013} is typeset using the \LaTeX package
+manual \citep{Stan:2013} is typeset using the \LaTeX\ package
 \citep{MittelbachEtAl:2004}.
@@
 analysis courses at Columbia Univesity and Harvard University, who
-served as trial subjects for both \proglang{Stan} and \cite{GelmanEtAl:2013}.
+served as trial subjects for both \proglang{Stan} and \citep{GelmanEtAl:2013}.
 

### 2014-02-12 src/docs/papers/stan-language/article.tex
updated Stan paper and bib submission style for JStatSoft

@@
   \proglang{Stan} provides full Bayesian inference for continuous-variable
-  models through Markov chain Monte Carlo (MCMC) methods such as
-  the No-U-Turn sampler (NUTS), an adaptive form of Hamiltonian Monte
-  Carlo (HMC) sampling.  Penalized maximum likelihood estimates are
-  calculated using optimization methods such as BFGS.
+  models through Markov chain Monte Carlo methods such as
+  the No-U-Turn sampler, an adaptive form of Hamiltonian Monte
+  Carlo sampling.  Penalized maximum likelihood estimates are
+  calculated using optimization methods such as the
+  Broyden–Fletcher–Goldfarb–Shanno algorithm.
 
@@
   algorithms such as variational Bayes, expectation propagation, and
-  marginal inference using approximate integration.  To this end, Stan
+  marginal inference using approximate integration.  To this end, \proglang{Stan}
   is set up so that the densities, gradients, and Hessians, along with
@@
   \proglang{R} using the \pkg{RStan} package, or through
-  \proglang{Python} using the \pkg{PyStan} package.  All three
+<  \proglang{Python} using the \pkg{PyStan} package.  All three
   interfaces support sampling or optimization-based inference and
@@
 \footnote{\proglang{Stan}'s home page, \url{http://mc-stan.org/},
-  links to the Stan manual, example models including translations of
-  the BUGS examples and the models from \citep{GelmanHill:2007},
-  getting started instructions and full documentation for Stan's
+  links to the \proglang{Stan} manual, example models including translations of
+  the \proglang{BUGS} examples and the models from \citep{GelmanHill:2007},
+  getting started instructions and full documentation for \proglang{Stan}'s
   interfaces for the command line shell (\pkg{CmdStan}), \proglang{Python}
@@
 implementations that allowed the parameter scales to be standardized
-automatically; Stan 2.0 added an option to estimate a dense metric
+automatically; \proglang{Stan} 2.0 added an option to estimate a dense metric
 (mass matrix) and a full covariance standardization of the posterior.
@@
 and latent models, can suffer (hierarchical modeling is covered in
-Sections~\ref{hierarchical-modeling.section} and reparameterizations to speed it
-up in \ref{centering.section}).  Riemannian manifold Hamiltonian Monte
+Section~\ref{hierarchical-modeling.section} and reparameterizations to speed it
+up in Section~\ref{centering.section}).  Riemannian manifold Hamiltonian Monte
 Carlo (RHMC) introduces a location-dependent metric that can overcome
-these final hurdles \citep{GirolamiCalderhead:2011}; Stan has a
+these final hurdles \citep{GirolamiCalderhead:2011}; \proglang{Stan} has a
 prototype implementation of RHMC based on the SoftAbs metric of
@@
 \code{theta}.  The vectorized likelihood statement can also be coded
-using a loop as in BUGS, although it will run more slowly than the
+using a loop as in \proglang{BUGS}, although it will run more slowly than the
 vectorized form:
@@
 The output contiues with a CSV header naming the columns of the
-output. For the default NUTS sampler in Stan 2.1.0, these are
+output. For the default NUTS sampler in \proglang{Stan} 2.1.0, these are
 %
@@
 
-Stan follows the \proglang{C++} convention of separating words in
+\proglang{Stan} follows the \proglang{C++} convention of separating words in
 variable names using underbars (\code{\_}), rather than dots
@@
 values of x near 0 than \code{log(1 - x)}.  \proglang{Stan}'s special
-functions are listed in Figures~\ref{special-functions.fig} and
-\ref{special-functions-cont.fig}.
-
-In addition to special functions, Stan includes distributions with
+functions are listed in Figure~\ref{special-functions.fig} and
+Figure~\ref{special-functions-cont.fig}.
+
+In addition to special functions, \proglang{Stan} includes distributions with
 alternative parameterizations, such as \code{bernoulli_logit}, which
@@
 
-Stan supports a growing collection of built-in univariate and
+\proglang{Stan} supports a growing collection of built-in univariate and
 multivariate probability density and mass functions.  These
@@
 
-The list of probability functions is provided in Figures
-\ref{prob-functions.fig}, \ref{prob-functions-cont.fig}, and
-\ref{prob-functions-cont-2.fig}.
+The list of probability functions is provided in
+Figure~\ref{prob-functions.fig}, 
+Figure~\ref{prob-functions-cont.fig}, and
+Figure~\ref{prob-functions-cont-2.fig}.
 
@@
 In addition to performing full Bayesian inference via posterior
-sampling, Stan also can perform optimization (i.e., computation of the
+sampling, \proglang{Stan} also can perform optimization (i.e., computation of the
 posterior mode).  We are currently working on implementing other
@@
 
-Stan was and continues to be supported largely through grants from the
+\proglang{Stan} was and continues to be supported largely through grants from the
 U.~S. government.  Grants which indirectly supported the initial
@@
 through a grant from the National Institutes of Health
-(1G20RR030893-01).  Stan is currently supported in part by a grant
+(1G20RR030893-01).  \proglang{Stan} is currently supported in part by a grant
 from the National Science Foundation (CNS-1205516).

### 2014-02-18 src/docs/papers/stan-language/article.tex
updated stan paper

@@
   imperatively defines a log probability function over parameters
-  conditioned on specified data and constants.  As of version 2.1,
+  conditioned on specified data and constants.  As of version 2.2.0,
   \proglang{Stan} provides full Bayesian inference for continuous-variable
@@
 
-These models turned out to be a challenge for
-existing general purpose inference software.  A direct encoding in
-\proglang{BUGS} \citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
-\proglang{JAGS} \citep{Plummer:2003} can grind these tools to a halt.
-We began by adding custom vectorized logistic regressions to \proglang{JAGS}
-using \proglang{C++} to overcome the cost of interpretation.  Although
-this is much faster than looping in \proglang{JAGS}, it quickly became
-clear that the root of the problem was the slow convergence of
-conditional sampling when parameters were highly correlated 
-in the posterior, as for time-series models and hierarchical models 
-with interacted predictors.  We finally realized we needed a better sampler, 
-not a more efficient implementation of Gibbs sampling.
+These models turned out to be a challenge for existing general purpose
+inference software.  A direct encoding in \proglang{BUGS}
+\citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
+\proglang{JAGS}~\citep{Plummer:2003} can grind these tools to a halt.
+We began by adding custom vectorized logistic regressions to
+\proglang{JAGS} using \proglang{C++} to overcome the cost of
+interpretation.  Although this is much faster than looping in
+\proglang{JAGS}, it quickly became clear that the root of the problem
+was the slow convergence of conditional sampling when parameters were
+highly correlated in the posterior, as for time-series models and
+hierarchical models with interacted predictors.  We finally realized
+we needed a better sampler, not a more efficient implementation of
+Gibbs sampling.
 
@@
 
-We thought we were home free at this point, but when we measured the
+We thought we were home free, but when we measured the
 speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
@@
 Gradient evaluation took 4e-06 seconds
-1000 transitions using 10 leapfrog steps per transition would take 0.04 seconds.
+1000 transitions using 10 leapfrog steps per transition would take
+0.04 seconds.
 Adjust your expectations accordingly!
@@
 The output contiues with a CSV header naming the columns of the
-output. For the default NUTS sampler in \proglang{Stan} 2.1.0, these are
+output. For the default NUTS sampler in \proglang{Stan} 2.2.0, these are
 %
@@
 \clearpage
-%\bibliographystyle{jss}
+% implicit: \bibliographystyle{jss}
 \bibliography{stan-paper}

### 2014-03-01 src/docs/papers/stan-language/article.tex
added corrections based on comments from Aki

@@
   calculated using optimization methods such as the
-  Broyden–Fletcher–Goldfarb–Shanno algorithm.
+  Broyden-Fletcher-Goldfarb-Shanno algorithm.
 
@@
 intended to have unconstrained support (i.e., no points of zero
-probabilty), which greatly simplifies the task of writing samplers or
+probability), which greatly simplifies the task of writing samplers or
 optimizers.  To achieve unbounded support, variables declared with
@@
 matrices (Cholesky factorization).  The dimensionality of the
-resulting probabilty function may change as a result of the
+resulting probability function may change as a result of the
 transform. For example, a $K \times K$ covariance matrix requires only
@@
 absolute Jacobian determinant of the inverse transform is added to the
-overall log probabilty function.%
+overall log probability function.%
 %
@@
 %
-The gradients of the log probabilty function exposed include the
+The gradients of the log probability function exposed include the
 Jacobian term.
@@
   real sum_of_squares;
-  sum <- 0;
+  sum_of_squares <- 0;
   for (n in 1:N)
-    sum <- sum + y[n] * y[n];
+    sum_of_squares <- sum_of_squares + y[n] * y[n];
 }  
@@
 parameters {
-  vector[K] beta_raw;    
+  ...
+  vector[K] beta_raw;  // replaces beta as parameter  
 }
 transformed parameters {
-  vector[K] beta;
+  vector[K] beta;      // beta now a transformed parameter
   beta <- mu + sigma * beta_raw;
@@
 model {
-  beta ~ normal(0,1);
+  beta_raw ~ normal(0,1);  // beta now sampled from unit normal
+  ...
 }
@@
 Vectors are declared with their sizes and matrices with their number
-of rows and columns.
+of rows and columns.  Vector, row vector, and matrix elements are
+accessed using bracket notation, as in \code{y[3]} for the third
+element of a vector or row vector and \code{a[2,3]} for the element in
+the third column of the second row of a matrix.  Indexing begins from
+1.
+
+The notation \code{a[2]} accesses the second row of matrix \code{a}.
+Rows, columns and subblocks of matrices can be accessed using
+\code{row}, \code{col}, and \code{block} functions.  Slices of arrays
+can be accessed using the \code{head}, \code{tail}, and \code{segment}
+functions.  There are also special functions for creating a diagonal
+matrix from a vector and accessing the diagonal of a vector.
 
@@
   standard library containers in \proglang{C++}.}
+
+The manual contains a chapter discussing the efficiency tradeoffs and
+motivations for separating arrays and matrices.
 
@@
 The default optimizer in \proglang{Stan} is the
-Broyden-Fletcher-Goldfarb-Shannon-Boyd (BFGS) optimizer.  BFGS is a
+Broyden-Fletcher-Goldfarb-Shanno (BFGS) optimizer.  BFGS is a
 quasi-Newton optimizer that evaluates gradients directly, then uses
@@
 served as trial subjects for both \proglang{Stan} and \citep{GelmanEtAl:2013}.
+
+We'd like to thank Aki Vehtari for comments and corrections on a draft
+of the paper.
 

### 2014-03-10 src/docs/papers/agrad/stan-autodiff.tex
first version of start of agrad paper

\documentclass[10pt]{article}
\usepackage{stan-papers}

\title{\Large\bfseries \code{\bfseries agrad}: A C++ Library for Efficient, Scalable,
  and Extensible Automatic Differentiation}
\author{\large Bob Carpenter \\ {\small Columbia University}
        \\[8pt]
        \large Marcus Brubaker \\ {\small Toyota Technical Institute}
   \and \large Matt Hoffman \\ {\small Adobe Research Labs}
        \\[8pt]
        \large Peter Li \\ {\small Columbia University}
   \and \large Daniel Lee \\ {\small Columbia University}
        \\[8pt]
        \large Michael Betancourt \\ {\small University of Warwick}
}
\date{\vspace*{8pt}\normalsize \today}

\begin{document}
\maketitle

\begin{abstract} 
  \noindent
  The \code{agrad} C++, forward- and reverse-mode automatic
  differentiation library was defined to satisfy
  six competing goals: extensibility, efficiency, scalability,
  stability, portability, and usability. Its underlying
  object-oriented design
  and functional template abstraction supports the addition of new
  functions based on value and first-order derivative calculations
  abstracted from the underlying automatic differentiation mechanisms.
  It uses a custom memory arena and template metaprogramming for
  efficiency.  Its underlying reverse-mode 
  evaluation is lazy for both efficiency and scalability.
  First- and higher-order derivatives are avaialable through a
  functional abstraction which hides all of the details of the
  library's internals. For broad applicability and numerical
  stability, it implements an extensive library of special functions,
  matrix and linear algebra operations, probability functions, 
  and ordinary differential equation solvers.  Evaluations on test
  problems show that both its efficiency and scalability are state of
  the art.  The \code{agrad} library is open-source licensed under the
  new BSD license and has been tested for all major compilers for
  Windows, Mac OS X, and Linux. 
\end{abstract}

\section{Introduction}

Many scientific computing applications, including optimization,
simulation, and differential equation solvers, require the computation
of first-order and higher-order derivatives over functions defined on
an application-by-application basis.

Although derivatives can be calculated and coded by hand, the process
is both tedious and error prone, especially for algorithms represented
in large amounts of code. Two popular alternatives are finite
differencing and automatic differentiation. Finite differencing calculates a
derivative of a function $f$ at $x$ by the linear approximation
\[ 
\frac{d}{dx}f(x) 
\ \approx \
\frac{f(x + \epsilon) - f(x - \epsilon)}{2\epsilon}.
\]
Automatic differentiation directly differentiates a computer program
for $f$ by applying the chain rule to the expressions produced by the
program's statements.

In contrast to finite differencing methods, automatic differentiation
is efficient and accurate. As an example of efficiency, gradient
calculations require only a constant multiple of the time taken to
evaluate a function independent of dimension, whereas finite
differences require a multiple determined by the dimensionality.
Automatic differentiation computes derivatives up to machine
precision, whereas finite differences are notoriously imprecise
because they require a tradeoff of arithmetic precision and accuracy
in linear approximations that depends on scale and rarely allows more
than 1e-6 relative accuracy even with scale-based tuning.

The \code{agrad} library was introduced to overcome several
shortcomings of existing implementations such as CppAD, Saccado,
AdolC, and Adept. The main motivation was to allow easy extensibility
to enable the implementation of a wide range of special functions,
probability density, mass and cumulative distribution functions, and
matrix and linear-algebra operations. A secondary motivation was to
increase scalability and efficiency, particularly for the calculations
of log probability functions as required for statistical inference
through optimization and sampling. A third motivation was to provide
simple interfaces that were portable and required no
platform-dependent installation and could be both used and extended to
new functions without understanding the internal differentiation data
types. The rest of the paper describes how \code{agrad} is designed
and implemented and why it improves the state of the art in
efficiency, scalability, and usability.

\section{Derivative Calculation Interface}

The \code{agrad} library application programmer interface (API) is
designed to allow client code to calculate derivatives and extend the
built-in library without needing to know anything about the internals
of the library's underlying automatic differentiation data types.
This section describes how to calculate derivatives and the next how
to add user-defined functions.

To use the \code{agrad} library to calculate derivatives, client code
need only define a templated functor, define an argument vector,
define the appropriate vector or matrix to hold the result, and then
call the derivative functional of the apporpriate order.

The library also exposes an interface for direct access to the
variables and methods used for automatic differentiation for
finer-grained control and integration with other software.

\subsection{Example Function: Neal's Funnel}

The log of the funnel probability density function is a simple example
of the kinds of probability functions \code{agrad} was designed to
differentiation.  
kinds of functions
Consider the log of the funnel density
\citep{Neal:2003}, which is defined for
a positive scalar $v$ and $N$-vector $y$ by
\[
\mbox{funnel}(v,y) = \frac{1}{2} \left(
  N v
  + \frac{v^2}{9}
  + \exp(-v) \sum_{n=1}^N y_n^2
\right)
\]
\reffigure{funnel-cpp} provides a C++ implementation of the
funnel function suitable for use with \code{agrad}.
%
\begin{figure}
\begin{quote}
\begin{Verbatim}
struct funnel {
  template <typename T>
  T operator()(const std::vector<T>& x) const {
    T v = x[0];
    T sum_y_sq = 0;
    for (size_t n = 1; n < x.size(); ++n)
      sum_y_sq += x[n] * x[n];
    T p1 = N * v;
    T p2 = (v * v) / 9;
    T p3 = sum_y_sq * exp(-v);
    return (p1 + p2 + p3) / 2;
  }
};
\end{Verbatim}
\end{quote}
\vspace*{-6pt}
\caption{\it The funnel density implemented as a C++ functor operating
  on a C++ standard template library vector containing $v$ followed by
  $y$. The two quadratic terms could also be expressed using \code{agrad}'s
  built-in functions\, \code{square} or\, \code{pow}.  The function
  could alternatively be defined in terms of any class definining a
  \code{size()} method and allowing \code{operator[]} indexing, such 
  as an Eigen library vector or C++11 standard library array.
}\label{funnel-cpp.figure}
\end{figure}
%

\subsection{Derivative Calculation}

Derivatives can be calculated using either C++ functionals operating
on functions defined as functors, or directly using the underlying
automatic differentiation data types.

\subsubsection{Functional Approach}

The \code{agrad} library supports a functional approach to calculating
derivatives which is most easily explained with an example. The
following code calculates the value \code{fx} and gradient \code{grad}
of the funnel functor \code{f} evaluated at the argument vector
\code{x}.
%
\begin{quote}
\begin{Verbatim}
funnel f;
std::vector<double> x = ...;
double fx;
std::vector<double> grad;
agrad::gradient(f,x,fx,grad);
\end{Verbatim}
\end{quote}

\subsubsection{Direct Approach}

The \code{agrad} library also supplies the same kind of low-level
methods as other automatic differentiation libraries such as Adept,
CppAD, and Saccado, although without the need for turning on or off
``recording'' to explicitly manage memory. For example, the following
function defines the log of the Beta density function
%
\begin{eqnarray*}
f(x,\alpha,\beta) 
& = & \log \distro{Beta}(x|\alpha,\beta)
\\
& = & (\alpha - 1) \log x + (\beta - 1) \log (1 - x) 
  \log \Gamma(\alpha + \beta) - \log \Gamma(\alpha) - \log
  \Gamma(\beta),
\end{eqnarray*}
%
which can be implemented using built-in special functions as
%
\begin{quote}
\begin{Verbatim}
agrad::var f(double x, 
             const agrad::var& a, const agrad::var& b) {
  return (a - 1) * log(x) + (b - 1) * log1m(x)
         + lgamma(a + b) - lgamma(a) - lgamma(b);
}
\end{Verbatim}
\end{quote}
%
To calculate the derivatives of $y = f(x,\alpha,\beta)$ with respect to
$\alpha,\beta$ at
$\alpha = 1.9$, $\beta = 17.2$, and  $x = 0.329$, the code is
%
\begin{quote}
\begin{Verbatim}
agrad::var a = 1.9;    agrad::var b = 17.2;
double x = 0.329;
agrad::var y = f(x,a,b);
fab.grad();
double dy_da = a.adjoint();    double dy_db = b.adjoint();
\end{Verbatim}
\end{quote}
%
This approach can be used for arbitrary functions defined for
arguments including \code{var}; see \refsection{flexible-signatures}
for an example of how the beta density function is defined in
\code{agrad}.

\section{Flexible Type Signatures}\label{flexible-signatures.section}

To allow for maximum flexibility and efficiency in calling functions, 
\code{agrad}'s signature for the log Beta density is declared as
%
\begin{quote}
\begin{Verbatim}
template <typename Ty, typename Ta, typename Tb>
inline typename agrad::return_type<Ty,Ta,Tb>::type
beta_log(const Ty& y, const Ta& a, const Tb& b); 
\end{Verbatim}
\end{quote}
%
where a mechanism like that of Boost library's promotion template
metaprogram is used to calculate the return type automatically as the
most general type able to hold the result (e.g., \code{agrad::var} if
used in place of \code{f} in the previous example). In general,
\code{agrad}'s density functions are vectorized to allow any argument
to be a scalar, standard vector, or Eigen vector of any primitive type
or of type \code{var} (or of type \code{fvar<T>}, the forward-mode
automatic differentiation variable type).  The need for this
flexibility is why a simple log density library like Boost's, which
template on a single type, cannot be directly used for \code{agrad}.
%
\footnote{For maximum portability, the \code{agrad} library does not
  depend on C++11 extensions, but even if it did, the \code{auto}
  mechanism is not powerful to determine the result type here.}

\section{Reverse-Mode Automatic Differentiation}

\appendix

\section{Functional Derivative API}

The high-level \code{agrad} API for calculating derivatives is defined
in terms of C++ functionals, i.e., functions operating on functions.%
%
\footnote{To simplify notation, \code{Vector<T>} is used as shorthand
for \code{Eigen::Matrix<T,Eigen::Dynamic,1>} and 
\code{Matrix<T>} for \code{Eigen::Matrix<T,Eigen::Dynamic,Eigen::Dynamic,1>}.}

\subsection{Simple Derivatives}

If \code{f} is a functor implementing the signature
%
\begin{quote}
\begin{Verbatim}
template <typename T>
T operator()(const T& x) const;
\end{Verbatim}
\end{quote}
%
and \code{x} is a point,
%
\begin{quote}
\begin{Verbatim}
double x = ...;  
\end{Verbatim}
\end{quote}
%
then
%
\begin{quote}
\begin{Verbatim}
double fx, df_dx;  
agrad::derivative(f,x,fx,df_dx);
\end{Verbatim}
\end{quote}
%
sets \code{fx} to be the value of \code{f} at \code{x} and
sets \code{df\_dx} to be the derivative of \code{f} at \code{x}.

\subsection{Gradients and Hessians}

If \code{f} is a functor implementing the signature
%
\begin{quote}
\begin{Verbatim}
template <typename T>
T operator()(const Vector<T>& x) const;
\end{Verbatim}
\end{quote}
%
and \code{x} is a vector of inputs,
%
\begin{quote}
\begin{Verbatim}
Vector<double> x = ...;
\end{Verbatim}
\end{quote}
%
then
%
\begin{quote}
\begin{Verbatim}
double fx;    Vector<double> g;
agrad::gradient(f,x,fx,grad);
\end{Verbatim}
\end{quote}
%
sets \code{fx} to be the value of \code{f} at \code{x} and sets
\code{grad} to be the gradient of \code{f} at \code{x} and
%
\begin{quote}
\begin{Verbatim}
double fx;    Matrix<double> H;
agrad::hessian(funnel(),x,H);
\end{Verbatim}
\end{quote}
%
sets \code{fx} to be the value of \code{f} at \code{x} and sets
\code{H} to be the Hessian of \code{f} at \code{x}.

\subsection{Directional Gradients and Hessians}

There are also directional forms of gradients and Hessians, so that if
\code{v} is a vector representing a direction, 
%
\begin{quote}
\begin{Verbatim}
Vector<double> v = ...;
\end{Verbatim}
\end{quote}
%
then
%
\begin{quote}
\begin{Verbatim}
double fx;    double grad_v;
agrad::gradient_dot_vector(f,x,v,fx,grad_v);
\end{Verbatim}
\end{quote}
%
sets \code{fx} to be the value of \code{f} at \code{x} and sets
\code{grad\_v} to be the product of the gradient of \code{f} at
\code{x} and the vector \code{v}, and
%
\begin{quote}
\begin{Verbatim}
double fx;    Vector<double> H_v;
agrad::gradient_dot_vector(f,x,v,fx,Hv);
\end{Verbatim}
\end{quote}
%
sets \code{fx} to be the value of \code{f} at \code{x} and sets
\code{H\_v} to be the product of the Hessian of \code{f} at
\code{x} and the vector \code{v}.

\subsection{Jacobians}

If \code{f} is a functor implementing the signature
%
\begin{quote}
\begin{Verbatim}
template <typename T>
Vector<T> operator()(const Vector<T>& x) const;
\end{Verbatim}
\end{quote}
%
and \code{x} is an input vector, then
%
\begin{quote}
\begin{Verbatim}
VectorXd x = ...;
\end{Verbatim}
\end{quote}
%
then
%
\begin{quote}
\begin{Verbatim}
VectorXd fx;    MatrixXd J;
agrad::jacobian(f,x,fx,J);
\end{Verbatim}
\end{quote}
%
sets \code{fx} to be the value of \code{f} at \code{x} and 
\code{J} to be the Jacobian of \code{f} at \code{x}.

\bibliographystyle{apalike}
\bibliography{../../bibtex/all}

\end{document}

### 2014-06-23 src/docs/talks/paris-dauphine-2014/stan-overview.tex
added complete talk

\documentclass[10pt]{report}

\usepackage[
  scale=0.875,
  stdmathitalics=true,
  stdmathdigits=true]{lucimatx}
\linespread{1.02}
%\usepackage{helvet}

\usepackage{fancyvrb}
\usepackage{amsmath}
% \usepackage{amssymb}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\setlength{\paperheight}{3in}
\setlength{\paperwidth}{4in}
\pdfpagewidth=\paperwidth
\pdfpageheight=\paperheight

\setlength{\textwidth}{3.75in}
\setlength{\textheight}{2.5in}

\setlength{\oddsidemargin}{-1.0in}
\setlength{\evensidemargin}{-1.0in}
\setlength{\topmargin}{-1.25in}

\newcommand{\sld}[1]{\newpage{\noindent\LARGE \ \ \
    \textcolor{MidnightBlue}{\bfseries #1}}\vspace*{4pt}}

\newcommand{\code}[1]{{\tt #1}}
\newcommand{\spc}{\hspace*{0.25in}}

\newcounter{gmlrx}
\newcounter{gmlry}
\newcommand{\gmnode}[3]{\put(#1,#2){\circle{20}}\put(#1,#2){\makebox(0,0){$#3$}}}
\newcommand{\gmplate}[5]{
\setcounter{gmlrx}{#1}\addtocounter{gmlrx}{#3}
\setcounter{gmlry}{#2}\addtocounter{gmlry}{-#4}
\put(#1,#2){\line(1,0){#3}}
\put(#1,#2){\line(0,-1){#4}}
\put(\value{gmlrx},\value{gmlry}){\line(-1,0){#3}}
\put(\value{gmlrx},\value{gmlry}){\line(0,1){#4}}
\setcounter{gmlrx}{#1}\addtocounter{gmlrx}{5}
\setcounter{gmlry}{#2}\addtocounter{gmlry}{-6}
\put(\value{gmlrx},\value{gmlry}){\makebox(0,0){$#5$}}
}

\newcommand{\myemph}[1]{{\color{MidnightBlue}{\bfseries #1}}}
\newcommand{\mypart}[2]{{\newpage 
\mbox{ }
\vfill
\noindent\spc\color{MidnightBlue}{\LARGE\bfseries #1\\[10pt]\spc\Huge{#2}}
\vfill\vfill}
\mbox{ }}

\begin{document}
\sf%
\vspace*{6pt}
%
\noindent
\spc{\huge\bfseries \color{MidnightBlue}{Stan}}
\\[8pt]
\spc{\Large\bfseries \color{MidnightBlue}{Probabilistic Programming Language}}
\\
\vfill
\noindent
\spc\spc{\slshape\footnotesize Core Development Team:}
\\[2pt]
\spc\spc{\small Andrew Gelman, \  \myemph{Bob Carpenter}, \  Matt Hoffman}
\\
\spc\spc{\small Daniel Lee, \  Ben Goodrich, \  Michael Betancourt, \ }
\\
\spc\spc{\small Marcus Brubaker, \   Jiqiang Guo, \  Peter Li, \ }
\\
\spc\spc{\small Allen Riddell, \  Marco Inacio}
\vfill
\mbox{ } 
\hfill
\includegraphics[width=27.5pt]{img/stanlogo-main.pdf}\hspace*{9pt}
\\
\spc{\footnotesize Paris-Dauphine 2014}
\hfill 
{\footnotesize \tt mc-stan.org}

\sld{What is Stan?}

\begin{itemize}
\item Stan is an \myemph{imperative} probabilistic programming language
\vspace*{-12pt}
\begin{itemize}\footnotesize
\item  cf., BUGS: declarative; \ Church: functional; \ Figaro: object-oriented
\end{itemize}
\item Stan \myemph{program}
\vspace*{-4pt}
\begin{itemize}\small
\item declares variables
\item codes log posterior (penalized likelihood)
\end{itemize}
\item Stan \myemph{inference}
\vspace*{-4pt}
\begin{itemize}\small
\item MCMC for full Bayesian inference
\item MLE for point estimation
\end{itemize}
\item Stan is \myemph{open source} \ {\footnotesize (BSD core C++, GPLv3 interfaces)}
\\
{\footnotesize hosted on GitHub; uses Eigen matrix lib, Boost, googletest
  C++ lib}
\end{itemize}

\sld{Who's Using Stan?}
\begin{itemize}
\item 750+ registrations for mailing list across the physical,
biomedical, and social sciences, plus engineering, finance and marketing

\item Subjects of Papers on Google Scholar in the sciences
\\
{\footnotesize clinical drug
trials, general computational statistics, entomology, opthalmology,
neurology, sociology and population dynamics, genomics, agriculture,
psycholinguistics, molecular biology, population dynamics, materials
engineering, botany, astrophysics, oceanography, election prediction,
fisheries, cancer biology, public health and epidemiology, population
ecology, collaborative filtering for recommender systems, climatology,
educational testing, natural language processing}
\end{itemize}

\sld{Books and Model Sets}
%
\begin{itemize}
\item 400+ page core language tutorial manual
\item separate installation, getting started, interface manuals 
\\ {\footnotesize  RStan, PyStan, CmdStan}
\item BUGS and JAGS examples (all 3 volumes), 
\item Gelman and Hill, {\slshape Data Analysis Using Regression and
    Multilevel/Hierarchical Models}
\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
\item two other books in progress
\end{itemize}

\mypart{Part I}{Stan Front End}

\sld{Interface Languages}

\begin{itemize}
\item \myemph{Platforms}
\\
Linux, Mac OS X, Windows
\item \myemph{C++ API}
\\
{\footnotesize portable, standards compliant (C++03 now, moving to C++11)}
\item \myemph{Interfaces}
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item \myemph{CmdStan}: Command-line or shell interface (direct executable)
\item \myemph{RStan}: R interface (Rcpp in memory)
\item \myemph{PyStan}: Python interface (Cython in memory)
\item \myemph{MStan$^*$}: MATLAB interface (lightweight external process)
\item \myemph{JuliaStan$^*$}: Julia interface (lightweight external process)
\end{itemize}
{\footnotesize ${}^*$ User contributed}
\end{itemize}
\vfill

\sld{Example: \ Bernoulli}
\vfill
\spc\spc
\begin{minipage}[t]{0.8\textwidth}
\begin{Verbatim}
data {
  int<lower=0> N;
  int<lower=0,upper=1> y[N];
}
parameters {
  real<lower=0,upper=1> theta;
} 
model {
  y ~ bernoulli(theta);
}
\end{Verbatim}
\end{minipage}
\vfill
\vfill
\vfill
\mbox{ } \hfill {\small\slshape notes: {\tt theta} uniform on $[0,1]$, \ {\tt y} vectorized}
\vfill

\sld{RStan Execution}

\begin{minipage}[t]{\textwidth}
\footnotesize
\begin{Verbatim}
> N <- 5;   y <- c(0,1,1,0,0);
> fit <- stan("bernoulli.stan", data = c("N", "y"));
> print(fit, digits=2)
\end{Verbatim}
\vspace*{1pt}
\begin{Verbatim}[fontshape=sl]
Inference for Stan model: bernoulli.
4 chains, each with iter=2000; warmup=1000; thin=1;

       mean  se_mean    sd   2.5%    50%  97.5%  n_eff  Rhat
theta  0.43     0.01  0.18   0.11   0.42   0.78   1229     1
lp__  -5.33     0.02  0.80  -7.46  -5.04  -4.78   1201     1
\end{Verbatim}
\vspace*{6pt}
\begin{Verbatim}
> hist( extract(fit)$theta )
\end{Verbatim}
\vspace*{-24pt}
\hfill\includegraphics[width=1.25in,height=0.9in]{img/bernoulli-posterior-histo.pdf}
\hspace*{24pt}
\end{minipage}

\sld{Basic Program Blocks}

\begin{itemize}
\item \myemph{\tt\bfseries data} \ (once) 
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare data types, sizes, and constraints
\item {\slshape execute}: read from data source, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries parameters} \ (every log prob eval)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare parameter types, sizes, and constraints
\item {\slshape execute}: transform to constrained, Jacobian
\end{itemize}
%
\item \myemph{\tt\bfseries model} \ (every log prob eval) 
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: statements definining posterior density
\item {\slshape execute}: execute statements
\end{itemize}
\end{itemize}

\sld{Derived Variable Blocks}

\begin{itemize}
\item \myemph{\tt\bfseries transformed data} (once after data)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define transformed data variables
\item {\slshape execute}: execute definition statements, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries transformed parameters} (every log prob eval)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define transformed parameter vars
\item {\slshape execute}: execute definition statements, validate constraints
\end{itemize}
%
\item \myemph{\tt\bfseries generated quantities} (once per draw, 
  \code{double} type)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define generated quantity
  variables
\\
{\footnotesize (for posterior predictions, event probabilities,
  decision making)}
\item {\slshape execute}: execute definition statements, validate constraints
\end{itemize}
%
\end{itemize}

\sld{User-Defined Functions (Stan 2.3)}

\begin{itemize}
\item \myemph{\tt\bfseries functions} \ (compiled with model)
\vspace*{-4pt}
\begin{itemize}\small
\item {\slshape content}: declare and define functions
\item {\slshape execute}: compile with model
\end{itemize}
\vspace*{12pt}
\item Example
\\[6pt]
\begin{minipage}[t]{0.8\textwidth}
\footnotesize
\begin{Verbatim}
functions {

  real relative_difference(real x, real y) {
    return 2 * fabs(x - y) / (fabs(x) + fabs(y));
  }

}
\end{Verbatim}
\end{minipage}
\end{itemize}

\sld{Variable and Expression Types}
\\[3pt]
\hspace*{17pt}Variables and expressions are \myemph{strongly, statically typed}.
\begin{itemize}
\item \myemph{Primitive}: {\tt\small int}, \ {\tt\small real}
\item \myemph{Matrix}: {\tt\small matrix[M,N]}, \ {\tt\small vector[N]}, \ {\tt\small row\_vector[M]}
\item \myemph{Bounded}: primitive or matrix, with 
\\ {\tt\small <lower=L>}, \ {\tt\small <upper=U>}, \ {\tt\small <lower=L,upper=U>}
\item \myemph{Constrained Vectors}: {\tt\small simplex[K]}, \ {\tt\small
    ordered[N]},
\\ {\tt\small positive\_ordered[N]}, \ {\tt\small unit\_length[N]}
\item \myemph{Constrained Matrices}: {\tt\small cov\_matrix[K]}, \ {\tt\small
    corr\_matrix[K]}, \ {\tt\small cholesky\_factor\_cov[M,N]}, \
  {\tt\small cholesky\_factor\_corr[K]}
\item \myemph{Arrays:}  of any type (and dimensionality)
\end{itemize}

\sld{Logical Operators}
\vfill
\noindent\spc
{\footnotesize
\begin{tabular}{c|ccl|l}
{\it Op.} & {\it Prec.} & {\it Assoc.} & {\it
  Placement} & {\it Description}
\\ \hline \hline
\code{||} & 9 & left & binary infix & logical or
\\ \hline
\Verb|&&| & 8 & left & binary infix & logical and
\\ \hline
\Verb|==| & 7 & left & binary infix & equality
\\
\Verb|!=| & 7 & left & binary infix & inequality
\\ \hline
\Verb|<| & 6 & left & binary infix & less than
\\
\Verb|<=| & 6 & left & binary infix & less than or equal
\\
\Verb|>| & 6 & left & binary infix & greater than 
\\
\Verb|>=| & 6 & left & binary infix & greater than or equal
\end{tabular}
}
\vfill
\vfill

\sld{Arithmetic and Matrix Operators}
\vfill
\noindent\spc
{\footnotesize
\begin{tabular}{c|ccl|l}
{\it Op.} & {\it Prec.} & {\it Assoc.} & {\it
  Placement} & {\it Description}
\\ \hline \hline

\code{+} & 5 & left & binary infix & addition
\\
\code{-} & 5 & left & binary infix & subtraction
\\ \hline
\code{*} & 4 & left & binary infix & multiplication
\\
\code{/} & 4 & left & binary infix & (right) division
\\ \hline
\Verb|\| & 3 & left & binary infix & left division
\\ \hline
\code{.*} & 2 & left & binary infix & elementwise multiplication
\\
\code{./} & 2 & left & binary infix & elementwise division
\\ \hline
\code{!} & 1 & n/a & unary prefix & logical negation
\\
\code{-} & 1 & n/a & unary prefix & negation
\\ 
\code{+} & 1 & n/a & unary prefix & promotion (no-op in Stan)
\\ \hline
\Verb|^| & 2 & right & binary infix & exponentiation
\\ \hline
\code{'} & 0 & n/a & unary postfix & transposition
\\ \hline \hline
\code{()} & 0 & n/a & prefix, wrap & function application
\\
\code{[]} & 0 & left & prefix, wrap & array, matrix indexing
\end{tabular}
}
\vfill

\sld{Built-in Math Functions}

\begin{itemize}
\item All built-in \myemph{C++ functions}
\\
{\footnotesize C math, TR1, C++11, including all trig, pow, and
  special log1m, etc.}
\item Extensive library of \myemph{statistical functions}
\\
{\footnotesize e.g., softmax,
log gamma and digamma functions, beta functions, Bessel functions}
%
\item Efficient, arithmetically stable \myemph{compound functions}
\\
{\footnotesize e.g., multiply log, log sum of
exponentials, log inverse logit}
\end{itemize}

\sld{Built-in Matrix Functions}

\begin{itemize}
\item \myemph{Basic arithmetic}: all arithmetic operators
\item \myemph{Elementwise arithmetic}: vectorized operations
\item \myemph{Solvers}: matrix left and right division, (log) determinant,
inverse 
\item \myemph{Decompositions}: QR, Eigenvalues and Eigenvectors, 
\\
Cholesky factorization, singular value decomposition
\item \myemph{Compound Operations}: quadratic forms, variance scaling
\vfill
\item \myemph{Ordering, Slicing, Broadcasting}: sort, rank, block, rep
\item \myemph{Reductions}: sum, product, norms
\end{itemize}

\sld{Distribution Library}

\begin{itemize}
\item Each distribution has
\vspace*{-4pt}
\begin{itemize}\small
\item log density or mass function
\item cumulative distribution functions, plus complementary versions,
  plus log scale
\item pseudo Random number generators
\end{itemize}
\item Alternative parameterizations
\\
{\footnotesize (e.g., Cholesky-based multi-normal,
log-scale Poisson, logit-scale Bernoulli)}
\item New multivariate correlation matrix density: LKJ
\\
{\footnotesize degrees of freedom controls 
shrinkage to (expansion from) unit matrix}
\end{itemize}

\sld{Statements}

\begin{itemize}
\item \myemph{Sampling}: \ {\footnotesize \Verb|y ~ normal(mu,sigma)|}
\ \ \ {\footnotesize (increments log probability)}
%
\item \myemph{Log probability}: \ {\footnotesize increment\_log\_prob(lp);}
\item \myemph{Assignment}: \  {\footnotesize \code{y\_hat <- x * beta};}
%
\item \myemph{For loop}: \ {\footnotesize \code{for (n in 1:N) ...}}
%
\item \myemph{While loop}: \ {\footnotesize \code{while (cond) ...}}
%
\item \myemph{Conditional}: \ {\footnotesize
\code{if (cond) ...; else if (cond) ...;  else ...;}}
\item \myemph{Block}: \ {\footnotesize \Verb|{ ... }|}  \ \ \ {\footnotesize
   (allows local variables)}
\item \myemph{Print}: \ {\footnotesize print("theta=",theta);}
\end{itemize}

\sld{Full Bayes with MCMC}

\begin{itemize}
\item Adaptive \myemph{Hamiltonian Monte Carlo} (HMC)
%
\item Adaptation \myemph{during warmup}
\vspace*{-4pt}
\begin{itemize}\small
\item step size adapted to target Metropolis acceptance rate
\item mass matrix estimated with regularization
{\footnotesize
\\ sample covariance of second half of warmup iterations
\\ (assumes constant posterior curvature)
}
\end{itemize}
%
\item Adaptation \myemph{during sampling}
\vspace*{-4pt}
\begin{itemize}\small
\item number of steps
\\
{\footnotesize aka no-U-turn sampler; aka NUTS}
\end{itemize}
%
\item \myemph{Initialization} user-specified or random unconstrained
\end{itemize}

\sld{Posterior Inference}

\begin{itemize}
\item Generated quantities block for predictions, decisions,
  and event probabilities
\item Extractors for samples in RStan and PyStan
\item Coda-like posterior summary
\vspace*{-4pt}
\begin{itemize}\small
\item posterior mean w.\ standard error, standard deviation, quantiles
\item split-$\hat{R}$ multi-chain convergence diagnostic (Gelman and ...)
\item multi-chain effective sample size estimation (FFT algorithm)
\end{itemize}
\item Model comparison with WAIC
\\
{\footnotesize (internal log likelihoods; external cross-sample statistics)}
\end{itemize}

\sld{Penalized MLE}

\begin{itemize}
\item Posterior mode finding via BFGS optimization
\\ {\footnotesize (uses model gradient, efficiently approximates Hessian)}
\item Disables Jacobians adjusting for parameter transforms
\item Models, data, initialization as in MCMC
\vfill
\item Near Future:
\vspace*{-4pt}
\begin{itemize}\small
\item  Standard errors on unconstrained scale
\\
{\footnotesize  (estimated using curvature of penalized log likelihood function}
\item Standard errors on constrained scale)
\\
{\footnotesize  (sample unconstrained approximation and inverse transform)}
\item L-BFGS optimizer
\end{itemize}      
\end{itemize}

\sld{Stan as a Research Tool}

\begin{itemize}
\item Stan can be used to explore statistical algorithms
\item Models transformed to have support on $\mathbb{R}^n$
\item Once a model is compiled, have
\vspace*{-4pt}
\begin{itemize}\small
\item log probability, gradient, and Hessian
\item data I/O and parameter initialization
\item model provides variable names and dimensionalities
\item transforms to and from constrained representation 
  \\ {\footnotesize (with or without Jacobian)}
\end{itemize}
\item Future:
\vspace*{-4pt}
\begin{itemize}\small
\item second- and higher-order derivatives via auto diff
\end{itemize}
\end{itemize}

\mypart{Part II}{Under the Hood}

\sld{Euclidean Hamiltonian}

\begin{itemize}
\item \myemph{Phase space}: $q$ position (parameters); \ $p$ momentum
\item \myemph{Posterior density}: $\pi(q)$
\item \myemph{Mass matrix}: $M$
\item \myemph{Potential energy}: $V(q) = -\log \pi(q)$
\item \myemph{Kinetic energy}: $T(p) = \frac{1}{2} p^{\top} M^{-1} p$
\item \myemph{Hamiltonian}:  $H(p,q) = V(q) + T(p)$
\item \myemph{Diff eqs}:
\[
\frac{dq}{dt} \ = \  + \frac{\partial H}{\partial p}
\hspace*{48pt}
\frac{dp}{dt} \ = \ - \frac{\partial H}{\partial q}
\]
\end{itemize}

\sld{Leapfrog Integrator Steps}
\begin{itemize}
\item Solves Hamilton's equations by \myemph{simulating dynamics}
\\
{\footnotesize (symplectic [volume preserving]; $\epsilon^3$ error per step, $\epsilon^2$ total error)}
\item Given: \myemph{step size} $\epsilon$, \myemph{mass matrix} $M$, \myemph{parameters} $q$
\item \myemph{Initialize kinetic} energy, $p \sim {\sf
    Normal}(0,\mbox{\bf I})$
\item \myemph{Repeat} for $L$ leapfrog steps:
\begin{eqnarray*}
p & \leftarrow &
p - \frac{\epsilon}{2} \, \frac{\partial V(q)}{\partial q}
\ \ \ \ \ \ \mbox{[half step in momentum]}
\\[6pt]
q & \leftarrow &
q + \epsilon \, M^{-1} \, p
\ \ \ \ \ \ \ \  \mbox{[full step in position]}
\\[6pt]
p & \leftarrow &
p - \frac{\epsilon}{2} \, \frac{\partial V(q)}{\partial q}
\ \ \ \ \ \ \mbox{[half step in momentum]}
\end{eqnarray*}
\end{itemize}

\sld{Standard HMC}
\begin{itemize}
\item \myemph{Initialize parameters} diffusely
\\ {\footnotesize Stan's default: \ $q \sim \mbox{\sf
    Uniform}(-2,2)$ on unconstrained scale}
\item For each draw
\vspace*{-3pt}
\begin{itemize}\small
\item leapfrog integrator \myemph{generates proposal}
\item \myemph{Metropolis accept} step ensures detailed balance
\end{itemize}
\vfill
\item \myemph{Balancing act}: small $\epsilon$ has low error, requires many steps
\item Results \myemph{highly sensitive} to step size $\epsilon$ and mass matrix
  $M$
\end{itemize}

\sld{Tuning HMC During Warmup}

\begin{itemize}
\item \myemph{Chicken-and-egg} problem
\vspace*{-4pt}
\begin{itemize}\small
\item convergence to high mass volume requires adaptation 
\item adaptation requires convergence
\end{itemize}
\item During warmup, tune
\vspace*{-4pt}
\begin{itemize}\small
\item \myemph{step size}: line search to achieve target acceptance
  rate
\item \myemph{mass matrix}: estimate with second half of warmup
\end{itemize}
\item Use exponentially growing adaptation block sizes
\end{itemize}

\sld{Position-Independent Curvature}

\begin{itemize}
\item \myemph{Euclidean} HMC uses \myemph{global mass matrix} $M$
\item Works for densities with \myemph{position-independent curvature}
\item \myemph{Counterexample}: hierarchical model
\vspace*{-4pt}
\begin{itemize}\small
\item hierarchical variance parameter controls lower-level scale
\item mitigate by reducing target acceptance rate
\end{itemize}
\vfill
\item \myemph{Riemannian-manifold} HMC (coming soon)
\vspace*{-4pt}
\begin{itemize}\small
\item automatically adapts to varying curvature
\item no need to estimate mass matrix
\item need to regularize Hessian-based curvature estimate
\\ {\footnotesize (Betancourt arXiv; SoftAbs metric)}
\end{itemize}
\end{itemize}

\sld{Adapting HMC During Sampling}

\begin{itemize}
\item No-U-turn sampler (\myemph{NUTS})
\item Subtle algorithm to maintain \myemph{detailed balance}
\item Move randomly \myemph{forward or backward in time}
\item Double number of leapfrog steps each move (\myemph{binary tree})
\item Stop when a subtree makes a \myemph{U-turn}
\\ {\footnotesize (rare: throw away second half if not end to end U-turn)}
\item \myemph{Slice sample} points along last segment of path
\item \myemph{Generalized} to Riemannian-manifold HMC
\\ {\footnotesize (Betancourt, arXiv)}
\end{itemize}

\sld{Reverse-Mode Auto Diff}
\begin{itemize}
\item Eval gradient in small multiple of function eval time
\\
{\footnotesize (independent of dimensionality)}
\item Templated \myemph{C++ overload} for all functions
\item Code \myemph{partial derivatives} for basic operations
\item Function evaluation builds up \myemph{expression tree}
\item Dynamic program propagates \myemph{chain rule} in reverse pass
\item \myemph{Object-oriented} custom partial propagation
\item Arena-based \myemph{memory management}
\\ {\footnotesize (customize \code{operator new})}
\end{itemize}

\sld{Forward-Mode Auto Diff}
\begin{itemize}
\item Templated \myemph{C++ overload} for all functions
\item Code \myemph{partial derivatives} for basic operations
\item Function evaluation propagates chain rule forward
\item Nest reverse-mode in forward for higher-order
\vfill
\item Faster autodiff rewrite coming in six months to one year
\end{itemize}

\sld{Autodiff Functionals}

\begin{itemize}
\item Fully encapsulates autodiff in C++
\item Autodiff operations are functionals {\footnotesize (higher-order functions)}
\begin{itemize}\small
\item gradients, Jacobians, gradient-vector product
\item directional derivative
\item Hessian-vector product
\item Hessian
\item gradient of trace of matrix-Hessian product
\\ {\footnotesize (for SoftAbs RHMC)}
\end{itemize}
\item Functions to differentiate coded as functors (or pointers)
\\ {\footnotesize (use bind or lambda)}
\end{itemize}

\sld{Variable Transforms}
\begin{itemize}
\item Code HMC and optimization with $\mathbb{R}^n$ \myemph{support}
\item Transform constrained parameters to unconstrained
\vspace*{-2pt}
{\small
\begin{itemize}
\item lower (upper) bound: offset (negated) log transform
\item lower and upper bound: scaled, offset logit transform
\item simplex: centered, stick-breaking logit transform
\item ordered: free first element, log transform offsets
\item unit length: spherical coordinates
\item covariance matrix: Cholesky factor positive diagonal 
\item correlation matrix: rows unit length via quadratic stick-breaking
\end{itemize}
}
\end{itemize}

\sld{Variable Transforms (cont.)}
\begin{itemize}
\item Inverse transform from unconstrained $\mathbb{R}^n$
\item Evaluate log probability in model block on natural scale
\item Optionally adjust log probability for change of variables
\\ {\footnotesize (add log determinant of inverse transform Jacobian)}
\end{itemize}

\sld{Parsing and Compilation}
\begin{itemize}
\item Stan code \myemph{parsed} to abstract syntax tree (AST)
\\ {\footnotesize (Boost Spirit Qi, recursive descent, lazy semantic
  actions)}
\item C++ model class \myemph{code generation} from AST
\\ {\footnotesize (Boost Variant)}
\item C++ code compilation
\item Dynamic linking for RStan, PyStan
\end{itemize}

\sld{Coding Probability Functions}
\begin{itemize}
\item \myemph{Vectorized} to allow scalar or container arguments
\\ {\footnotesize (containers all same shape; scalars broadcast as necessary)}
\item Avoid \myemph{repeated computations}, e.g. $\log \sigma$ in
\hspace*{-18pt}
{\small
\begin{eqnarray*}
\textstyle \log \, \mbox{\sf Normal}(y | \mu, \sigma)
& = & \textstyle \sum_{n=1}^N \log \, \mbox{\sf Normal}(y_n | \mu,\sigma)
\\[4pt]
& = & \textstyle \sum_{n=1}^N  - \log \sqrt{2\pi} \ - \log \sigma \ -
\frac{\textstyle y_n - \mu}{\textstyle 2\sigma^2}
\end{eqnarray*}
}
\item \myemph{expression templates} to broadcast and cache scalars
\item \myemph{traits} metaprogram to \myemph{drop constants} (e.g., $-\log
  \sqrt{2 \pi}$) and calculate return types
\end{itemize}

\sld{Models with Discrete Parameters}
\begin{itemize}
\item e.g., simple mixture models, survival models, HMMs,
      discrete measurement error models, missing data
\item \myemph{Marginalize out} discrete parameters
\item Efficient sampling due to \myemph{Rao-Blackwellization}
\item Inference straightforward with expectations
\item \myemph{Difficult} for many users
\end{itemize}

\sld{Models with Missing Data}
\begin{itemize}
\item In principle, missing data just \myemph{additional parameters}
\item In practice, how to declare? 
\vspace*{-3pt}
\begin{itemize}\small
\item observed data as data variables
\item missing data as parameters
\item combine into single vector 
\\ {\footnotesize (in transformed parameters or local in model)}
\end{itemize}
\end{itemize}

\mypart{Part III}{What's Next?}

\sld{Differential Equation Solver}
\begin{itemize}
\item Auto-diff solutions w.r.t.\ parameters
\item Integrate coupled system for solution with partials
\item Auto-diff coupled Jacobian for stiff systems
\vfill
\item C++ prototype integrated for large PK/PD models
\\ {\footnotesize (project with Novartis for clinical trial)}
\end{itemize}

\sld{L-BFGS Optimizer}
\begin{itemize}
\item Faster, lower memory, more stable than BFGS
\item Plug-and-play in interfaces
\vfill
\item Code and doc complete; under testing
\end{itemize}

\sld{Higher-Order Auto-diff}
\begin{itemize}
\item Finish higher-order auto-diff for probability functions
\item May punt some cumulative distribution functions
\\
{\footnotesize (Black art iterative algorithms required)}
\vfill
\item Code complete; under testing
\end{itemize}

\sld{Riemannian Manifold HMC}
\begin{itemize}
\item \myemph{SoftAbs} metric
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item Eigendecompose Hessian
\item \myemph{positive definite} with positive eigenvalues
\item \myemph{condition} by narrowing eigenvalue range
\item Betancourt {\slshape arXiv}\ paper
\end{itemize}
\vfill
\item Code complete; awaiting higher-order auto-diff
\end{itemize}

\sld{Thermodynamic Sampler}
\begin{itemize}
\item Physically motivated alternative to ``simulated''
  \myemph{annealing and tempering} (not really simulated!)
\item Supplies external \myemph{heat bath}
\item Operates through \myemph{contact manifold}
\item System relaxes more naturally between energy levels
\vfill
\item Prototypes evaluated
\item Betancourt paper on {\slshape arXiv} (for geometers)
\end{itemize}

\sld{Marginal Maximum Likelihood}
\begin{itemize}
\item Marginalize out hierarchical parameters
\item Gradient-based nested optimization algorithm
\item Errors estimated via curvature as in MLE
\vfill
\item Design complete; awaiting parameter tagging
\end{itemize}

\sld{``Black Box'' VB and EP}
\begin{itemize}
\item \myemph{Black box} so can run any model
\\
{\footnotesize (Laplace or other approximations)}
\item Stochastic, data-streaming \myemph{variational Bayes} (VB)
\item Data-parallel \myemph{expectation propagation} (EP)
\\ {\footnotesize (cavity distributions provide general shard combination)}
\item Both use point estimates of parametric approximation to posterior
\item Optimize parameters to minimize KL divergence
\\
{\footnotesize (VB, EP measure divergence in opposite directions)}
\vfill
\item Prototype stage
\end{itemize}

\mypart{}{The End}

\sld{Stan's Namesake}
\begin{itemize}
\item Stanislaw Ulam (1909--1984)
\item Co-inventor of Monte Carlo method (and hydrogen bomb)
\item[]
\begin{center}
\includegraphics[width=0.25\textwidth]{img/ulam-fermiac.jpg}
\end{center}
{\small
\item Ulam holding the Fermiac, Enrico Fermi's physical Monte Carlo simulator
for random neutron diffusion}
\end{itemize}

\mypart{Appendix I}{Bayesian Data Analysis}

\sld{Bayesian Data Analysis}
\begin{itemize}
\item ``By {Bayesian data analysis}, we mean {practical methods}
for making {inferences} from {data} using {probability models}
for quantities we {observe} and about which we {wish to learn}.''
%
\item ``The essential characteristic of Bayesian methods is
their \myemph{explict use of probability for quantifying uncertainty}
in inferences based on statistical analysis.''
\end{itemize}
%
\vfill\hfill{\footnotesize Gelman et al., {\slshape Bayesian Data Analysis},
  3rd edition, 2013}

\sld{Bayesian Mechanics}
%
\begin{enumerate}
\item Set up full probability model 
\vspace*{-4pt}
\begin{itemize}
\item for all observable \& unobservable quantities
\item consistent w. problem knowledge \& data collection
\end{itemize}
%
\item Condition on observed data
\vspace*{-4pt}
\begin{itemize}
\item caclulate posterior probability of unobserved quantities
  conditional on observed quantities
\end{itemize}
%
\item Evaluate 
\vspace*{-4pt}
\begin{itemize}
\item model fit 
\item implications of posterior
\end{itemize}
\end{enumerate}

\vfill\hfill {\footnotesize {\slshape Ibid.}}

\sld{Basic Quantities}
%
\begin{itemize}
\item Basic Quantities
\vspace*{-4pt}
\begin{itemize}
\item $y$: \ observed data
\item $\tilde{y}$: \ unknown, potentially observable quantities
\item $\theta$: \ parameters (and other unobserved quantities)
\item $x$: \ constants, predictors for conditional models
\end{itemize}
\item Random models for things that could've been otherwise
\begin{itemize}
\item Everyone: Model data $y$ as random
\item Bayesians:  Model parameters $\theta$ as random
\end{itemize}
\end{itemize}

\sld{Distribution Naming Conventions}

\begin{itemize}
\item \myemph{Joint}: \ $p(y,\theta)$
\item \myemph{Sampling / Likelihood}: \ $p(y|\theta)$
\item \myemph{Prior}: \ $p(\theta)$
\item \myemph{Posterior}: \ $p(\theta|y)$
\item \myemph{Data Marginal}: \ $p(y)$
\item \myemph{Posterior Predictive}: \ $p(\tilde{y}|y)$
\end{itemize}

\noindent
\spc
{\footnotesize $y$ modeled data, \, $\theta$ parameters, \, $\tilde{y}$ predictions,}
\\[4pt]
\spc
{\footnotesize implicit: $x, \tilde{x}$ unmodeled data (for $y$, $\tilde{y}$), \ size constants}

\sld{Bayes's Rule for the Posterior}
%
\begin{itemize}
\item Suppose the data $y$ is fixed (i.e., observed).  Then
%
\vspace*{2pt}
\begin{eqnarray*}
p(\theta|y) 
 \hspace*{8pt} = \hspace*{8pt} \frac{p(y,\theta)}{p(y)}
& = & \frac{p(y|\theta) \, p(\theta)}{p(y)}
\\[6pt]
& = & \frac{p(y|\theta) \, p(\theta)}{\int p(y,\theta) \ d\theta}
\\[6pt]
& = & \frac{p(y|\theta) \, p(\theta)}{\int p(y|\theta) \, p(\theta) \ d\theta}
\\[6pt]
& \propto & p(y|\theta) \, p(\theta) \ \ = \ \ p(y,\theta)
\end{eqnarray*}
\item Posterior proportional to likelihood times prior (i.e., joint)
\end{itemize}

\sld{Monte Carlo Methods}
\begin{itemize}
\item For integrals that are impossible to solve analytically
\item But for which sampling and evaluation is tractable
\item Compute plug-in estimates of statistics based on
randomly generated variates (e.g., means, variances,
quantiles/intervals, comparisons)
\item Accuracy with $M$ (independent) samples proportional to
\[
\frac{1}{\sqrt{M}}
\]
e.g., 100 times more samples per decimal place!
\end{itemize}
\vfill\hfill
{\small (Metropolis and Ulam 1949)}

\sld{Monte Carlo Example}
\begin{itemize}
\item Posterior expectation of $\theta$:
\[
\mathbb{E}[\theta|y] = \int \theta \ p(\theta|y) \ d\theta.
\]
\item Bayesian estimate minimizing expected square error: 
\[
\hat{\theta} 
= \arg\min_{\theta'}
\mathbb{E}[(\theta - \theta')^2|y]
= \mathbb{E}[\theta|y] 
\]
\item Generate samples $\theta^{(1)}, \theta^{(2)}, \ldots,
 \theta^{(M)}$ drawn from $p(\theta|y)$
\item Monte Carlo Estimator plugs in average for expectation:
\[
\mathbb{E}[\theta|y] \approx \frac{1}{M} \sum_{m=1}^M \theta^{(m)}
\]
\end{itemize}

\sld{Monte Carlo Example II}
\begin{itemize}
\item Bayesian alternative to frequentist hypothesis testing
\item Use probability to summarize results
\item Bayesian comparison: probability $\theta_1 > \theta_2$ given
  data $y$?
\begin{eqnarray*}
\mbox{Pr}[\theta_1 > \theta_2|y] 
& = & 
\int \int 
\mathbb{I}(\theta_1 > \theta_2) \ p(\theta_1|y) \ p(\theta_2|y)  
\ d\theta_1 \ d\theta_2
\\
& \approx & 
\frac{1}{M} \sum_{m=1}^M \mathbb{I}(\theta_1^{(m)} > \theta_2^{(m)})
\end{eqnarray*}
%
\item (Bayesian hierarchical model ``adjusts'' for multiple comparisons)
\end{itemize}

\sld{Markov Chain Monte Carlo}
\begin{itemize}
\item When sampling independently from $p(\theta|y)$ impossible
\item $\theta^{(m)}$ drawn via a Markov chain $p(\theta^{(m)}|y,\theta^{(m-1)})$
\item Require MCMC marginal $p(\theta^{(m)}|y)$ equal to true
  posterior marginal
\item Leads to auto-correlation in samples
  $\theta^{(1)},\ldots, \theta^{(m)}$
\item Effective sample size $N_{\mbox{\tiny eff}}$ divides out
  autocorrelation (must be estimated)
\item Estimation accuracy proportional to $1 / \sqrt{N_{\mbox{\tiny eff}}}$
\end{itemize}

\sld{Gibbs Sampling}
\begin{itemize}
\item Samples a parameter given data and other parameters
\item Requires conditional posterior $p(\theta_n|y,\theta_{-n})$
\item Conditional posterior easy in directed graphical model
\item Requires general unidimensional sampler for non-conjugacy

\begin{itemize}
\item JAGS uses slice sampler
\item BUGS uses adaptive rejection sampler
\end{itemize}
\item Conditional sampling and general unidimensional sampler 
can both lead to slow convergence and mixing
\end{itemize}
\vfill\hfill
{\small (Geman and Geman 1984)}

\sld{Metropolis-Hastings Sampling}
\begin{itemize}
\item Proposes new point by changing all parameters randomly
\item Computes accept probability of new point based
on ratio of new to old log probability (and proposal density)
\item Only requires evaluation of $p(\theta|y)$
\item Requires good proposal mechanism to be effective
\item Acceptance requires small changes in log probability
\item But small step sizes lead to random walks and slow convergence
  and mixing
\end{itemize}
\vfill\hfill
{\small (Metropolis et al. 1953; Hastings 1970)}

\end{document}

### 2014-06-26 src/docs/talks/paris-dauphine-2014/stan-overview.tex
final talk used at Dauphine

@@
 \\
-{\footnotesize hosted on GitHub; uses Eigen matrix lib, Boost, googletest
-  C++ lib}
+{\footnotesize hosted on GitHub; uses Eigen matrix lib, Boost C++ lib, googletest}
 \end{itemize}
@@
 \item separate installation, getting started, interface manuals 
-\\ {\footnotesize  RStan, PyStan, CmdStan}
+\\ {\footnotesize  (RStan, PyStan, CmdStan)}
 \item BUGS and JAGS examples (all 3 volumes), 
@@
 
+\sld{Scaling and Evaluation}
+\begin{itemize}
+\item Types of scaling
+\vspace*{-4pt}
+\begin{itemize}\small
+\item more data
+\item more parameters
+\item \myemph{more complex models}
+\end{itemize}
+\item for MCMC, measure (as function of scaling)
+\vspace*{-4pt}
+\begin{itemize}\small
+\item time to convergence
+\item time per effective sample at mixing
+\item memory usage
+\end{itemize}
+\end{itemize}
+
+\sld{Stan vs. BUGS / JAGS}
+\begin{itemize}
+\item JAGS is the former \myemph{state-of-the-art}, \myemph{black-box} MCMC method
+\\ {\footnotesize (Gibbs with slice sampling; chokes w.\ high posterior correlation)}
+\item Stan (2.0) vs. BUGS on scaling IRT model
+\vspace*{-4pt}
+\begin{itemize}\footnotesize
+\item Stan about \myemph{10 times faster}
+\item Stan memory and time scales linearly
+\item linearly scales with number of question answers
+\end{itemize}
+\vspace*{-3pt}
+\vfill
+\hspace*{-12pt}
+\includegraphics[height=0.6in]{img/irt_scale-eps-converted-to.pdf}
+\includegraphics[height=0.45in]{img/irt-memory-2.png}
+\item Stan can handle problems that choke BUGS and JAGS
+\end{itemize}
+
 \mypart{Part I}{Stan Front End}
 
-\sld{Interface Languages}
+\sld{Platforms and Interfaces}
 
@@
 \item {\slshape content}: declare and define generated quantity
-  variables
+  variables; \\
+includes pseudo-random number generators
 \\
@@
 \begin{itemize}\small
-\item {\slshape content}: declare and define functions
+\item {\slshape content}: declare and define general (recursive) functions
+\\
+{\small (use them elsewhere in program)}
 \item {\slshape execute}: compile with model
 \end{itemize}
-\vspace*{12pt}
+\vspace*{6pt}
 \item Example
@@
 
-  real relative_difference(real x, real y) {
-    return 2 * fabs(x - y) / (fabs(x) + fabs(y));
+  real relative_difference(real u, real v) {
+    return 2 * fabs(u - v) / (fabs(u) + fabs(v));
   }
@@
 \item \myemph{Primitive}: {\tt\small int}, \ {\tt\small real}
-\item \myemph{Matrix}: {\tt\small matrix[M,N]}, \ {\tt\small vector[N]}, \ {\tt\small row\_vector[M]}
+\item \myemph{Matrix}: {\tt\small matrix[M,N]}, \ {\tt\small vector[M]}, \ {\tt\small row\_vector[N]}
 \item \myemph{Bounded}: primitive or matrix, with 
@@
 \begin{itemize}
-\item All built-in \myemph{C++ functions}
+\item All built-in \myemph{C++ functions and operators}
 \\
 {\footnotesize C math, TR1, C++11, including all trig, pow, and
-  special log1m, etc.}
+  special log1m, erf, erfc, fma, atan2, etc.}
 \item Extensive library of \myemph{statistical functions}
@@
 {\footnotesize e.g., softmax,
-log gamma and digamma functions, beta functions, Bessel functions}
+log gamma and digamma functions, beta functions, Bessel functions of
+ and second kind, etc.}
 %
@@
 \item \myemph{Elementwise arithmetic}: vectorized operations
-\item \myemph{Solvers}: matrix left and right division, (log) determinant,
+\item \myemph{Solvers}: matrix division, (log) determinant,
 inverse 
@@
 \item \myemph{Compound Operations}: quadratic forms, variance scaling
-\vfill
 \item \myemph{Ordering, Slicing, Broadcasting}: sort, rank, block, rep
 \item \myemph{Reductions}: sum, product, norms
+\item \myemph{Specializations}: triangular, positive-definite, etc.
 \end{itemize}
@@
 \\
-{\footnotesize aka no-U-turn sampler; aka NUTS}
+{\footnotesize aka no-U-turn sampler (NUTS)}
 \end{itemize}
@@
 \begin{itemize}
-\item Generated quantities block for predictions, decisions,
-  and event probabilities
-\item Extractors for samples in RStan and PyStan
-\item Coda-like posterior summary
+\item Generated quantities block for \myemph{inference}
+\\ {\footnotesize (predictions, decisions, and event probabilities)}
+\item \myemph{Extractors} for samples in RStan and PyStan
+\item Coda-like \myemph{posterior summary}
 \vspace*{-4pt}
@@
 \end{itemize}
-\item Model comparison with WAIC
+\item Model comparison with \myemph{WAIC}
 \\
@@
 \begin{itemize}
-\item Posterior mode finding via BFGS optimization
+\item Posterior \myemph{mode finding} via BFGS optimization
 \\ {\footnotesize (uses model gradient, efficiently approximates Hessian)}
-\item Disables Jacobians adjusting for parameter transforms
+\item \myemph{Disables Jacobians} for parameter inverse transforms
 \item Models, data, initialization as in MCMC
 \vfill
-\item Near Future:
-\vspace*{-4pt}
-\begin{itemize}\small
-\item  Standard errors on unconstrained scale
+\item \myemph{Very Near Future}
+\vspace*{-4pt}
+\begin{itemize}\small
+\item  \myemph{Standard errors} on unconstrained scale
 \\
 {\footnotesize  (estimated using curvature of penalized log likelihood function}
-\item Standard errors on constrained scale)
+\item Standard errors \myemph{on constrained scale})
 \\
@@
 \begin{itemize}
-\item Stan can be used to explore statistical algorithms
-\item Models transformed to have support on $\mathbb{R}^n$
+\item Stan can be used to \myemph{explore algorithms}
+\item Models transformed to \myemph{unconstrained support} on $\mathbb{R}^n$
 \item Once a model is compiled, have
@@
 \begin{itemize}\small
-\item log probability, gradient, and Hessian
+\item \myemph{log probability, gradient, and Hessian}
 \item data I/O and parameter initialization
@@
 \end{itemize}
-\item Future:
-\vspace*{-4pt}
-\begin{itemize}\small
-\item second- and higher-order derivatives via auto diff
+\item {Very Near Future:}
+\vspace*{-4pt}
+\begin{itemize}\small
+\item second- and \myemph{higher-order derivatives} via auto-diff
 \end{itemize}
@@
 \item leapfrog integrator \myemph{generates proposal}
-\item \myemph{Metropolis accept} step ensures detailed balance
+\item \myemph{Metropolis accept} step ensures \myemph{detailed balance}
 \end{itemize}
@@
 \item need to regularize Hessian-based curvature estimate
-\\ {\footnotesize (Betancourt arXiv; SoftAbs metric)}
+\\ {\footnotesize (Betancourt {\slshape arXiv}; SoftAbs metric)}
 \end{itemize}
@@
 \\ {\footnotesize (rare: throw away second half if not end to end U-turn)}
-\item \myemph{Slice sample} points along last segment of path
+\item \myemph{Slice sample} points along last branch of tree
 \item \myemph{Generalized} to Riemannian-manifold HMC
@@
 \item Dynamic program propagates \myemph{chain rule} in reverse pass
-\item \myemph{Object-oriented} custom partial propagation
+\item Extensible w.\ \myemph{object-oriented} custom partial propagation
 \item Arena-based \myemph{memory management}
@@
 \item Code \myemph{partial derivatives} for basic operations
-\item Function evaluation propagates chain rule forward
-\item Nest reverse-mode in forward for higher-order
+\item Function evaluation propagates \myemph{chain rule} forward
+\item Nest reverse-mode in forward for \myemph{higher-order}
+\item \myemph{Jacobians}
+\vspace*{-4pt}
+\begin{itemize}\small
+\item Rerun propagation pass in reverse mode
+\item Rerun forward construction with forward mode
+\end{itemize}
 \vfill
@@
 \item Functions to differentiate coded as functors (or pointers)
-\\ {\footnotesize (use bind or lambda)}
+\\ {\footnotesize (enables dynamic C++ bind or lambda)}
 \end{itemize}
@@
 \\ {\footnotesize (Boost Variant)}
-\item C++ code compilation
-\item Dynamic linking for RStan, PyStan
+\item C++ code \myemph{compilation}
+\item \myemph{Dynamic linking} for RStan, PyStan
 \end{itemize}
@@
 }
-\item \myemph{expression templates} to broadcast and cache scalars
+\item recursive \myemph{expression templates} to broadcast and cache scalars,
+generalize containers (arrays, matrices, vectors)
 \item \myemph{traits} metaprogram to \myemph{drop constants} (e.g., $-\log
-  \sqrt{2 \pi}$) and calculate return types
+  \sqrt{2 \pi}$) and calculate intermediate and return types
 \end{itemize}
@@
 \item Inference straightforward with expectations
-\item \myemph{Difficult} for many users
+\vspace*{12pt}
+\item Too \myemph{difficult} for many of our users
+\\
+{\small (exploring encapsulation options)}
 \end{itemize}
@@
 \item In practice, how to declare? 
-\vspace*{-3pt}
-\begin{itemize}\small
-\item observed data as data variables
-\item missing data as parameters
+\begin{itemize}
+\item \myemph{observed} data as data variables
+\item \myemph{missing} data as parameters
 \item combine into single vector 
@@
 \item C++ prototype integrated for large PK/PD models
-\\ {\footnotesize (project with Novartis for clinical trial)}
+\vspace*{-4pt}
+\begin{itemize}\footnotesize
+\item 
+Project with Novartis: 
+  longitudinal clinical trial w.\ multiple drugs, dosings, placebo control,
+  hierarchical model of patient-level effects, meta-analysis
+\item
+Collaborators: Frederic Bois, Amy Racine, Sebastian Weber
+\end{itemize}
 \end{itemize}
@@
 \begin{itemize}
+\item \myemph{NUTS} generalized to RHMC
+\\ {\footnotesize (Betancourt {\slshape arXiv} paper)}
 \item \myemph{SoftAbs} metric
@@
 \vfill
-\item Prototypes evaluated
-\item Betancourt paper on {\slshape arXiv} (for geometers)
+\item Prototype complete
+\\ {\small (Betancourt paper on {\slshape arXiv}; for geometers)}
 \end{itemize}

### 2014-08-21 src/docs/talks/msr-nyc-2014/stan-current-future.tex
added draft of talk for MSR NYC 2014 talk

--- 2014-06-26 src/docs/talks/paris-dauphine-2014/stan-overview.tex
@@
 \\
-\spc\spc{\small Allen Riddell, \  Marco Inacio}
+\spc\spc{\small Allen Riddell, \  Marco Inacio, \ Jeff Arnold, \ }
+\\
+\spc\spc{\small Mitzi Morris \ }
+\vfill
 \vfill
@@
 \\
-\spc{\footnotesize Paris-Dauphine 2014}
+\spc{\footnotesize Microsoft Research NYC 2014}
 \hfill 
 {\footnotesize \tt mc-stan.org}
+
+\mypart{{\LARGE The Present:} \\ \spc\spc{\Huge Stan 2.4}}
 
@@
 \item codes log posterior (penalized likelihood)
+\item calculates gradients, Hessians, etc., using auto-diff
 \end{itemize}
@@
 \begin{itemize}\small
-\item MCMC for full Bayesian inference
-\item MLE for point estimation
-\end{itemize}
-\item Stan is \myemph{open source} \ {\footnotesize (BSD core C++, GPLv3 interfaces)}
-\\
-{\footnotesize hosted on GitHub; uses Eigen matrix lib, Boost C++ lib, googletest}
-\end{itemize}
+\item MCMC for full Bayesian inference (+ posterior analysis)
+\item MLE for point estimates (+ Hessian-based error estimates)
+\end{itemize}
+\end{itemize}
+
+\sld{Stan is Open}
+
+\begin{itemize}
+\item Stan is \myemph{open source}
+\vspace*{-4pt}
+\begin{itemize} 
+\small
+\item Stan Core C++, CmdStan interface: {\slshape new BSD}
+\item PyStan, RStan interfaces: {\slshape GPLv3}
+\item Dependencies: Eigen, Boost, (googletest)
+\end{itemize}
+\item GitHub \myemph{hosted publicly} {\small ({\tt stan-dev})}
+\vspace*{-4pt}
+\begin{itemize}
+\small
+\item issue tracking (bug reports, feature requests)
+\item pull requests, code review, continuous integration hooks
+\end{itemize}
+\item Google Groups \myemph{public mailing lists} {\small ({\tt stan-dev}, {\tt stan-users})}
+\end{itemize}
+
+\sld{Platforms and Interfaces}
+
+\begin{itemize}
+\item \myemph{Platforms}
+\\
+Linux, Mac OS X, Windows
+\item \myemph{C++ API}
+\\
+{\footnotesize portable, standards compliant (C++03 now, moving to C++11)}
+\item \myemph{Interfaces}
+\vspace*{-4pt}
+\begin{itemize}\footnotesize
+\item \myemph{CmdStan}: Command-line or shell interface (direct executable)
+\item \myemph{RStan}: R interface (Rcpp in memory)
+\item \myemph{PyStan}: Python interface (Cython in memory)
+\item \myemph{MStan$^*$}: MATLAB interface (lightweight external process)
+\item \myemph{JuliaStan$^*$}: Julia interface (lightweight external process)
+\end{itemize}
+{\footnotesize ${}^*$ User contributed}
+\end{itemize}
+\vfill
 
@@
 \begin{itemize}
-\item 750+ registrations for mailing list across the physical,
-biomedical, and social sciences, plus engineering, finance and marketing
-
-\item Subjects of Papers on Google Scholar in the sciences
-\\
+\item 830+ registrations for mailing list
+\item spanning physical, biomedical, and social sciences, 
+\\ plus engineering, education, finance, and marketing
+
+\item Papers citing Stan on {\slshape Google Scholar} in subjects
+\\[3pt]
 {\footnotesize clinical drug
@@
 ecology, collaborative filtering for recommender systems, climatology,
-educational testing, natural language processing}
-\end{itemize}
-
-\sld{Books and Model Sets}
+educational testing, natural language processing, econometrics}
+\end{itemize}
+
+\sld{Documentation and Examples}
 %
@@
 \item two other books in progress
+\item user-contributed examples in group
 \end{itemize}
@@
 \begin{itemize}
-\item Types of scaling
-\vspace*{-4pt}
-\begin{itemize}\small
-\item more data
-\item more parameters
-\item \myemph{more complex models}
-\end{itemize}
-\item for MCMC, measure (as function of scaling)
+\item Type of scaling \hfill (improvement$^{*}$)
+\vspace*{-4pt}
+\begin{itemize}\small
+\item more data \hfill (1--2 orders of magnitude)
+\item more parameters \hfill (1--2 orders of magnitude)
+\item \myemph{more complex models} \hfill ({\large \myemph{$\infty$}})
+\end{itemize}
+\item imporvement vs.\ ``state-of-the-art'' \myemph{black-box MCMC}
+measured by
 \vspace*{-4pt}
@@
 
-\sld{Stan vs. BUGS / JAGS}
-\begin{itemize}
-\item JAGS is the former \myemph{state-of-the-art}, \myemph{black-box} MCMC method
-\\ {\footnotesize (Gibbs with slice sampling; chokes w.\ high posterior correlation)}
-\item Stan (2.0) vs. BUGS on scaling IRT model
-\vspace*{-4pt}
-\begin{itemize}\footnotesize
-\item Stan about \myemph{10 times faster}
-\item Stan memory and time scales linearly
-\item linearly scales with number of question answers
-\end{itemize}
-\vspace*{-3pt}
-\vfill
-\hspace*{-12pt}
-\includegraphics[height=0.6in]{img/irt_scale-eps-converted-to.pdf}
-\includegraphics[height=0.45in]{img/irt-memory-2.png}
-\item Stan can handle problems that choke BUGS and JAGS
-\end{itemize}
-
 \mypart{Part I}{Stan Front End}
-
-\sld{Platforms and Interfaces}
-
-\begin{itemize}
-\item \myemph{Platforms}
-\\
-Linux, Mac OS X, Windows
-\item \myemph{C++ API}
-\\
-{\footnotesize portable, standards compliant (C++03 now, moving to C++11)}
-\item \myemph{Interfaces}
-\vspace*{-4pt}
-\begin{itemize}\footnotesize
-\item \myemph{CmdStan}: Command-line or shell interface (direct executable)
-\item \myemph{RStan}: R interface (Rcpp in memory)
-\item \myemph{PyStan}: Python interface (Cython in memory)
-\item \myemph{MStan$^*$}: MATLAB interface (lightweight external process)
-\item \myemph{JuliaStan$^*$}: Julia interface (lightweight external process)
-\end{itemize}
-{\footnotesize ${}^*$ User contributed}
-\end{itemize}
-\vfill
 
@@
 \item {\slshape content}: declare parameter types, sizes, and constraints
-\item {\slshape execute}: transform to constrained, Jacobian
+\item {\slshape execute}: transform to constrained, optional Jacobian
 \end{itemize}
@@
 
-\sld{User-Defined Functions (Stan 2.3)}
+\sld{User-Defined Functions}
 
@@
 \begin{itemize}
-\item Posterior \myemph{mode finding} via BFGS optimization
+\item Posterior \myemph{mode finding} via L-BFGS optimization
 \\ {\footnotesize (uses model gradient, efficiently approximates Hessian)}
@@
 \item Models, data, initialization as in MCMC
-\vfill
-\item \myemph{Very Near Future}
-\vspace*{-4pt}
-\begin{itemize}\small
-\item  \myemph{Standard errors} on unconstrained scale
-\\
-{\footnotesize  (estimated using curvature of penalized log likelihood function}
-\item Standard errors \myemph{on constrained scale})
-\\
-{\footnotesize  (sample unconstrained approximation and inverse transform)}
-\item L-BFGS optimizer
-\end{itemize}      
+\item Curvature (Hessian) used to estimate posterior covariance (or
+  standard errors on unconstrained scale)
 \end{itemize}
@@
 \item \myemph{Generalized} to Riemannian-manifold HMC
-\\ {\footnotesize (Betancourt, arXiv)}
+\\ {\footnotesize (Betancourt arXiv paper)}
 \end{itemize}
@@
 
-\mypart{Part III}{What's Next?}
+\mypart{{\LARGE The Future:} \\ \spc\spc{\Huge Stan 3 \& Beyond}}
+
+\sld{Stan 3: The Refactoring}
+\begin{itemize}
+\item Third time's the charm in software (re)design
+\item Goal of removing duplicated code in interfaces
+\item Minimizing dependencies between interfaces and core
+\item More modular and flexible
+\vspace*{-4pt}
+\begin{itemize}\small
+\item user-facing C++ API
+\item interface API
+\end{itemize}
+\vfill
+\item Design and some coding underway
+\end{itemize}
 
@@
   hierarchical model of patient-level effects, meta-analysis
-\item
-Collaborators: Frederic Bois, Amy Racine, Sebastian Weber
-\end{itemize}
-\end{itemize}
-
-\sld{L-BFGS Optimizer}
-\begin{itemize}
-\item Faster, lower memory, more stable than BFGS
-\item Plug-and-play in interfaces
-\vfill
-\item Code and doc complete; under testing
+\end{itemize}
+\vfill
+\item Generalized code complete, under testing
+\item {\footnotesize With: Frederic Bois, Amy Racine-Poon, Sebastian Weber}
+\end{itemize}
+
+\sld{MLE/MML Error Estimation}
+\begin{itemize}
+\item Use curvature to estimate error on unconstrained scale
+\item Transform back to constrained scale 
+\item May require sampling from Laplace approximation
+\vfill
+\item Design complete
 \end{itemize}
@@
 \begin{itemize}
+\item Supports posteriors with \myemph{position-dependent
+    curvature} with local \myemph{mass matrix estimation}
+\\
+  {\small (e.g., hierarchical models)}
 \item \myemph{NUTS} generalized to RHMC
-\\ {\footnotesize (Betancourt {\slshape arXiv} paper)}
 \item \myemph{SoftAbs} metric
@@
 \item \myemph{condition} by narrowing eigenvalue range
-\item Betancourt {\slshape arXiv}\ paper
-\end{itemize}
-\vfill
-\item Code complete; awaiting higher-order auto-diff
+\end{itemize}
+\vfill
+\item Code complete; awaiting full higher-order auto-diff
+\item {\small (Betancourt {\slshape arXiv} papers)}
 \end{itemize}
@@
 \begin{itemize}
+\item Enhances posterior mode-finding with \myemph{multiple modes}
 \item Physically motivated alternative to ``simulated''
@@
 \item Prototype complete
-\\ {\small (Betancourt paper on {\slshape arXiv}; for geometers)}
-\end{itemize}
-
-\sld{Marginal Maximum Likelihood}
-\begin{itemize}
-\item Marginalize out hierarchical parameters
+\item {\small (Betancourt {\slshape arXiv} paper)}
+\end{itemize}
+
+\sld{\Large Marginal Maximum Likelihood (MML)}
+\begin{itemize}
+\item Enable MLE-like estimate with \myemph{infinite likelihoods}
+\\ {\small (e.g., hierarchical models, mixture models / clustering)}
+\item Marginalize out lower-level parameters to estimate hierarchical
+\item Estimate lower-level parameters based on hierarchical estimates
 \item Gradient-based nested optimization algorithm
-\item Errors estimated via curvature as in MLE
+\item Errors / posterior variance estimated as in MLE
 \vfill
@@
 
-\sld{``Black Box'' VB and EP}
-\begin{itemize}
-\item \myemph{Black box} so can run any model
-\\
-{\footnotesize (Laplace or other approximations)}
-\item Stochastic, data-streaming \myemph{variational Bayes} (VB)
-\item Data-parallel \myemph{expectation propagation} (EP)
-\\ {\footnotesize (cavity distributions provide general shard combination)}
-\item Both use point estimates of parametric approximation to posterior
+\sld{Variational Bayes (VB)}
+\begin{itemize}
+\item \myemph{Black box} to \myemph{scale} arbitrary models
+\item Batch or stochastic (data streaming)
+\item Compute expectations via approximations (e.g., Laplace)
+\item Point estimate parametric approximations to posterior
 \item Optimize parameters to minimize KL divergence
-\\
-{\footnotesize (VB, EP measure divergence in opposite directions)}
 \vfill
 \item Prototype stage 
+\item {\footnotesize With: Dave Blei, Alp Kucukelbir, and Rajesh Ranganath}
+\end{itemize}
+
+\sld{Expectation Propagation (EP)}
+\begin{itemize}
+\item \myemph{Black box} to \myemph{scale} arbitrary models
+\item Data-parallelization
+\vspace*{-4pt}
+\begin{itemize}\small
+\item cavity distributions guide shard combination
+\end{itemize}
+\item Point estimate parametric approximations to posterior
+\item Optimize parameters to minimize KL divergence
+\vfill
+\item Prototype stage
+\item {\footnotesize With: Aki Vehtari, Nicolas Chopin, Christian Robert,
+  John Cunningham}
 \end{itemize}

### 2014-08-22 src/docs/talks/msr-nyc-2014/stan-current-future.tex
version of talk for presentation

@@
 
-\mypart{{\LARGE The Present:} \\ \spc\spc{\Huge Stan 2.4}}
+\mypart{Current:}{Stan 2.4}
 
@@
 \item declares variables
-\item codes log posterior (penalized likelihood)
-\item calculates gradients, Hessians, etc., using auto-diff
+\item codes log posterior (or penalized likelihood)
 \end{itemize}
@@
 \begin{itemize}\small
-\item MCMC for full Bayesian inference (+ posterior analysis)
-\item MLE for point estimates (+ Hessian-based error estimates)
-\end{itemize}
-\end{itemize}
-
-\sld{Stan is Open}
-
-\begin{itemize}
-\item Stan is \myemph{open source}
-\vspace*{-4pt}
-\begin{itemize} 
-\small
-\item Stan Core C++, CmdStan interface: {\slshape new BSD}
-\item PyStan, RStan interfaces: {\slshape GPLv3}
-\item Dependencies: Eigen, Boost, (googletest)
-\end{itemize}
-\item GitHub \myemph{hosted publicly} {\small ({\tt stan-dev})}
-\vspace*{-4pt}
-\begin{itemize}
-\small
-\item issue tracking (bug reports, feature requests)
-\item pull requests, code review, continuous integration hooks
-\end{itemize}
-\item Google Groups \myemph{public mailing lists} {\small ({\tt stan-dev}, {\tt stan-users})}
-\end{itemize}
-
-\sld{Platforms and Interfaces}
-
-\begin{itemize}
-\item \myemph{Platforms}
-\\
-Linux, Mac OS X, Windows
-\item \myemph{C++ API}
-\\
-{\footnotesize portable, standards compliant (C++03 now, moving to C++11)}
-\item \myemph{Interfaces}
-\vspace*{-4pt}
-\begin{itemize}\footnotesize
-\item \myemph{CmdStan}: Command-line or shell interface (direct executable)
-\item \myemph{RStan}: R interface (Rcpp in memory)
-\item \myemph{PyStan}: Python interface (Cython in memory)
-\item \myemph{MStan$^*$}: MATLAB interface (lightweight external process)
-\item \myemph{JuliaStan$^*$}: Julia interface (lightweight external process)
-\end{itemize}
-{\footnotesize ${}^*$ User contributed}
-\end{itemize}
-\vfill
-
-\sld{Who's Using Stan?}
-\begin{itemize}
-\item 830+ registrations for mailing list
-\item spanning physical, biomedical, and social sciences, 
-\\ plus engineering, education, finance, and marketing
-
-\item Papers citing Stan on {\slshape Google Scholar} in subjects
-\\[3pt]
-{\footnotesize clinical drug
-trials, general computational statistics, entomology, opthalmology,
-neurology, sociology and population dynamics, genomics, agriculture,
-psycholinguistics, molecular biology, population dynamics, materials
-engineering, botany, astrophysics, oceanography, election prediction,
-fisheries, cancer biology, public health and epidemiology, population
-ecology, collaborative filtering for recommender systems, climatology,
-educational testing, natural language processing, econometrics}
-\end{itemize}
-
-\sld{Documentation and Examples}
-%
-\begin{itemize}
-\item 400+ page core language tutorial manual
-\item separate installation, getting started, interface manuals 
-\\ {\footnotesize  (RStan, PyStan, CmdStan)}
-\item BUGS and JAGS examples (all 3 volumes), 
-\item Gelman and Hill, {\slshape Data Analysis Using Regression and
-    Multilevel/Hierarchical Models}
-\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
-\item two other books in progress
-\item user-contributed examples in group
-\end{itemize}
-
-\sld{Scaling and Evaluation}
-\begin{itemize}
-\item Type of scaling \hfill (improvement$^{*}$)
-\vspace*{-4pt}
-\begin{itemize}\small
-\item more data \hfill (1--2 orders of magnitude)
-\item more parameters \hfill (1--2 orders of magnitude)
-\item \myemph{more complex models} \hfill ({\large \myemph{$\infty$}})
-\end{itemize}
-\item imporvement vs.\ ``state-of-the-art'' \myemph{black-box MCMC}
-measured by
-\vspace*{-4pt}
-\begin{itemize}\small
-\item time to convergence
-\item time per effective sample at mixing
-\item memory usage
-\end{itemize}
-\end{itemize}
-
-\mypart{Part I}{Stan Front End}
+\item Sampling for full Bayesian inference
+\item Optimization + curvature for maximum likelihood estimates
+\end{itemize}
+\end{itemize}
 
@@
 
+\sld{Stan is Open}
+
+\begin{itemize}
+\item Stan is \myemph{open source}
+\vspace*{-4pt}
+\begin{itemize} 
+\small
+\item Stan Core C++, CmdStan interface: {\slshape new BSD}
+\item PyStan, RStan interfaces: {\slshape GPLv3}
+\item Dependencies: Eigen, Boost, (googletest)
+\end{itemize}
+\item GitHub \myemph{hosted publicly} {\small ({\tt stan-dev})}
+\vspace*{-4pt}
+\begin{itemize}
+\small
+\item issue tracking (bug reports, feature requests)
+\item pull requests, code review, continuous integration hooks
+\end{itemize}
+\item Google Groups \myemph{public mailing lists} {\small ({\tt stan-dev}, {\tt stan-users})}
+\end{itemize}
+
+\sld{Platforms and Interfaces}
+
+\begin{itemize}
+\item \myemph{Platforms}
+\\
+Linux, Mac OS X, Windows
+\item \myemph{C++ API}
+\\
+{\footnotesize portable, standards compliant (C++03 now, moving to C++11)}
+\item \myemph{Interfaces}
+\vspace*{-4pt}
+\begin{itemize}\footnotesize
+\item \myemph{CmdStan}: Command-line or shell interface (direct executable)
+\item \myemph{RStan}: R interface (Rcpp in memory)
+\item \myemph{PyStan}: Python interface (Cython in memory)
+\item \myemph{MStan$^*$}: MATLAB interface (lightweight external process)
+\item \myemph{JuliaStan$^*$}: Julia interface (lightweight external process)
+\end{itemize}
+{\footnotesize ${}^*$ User contributed}
+\end{itemize}
+\vfill
+
+\sld{Who's Using Stan?}
+\begin{itemize}
+\item 830+ user mailing list; 100+ citations
+\vspace*{-4pt}
+\begin{itemize}\small
+\item physical, biomedical, and social sciences
+\item plus engineering, education, finance, and marketing
+\end{itemize}
+\item Application areas
+  \\[3pt]
+  {\footnotesize clinical drug trials, general computational
+    statistics, entomology, opthalmology, neurology, sociology and
+    population dynamics, genomics, agriculture, psycholinguistics,
+    molecular biology, population dynamics, materials engineering,
+    botany, astrophysics, oceanography, election prediction,
+    fisheries, cancer biology, public health and epidemiology,
+    population ecology, collaborative filtering, climatology,
+    educational testing, natural language processing, econometrics}
+\end{itemize}
+
+\sld{Documentation and Examples}
+%
+\begin{itemize}
+\item Documentation
+\vspace*{-4pt}
+\begin{itemize}\footnotesize
+\item 400+ page user's guide and reference
+\item For each interface: installation, getting started, reference
+\end{itemize}
+\item Examples
+\vspace*{-4pt}
+\begin{itemize}\footnotesize
+\item BUGS and JAGS examples (all 3 volumes), 
+\item Gelman and Hill, {\slshape Data Analysis Using Regression and \\
+    Multilevel/Hierarchical Models}
+\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
+\item two books in progress
+\item user-contributed examples in group
+\end{itemize}
+\end{itemize}
+
+\sld{Scaling and Evaluation}
+\begin{itemize}
+\item Type of scaling
+\vspace*{-4pt}
+\begin{itemize}\small
+\item more data (e.g., observations)
+\item more parameters (e.g., regression coefficients)
+\item \myemph{more complex models} (e.g., multilevel priors)
+\end{itemize}
+\item Metrics
+\vspace*{-4pt}
+\begin{itemize}\small
+\item time to convergence
+\item time per effective sample after convergence
+\item memory usage
+\vfill
+\item 0 to {\large $\infty$} orders of magnitude improvement on
+  black-box state-of-the-art (e.g., BUGS, JAGS)
+\item more improvement with more complex models
+\end{itemize}
+\end{itemize}
+
 \sld{Basic Program Blocks}
@@
 \begin{itemize}\small
-\item {\slshape content}: statements definining posterior density
+\item {\slshape content}: definine posterior density (up
+  to constant)
 \item {\slshape execute}: execute statements
@@
 
+\sld{Arithmetic and Matrix Operators}
+\vfill
+\noindent\spc
+{\footnotesize
+\begin{tabular}{c|ccl|l}
+{\it Op.} & {\it Prec.} & {\it Assoc.} & {\it
+  Placement} & {\it Description}
+\\ \hline \hline
+
+\code{+} & 5 & left & binary infix & addition
+\\
+\code{-} & 5 & left & binary infix & subtraction
+\\ \hline
+\code{*} & 4 & left & binary infix & multiplication
+\\
+\code{/} & 4 & left & binary infix & (right) division
+\\ \hline
+\Verb|\| & 3 & left & binary infix & left division
+\\ \hline
+\code{.*} & 2 & left & binary infix & elementwise multiplication
+\\
+\code{./} & 2 & left & binary infix & elementwise division
+\\ \hline
+\code{!} & 1 & n/a & unary prefix & logical negation
+\\
+\code{-} & 1 & n/a & unary prefix & negation
+\\ 
+\code{+} & 1 & n/a & unary prefix & promotion (no-op in Stan)
+\\ \hline
+\Verb|^| & 2 & right & binary infix & exponentiation
+\\ \hline
+\code{'} & 0 & n/a & unary postfix & transposition
+\\ \hline \hline
+\code{()} & 0 & n/a & prefix, wrap & function application
+\\
+\code{[]} & 0 & left & prefix, wrap & array, matrix indexing
+\end{tabular}
+}
+\vfill
+
 \sld{Logical Operators}
@@
 
-\sld{Arithmetic and Matrix Operators}
-\vfill
-\noindent\spc
-{\footnotesize
-\begin{tabular}{c|ccl|l}
-{\it Op.} & {\it Prec.} & {\it Assoc.} & {\it
-  Placement} & {\it Description}
-\\ \hline \hline
-
-\code{+} & 5 & left & binary infix & addition
-\\
-\code{-} & 5 & left & binary infix & subtraction
-\\ \hline
-\code{*} & 4 & left & binary infix & multiplication
-\\
-\code{/} & 4 & left & binary infix & (right) division
-\\ \hline
-\Verb|\| & 3 & left & binary infix & left division
-\\ \hline
-\code{.*} & 2 & left & binary infix & elementwise multiplication
-\\
-\code{./} & 2 & left & binary infix & elementwise division
-\\ \hline
-\code{!} & 1 & n/a & unary prefix & logical negation
-\\
-\code{-} & 1 & n/a & unary prefix & negation
-\\ 
-\code{+} & 1 & n/a & unary prefix & promotion (no-op in Stan)
-\\ \hline
-\Verb|^| & 2 & right & binary infix & exponentiation
-\\ \hline
-\code{'} & 0 & n/a & unary postfix & transposition
-\\ \hline \hline
-\code{()} & 0 & n/a & prefix, wrap & function application
-\\
-\code{[]} & 0 & left & prefix, wrap & array, matrix indexing
-\end{tabular}
-}
-\vfill
-
 \sld{Built-in Math Functions}
@@
 log gamma and digamma functions, beta functions, Bessel functions of
- and second kind, etc.}
+first and second kind, etc.}
 %
@@
 \begin{itemize}
-\item \myemph{Basic arithmetic}: all arithmetic operators
-\item \myemph{Elementwise arithmetic}: vectorized operations
+\item \myemph{Type inference}, {\small e.g., row vector times vector is scalar}
+\item \myemph{Basic and elementwise arithmetic}: all ops
 \item \myemph{Solvers}: matrix division, (log) determinant,
@@
 Cholesky factorization, singular value decomposition
-\item \myemph{Compound Operations}: quadratic forms, variance scaling
 \item \myemph{Ordering, Slicing, Broadcasting}: sort, rank, block, rep
 \item \myemph{Reductions}: sum, product, norms
+\item \myemph{Compound Operations}: quadratic forms, variance scaling
 \item \myemph{Specializations}: triangular, positive-definite, etc.
@@
 {\footnotesize degrees of freedom controls 
-shrinkage to (expansion from) unit matrix}
+shrinkage to (expansion from) unit matrix; \\
+independently scale by parameter}
 \end{itemize}
@@
 \sld{Statements}
-
+\vspace*{-2pt}
 \begin{itemize}
@@
    (allows local variables)}
-\item \myemph{Print}: \ {\footnotesize print("theta=",theta);}
+\item \myemph{Print}: \ {\footnotesize \code{print("theta=",theta);}}
+\item \myemph{Throw}: \ {\footnotesize \code{raise\_exception("x must be
+    positive; found x=", x);}}
 \end{itemize}
@@
 \end{itemize}
-\item Model comparison with \myemph{WAIC}
+\item Model comparison with \myemph{WAIC} or cross-validation
 \\
@@
 
-\mypart{Part II}{Under the Hood}
+\mypart{Future:}{Stan 3 \& Beyond}
+
+\sld{Stan 3: The Refactoring}
+\begin{itemize}
+\item Third time's the charm in software (re)design
+\item Goal of removing duplicated code in interfaces
+\item Minimizing dependencies between interfaces and core
+\item More modular and flexible
+\vspace*{-4pt}
+\begin{itemize}\small
+\item user-facing C++ API
+\item interface API
+\end{itemize}
+\vfill
+\item Design and some coding underway
+\end{itemize}
+
+\sld{Differential Equation Solver}
+\begin{itemize}
+\item Auto-diff solutions w.r.t.\ parameters \& initial states
+\item Integrate coupled system for solution with partials
+\item Auto-diff coupled Jacobian for stiff systems
+\vfill
+\item C++ prototype integrated for large PK/PD models
+\vspace*{-4pt}
+\begin{itemize}\footnotesize
+\item 
+Project with Novartis: 
+  longitudinal clinical trial w.\ multiple drugs, dosings, placebo control,
+  hierarchical model of patient-level effects, meta-analysis
+\end{itemize}
+\vfill
+\item Generalized code complete, under testing
+\item {\footnotesize With: Frederic Bois, Amy Racine-Poon, Sebastian Weber}
+\end{itemize}
+
+\sld{Higher-Order Auto-diff}
+\begin{itemize}
+\item Finish higher-order auto-diff for probability functions
+\item May punt some cumulative distribution functions
+\\
+{\footnotesize (Black art iterative algorithms required)}
+\vfill
+\item Code complete; under testing
+\end{itemize}
+
+\sld{Riemannian Manifold HMC}
+\begin{itemize}
+\item Supports posteriors with \myemph{position-dependent
+    curvature} with local \myemph{mass matrix estimation}
+\\
+  {\small (e.g., hierarchical models)}
+\item \myemph{NUTS} generalized to RHMC
+\item \myemph{SoftAbs} metric
+\vspace*{-4pt}
+\begin{itemize}\footnotesize
+\item Eigendecompose Hessian
+\item \myemph{positive definite} with positive eigenvalues
+\item \myemph{condition} by narrowing eigenvalue range
+\end{itemize}
+\vfill
+\item Code complete; awaiting full higher-order auto-diff
+\item {\small (Betancourt {\slshape arXiv} papers)}
+\end{itemize}
+
+\sld{Thermodynamic Sampler}
+\begin{itemize}
+\item Enhances posterior mode-finding with \myemph{multiple modes}
+\item Physically motivated alternative to ``simulated''
+  \myemph{annealing and tempering} (not really simulated!)
+\item Supplies external \myemph{heat bath}
+\item Operates through \myemph{contact manifold}
+\item System relaxes more naturally between energy levels
+\vfill
+\item Prototype complete
+\item {\small (Betancourt {\slshape arXiv} paper)}
+\end{itemize}
+
+\sld{Ensemble MCMC}
+\begin{itemize}
+\item Markov chain is ensemble of parameter vectors
+\item Interpolate/extrapolate in ensemble for Metropolis proposal
+\begin{itemize}\small
+\item Walk \& stretch moves (Goodman \& Weare)
+\item Differential evolution (ter Braak)
+\end{itemize}
+\item Highly parallelizable
+\item Derivative free
+\vfill
+\item Code complete; awaiting integration
+\end{itemize}
+
+\sld{\Large Marginal Maximum Likelihood (MML)}
+\begin{itemize}
+\item Enable MLE-like estimate with \myemph{infinite likelihoods}
+\\ {\small (e.g., hierarchical models, mixture models / clustering)}
+\item Marginalize out lower-level parameters to estimate hierarchical
+\item Estimate lower-level parameters based on hierarchical estimates
+\item Gradient-based nested optimization algorithm
+\item Errors / posterior variance estimated as in MLE
+\vfill
+\item Design complete; awaiting parameter tagging
+\end{itemize}
+
+\sld{MLE \& MML Errors}
+\begin{itemize}
+\item Simple posterior approximation
+\item Standard errors for estimators
+\item Approximate posterior covariance with multivariate normal
+\item Estimate with Laplace approximation (based on curvature)
+\item Sample and transform to constrained scale
+\vfill
+\item Design complete
+\end{itemize}
+
+\sld{Variational Bayes (VB)}
+\begin{itemize}
+\item \myemph{Black box} to \myemph{scale} arbitrary models
+\item Batch or stochastic (data streaming)
+\item Compute expectations via approximations (e.g., Laplace)
+\item Point estimate parametric approximations to posterior
+\item Optimize parameters to minimize KL divergence
+\vfill
+\item Prototype stage 
+\item {\footnotesize With: Dave Blei, Alp Kucukelbir, and Rajesh Ranganath}
+\end{itemize}
+
+\sld{Expectation Propagation (EP)}
+\begin{itemize}
+\item \myemph{Black box} to \myemph{scale} arbitrary models
+\item Data-parallelization
+\vspace*{-4pt}
+\begin{itemize}\small
+\item cavity distributions guide shard combination
+\end{itemize}
+\item Point estimate parametric approximations to posterior
+\item Optimize parameters to minimize KL divergence
+\vfill
+\item Prototype stage
+\item {\footnotesize With: Aki Vehtari, Nicolas Chopin, Christian Robert,
+  John Cunningham}
+\end{itemize}
+
+\mypart{}{The End}
+
+\sld{Stan's Namesake}
+\begin{itemize}
+\item Stanislaw Ulam (1909--1984)
+\item Co-inventor of Monte Carlo method (and hydrogen bomb)
+\item[]
+\begin{center}
+\includegraphics[width=0.25\textwidth]{img/ulam-fermiac.jpg}
+\end{center}
+{\small
+\item Ulam holding the Fermiac, Enrico Fermi's physical Monte Carlo simulator
+for random neutron diffusion}
+\end{itemize}
+
+\mypart{Appendix I}{Under the Hood}
 
@@
 
-\mypart{{\LARGE The Future:} \\ \spc\spc{\Huge Stan 3 \& Beyond}}
-
-\sld{Stan 3: The Refactoring}
-\begin{itemize}
-\item Third time's the charm in software (re)design
-\item Goal of removing duplicated code in interfaces
-\item Minimizing dependencies between interfaces and core
-\item More modular and flexible
-\vspace*{-4pt}
-\begin{itemize}\small
-\item user-facing C++ API
-\item interface API
-\end{itemize}
-\vfill
-\item Design and some coding underway
-\end{itemize}
-
-\sld{Differential Equation Solver}
-\begin{itemize}
-\item Auto-diff solutions w.r.t.\ parameters
-\item Integrate coupled system for solution with partials
-\item Auto-diff coupled Jacobian for stiff systems
-\vfill
-\item C++ prototype integrated for large PK/PD models
-\vspace*{-4pt}
-\begin{itemize}\footnotesize
-\item 
-Project with Novartis: 
-  longitudinal clinical trial w.\ multiple drugs, dosings, placebo control,
-  hierarchical model of patient-level effects, meta-analysis
-\end{itemize}
-\vfill
-\item Generalized code complete, under testing
-\item {\footnotesize With: Frederic Bois, Amy Racine-Poon, Sebastian Weber}
-\end{itemize}
-
-\sld{MLE/MML Error Estimation}
-\begin{itemize}
-\item Use curvature to estimate error on unconstrained scale
-\item Transform back to constrained scale 
-\item May require sampling from Laplace approximation
-\vfill
-\item Design complete
-\end{itemize}
-
-\sld{Higher-Order Auto-diff}
-\begin{itemize}
-\item Finish higher-order auto-diff for probability functions
-\item May punt some cumulative distribution functions
-\\
-{\footnotesize (Black art iterative algorithms required)}
-\vfill
-\item Code complete; under testing
-\end{itemize}
-
-\sld{Riemannian Manifold HMC}
-\begin{itemize}
-\item Supports posteriors with \myemph{position-dependent
-    curvature} with local \myemph{mass matrix estimation}
-\\
-  {\small (e.g., hierarchical models)}
-\item \myemph{NUTS} generalized to RHMC
-\item \myemph{SoftAbs} metric
-\vspace*{-4pt}
-\begin{itemize}\footnotesize
-\item Eigendecompose Hessian
-\item \myemph{positive definite} with positive eigenvalues
-\item \myemph{condition} by narrowing eigenvalue range
-\end{itemize}
-\vfill
-\item Code complete; awaiting full higher-order auto-diff
-\item {\small (Betancourt {\slshape arXiv} papers)}
-\end{itemize}
-
-\sld{Thermodynamic Sampler}
-\begin{itemize}
-\item Enhances posterior mode-finding with \myemph{multiple modes}
-\item Physically motivated alternative to ``simulated''
-  \myemph{annealing and tempering} (not really simulated!)
-\item Supplies external \myemph{heat bath}
-\item Operates through \myemph{contact manifold}
-\item System relaxes more naturally between energy levels
-\vfill
-\item Prototype complete
-\item {\small (Betancourt {\slshape arXiv} paper)}
-\end{itemize}
-
-\sld{\Large Marginal Maximum Likelihood (MML)}
-\begin{itemize}
-\item Enable MLE-like estimate with \myemph{infinite likelihoods}
-\\ {\small (e.g., hierarchical models, mixture models / clustering)}
-\item Marginalize out lower-level parameters to estimate hierarchical
-\item Estimate lower-level parameters based on hierarchical estimates
-\item Gradient-based nested optimization algorithm
-\item Errors / posterior variance estimated as in MLE
-\vfill
-\item Design complete; awaiting parameter tagging
-\end{itemize}
-
-\sld{Variational Bayes (VB)}
-\begin{itemize}
-\item \myemph{Black box} to \myemph{scale} arbitrary models
-\item Batch or stochastic (data streaming)
-\item Compute expectations via approximations (e.g., Laplace)
-\item Point estimate parametric approximations to posterior
-\item Optimize parameters to minimize KL divergence
-\vfill
-\item Prototype stage 
-\item {\footnotesize With: Dave Blei, Alp Kucukelbir, and Rajesh Ranganath}
-\end{itemize}
-
-\sld{Expectation Propagation (EP)}
-\begin{itemize}
-\item \myemph{Black box} to \myemph{scale} arbitrary models
-\item Data-parallelization
-\vspace*{-4pt}
-\begin{itemize}\small
-\item cavity distributions guide shard combination
-\end{itemize}
-\item Point estimate parametric approximations to posterior
-\item Optimize parameters to minimize KL divergence
-\vfill
-\item Prototype stage
-\item {\footnotesize With: Aki Vehtari, Nicolas Chopin, Christian Robert,
-  John Cunningham}
-\end{itemize}
-
-\mypart{}{The End}
-
-\sld{Stan's Namesake}
-\begin{itemize}
-\item Stanislaw Ulam (1909--1984)
-\item Co-inventor of Monte Carlo method (and hydrogen bomb)
-\item[]
-\begin{center}
-\includegraphics[width=0.25\textwidth]{img/ulam-fermiac.jpg}
-\end{center}
-{\small
-\item Ulam holding the Fermiac, Enrico Fermi's physical Monte Carlo simulator
-for random neutron diffusion}
-\end{itemize}
-
-\mypart{Appendix I}{Bayesian Data Analysis}
+\mypart{Appendix II}{Bayesian Data Analysis}
 

### 2014-10-29 src/docs/talks/paris-dauphine-2014/stan-overview.tex
updated overview talk slides

@@
 \linespread{1.02}
-%\usepackage{helvet}
+% \usepackage{helvet}
 
@@
 \noindent
-\spc\spc{\slshape\footnotesize Core Development Team:}
+\spc\spc{\slshape\small Core Development Team:}
 \\[2pt]
@@
 \\
-\spc\spc{\small Allen Riddell, \  Marco Inacio}
+\spc\spc{\small Allen Riddell, \  Marco Inacio, \ Jeffrey Arnold, \ }
+\\
+\spc\spc{\small Mitzi Morris, \ Rob Trangucci}
 \vfill
@@
 \\
-\spc{\footnotesize Paris-Dauphine 2014}
+\spc{\small Stan 2.5.0 \ \footnotesize (October 2014)}
 \hfill 
@@
   \begin{itemize}\small
-\item declares variables
-\item codes log posterior (penalized likelihood)
+  \item declares data and (constrained) parameter variables
+  \item defines log posterior (or penalized likelihood)
   \end{itemize}
@@
 \begin{itemize}
-\item 750+ registrations for mailing list across the physical,
-biomedical, and social sciences, plus engineering, finance and marketing
-
-\item Subjects of Papers on Google Scholar in the sciences
-\\
-{\footnotesize clinical drug
-trials, general computational statistics, entomology, opthalmology,
-neurology, sociology and population dynamics, genomics, agriculture,
-psycholinguistics, molecular biology, population dynamics, materials
-engineering, botany, astrophysics, oceanography, election prediction,
-fisheries, cancer biology, public health and epidemiology, population
-ecology, collaborative filtering for recommender systems, climatology,
-educational testing, natural language processing}
+\item 850+ mailing list registrations;  5500 manual downloads (2.4.0)
+
+\item Biological sciences: {\footnotesize
+clinical drug trials, entomology, opthalmology,
+neurology, genomics, agriculture, botany, fisheries,
+cancer biology, epidemiology, population ecology, neurology
+}
+\item Physical sciences: {\footnotesize 
+astrophysics, molecular biology, oceanography, climatology
+}
+\item Social sciences: {\footnotesize
+ population dynamics, psycholinguistics, social networks, political science
+}
+\item Other: {\footnotesize materials engineering, finance, actuarial,
+  sports, public health,
+  collaborative filtering, educational testing}
 \end{itemize}
@@
 \begin{itemize}
-\item 400+ page core language tutorial manual
-\item separate installation, getting started, interface manuals 
-\\ {\footnotesize  (RStan, PyStan, CmdStan)}
-\item BUGS and JAGS examples (all 3 volumes), 
+\item 450+ page core language tutorial manual
+\item Installation and getting started manuals by interface
+  \\ {\footnotesize  (RStan, PyStan, CmdStan, MatlabStan, Stan.jl)}
+\item BUGS and JAGS examples (most of all 3 volumes), 
 \item Gelman and Hill, {\slshape Data Analysis Using Regression and
@@
   \item more parameters
-\item \myemph{more complex models}
-\end{itemize}
-\item for MCMC, measure (as function of scaling)
-\vspace*{-4pt}
-\begin{itemize}\small
-\item time to convergence
-\item time per effective sample at mixing
-\item memory usage
-\end{itemize}
-\end{itemize}
-
-\sld{Stan vs. BUGS / JAGS}
-\begin{itemize}
-\item JAGS is the former \myemph{state-of-the-art}, \myemph{black-box} MCMC method
+  \item \myemph{more complex models} \hfill (why we built Stan)
+  \end{itemize}
+\item for MCMC, measure \hfill (vs.\ BUGS / JAGS)
+  \vspace*{-4pt}
+  \begin{itemize}\small
+  \item time to convergence \hfill (0.5--{\large$\infty$} faster)
+  \item time per effective sample at mixing \hfill (ditto)
+  \item memory usage \hfill (90\% less, linear scaling)
+  \end{itemize}
+\end{itemize}
+
+\sld{Stan vs. JAGS (BUGS)}
+\begin{itemize}
+\item JAGS is the new BUGS, and former \myemph{state-of-the-art}, \myemph{black-box} MCMC method
   \\ {\footnotesize (Gibbs with slice sampling; chokes w.\ high posterior correlation)}
-\item Stan (2.0) vs. BUGS on scaling IRT model
+\item Stan (2.0) vs. BUGS on scaling multilevel 2PL-IRT model
   \vspace*{-4pt}
   \begin{itemize}\footnotesize
-\item Stan about \myemph{10 times faster}
-\item Stan memory and time scales linearly
-\item linearly scales with number of question answers
+  \item Stan about \myemph{10 times faster}  ($n_{eff}/s$) and uses
+    10\% memory
   \end{itemize}
@@
   \item \myemph{PyStan}: Python interface (Cython in memory)
-\item \myemph{MStan$^*$}: MATLAB interface (lightweight external process)
-\item \myemph{JuliaStan$^*$}: Julia interface (lightweight external process)
+  \item \myemph{MatlabStan$^*$}: MATLAB interface (lightweight external process)
+  \item \myemph{Stan.jl$^*$}: Julia interface (lightweight external process)
   \end{itemize}
@@
 
+\sld{NUTS vs.\ Gibbs and Metropolis}
+
+\includegraphics[width=0.9\textwidth]{img/nuts-vs.pdf}
+\begin{itemize}
+\item Two dimensions of highly correlated 250-dim distribution
+\item 1M samples from Metropolis, 1M from Gibbs (thin to 1K)
+\item 1K samples from NUTS, 1K independent draws
+\end{itemize}
+
+\sld{NUTS vs.\ Basic HMC}
+
+\includegraphics[width=0.9\textwidth]{img/nuts-ess-1.pdf}
+{\small
+  \begin{itemize}
+  \item 250-D normal and logistic regression models
+  \item Vertical axis is effective sample size per sample (bigger better)
+  \item Left) NUTS; \ \ Right) HMC with increasing $t = \epsilon L$
+  \end{itemize}
+}
+
+\sld{NUTS vs.\ Basic HMC II}
+
+\includegraphics[width=0.9\textwidth]{img/nuts-ess-2.pdf}
+
+{\small
+  \begin{itemize}
+  \item Hierarchical logistic regression and stochastic volatility
+  \item Simulation time $t$ is $\epsilon \ L$, step size ($\epsilon$)
+    times number of steps ($L$)
+  \item NUTS can beat optimally tuned HMC (latter very expensive)
+  \end{itemize}
+}
+
 \sld{Reverse-Mode Auto Diff}
@@
 \item \myemph{traits} metaprogram to \myemph{drop constants} (e.g., $-\log
-  \sqrt{2 \pi}$) and calculate intermediate and return types
+  \sqrt{2 \pi}$ or $\log \sigma$ if constant) 
+  and calculate intermediate and return types
 \end{itemize}
@@
 
-\mypart{Part III}{What's Next?}
-
 \sld{Differential Equation Solver}
@@
 
-\sld{L-BFGS Optimizer}
-\begin{itemize}
-\item Faster, lower memory, more stable than BFGS
-\item Plug-and-play in interfaces
-\vfill
-\item Code and doc complete; under testing
-\end{itemize}
+\mypart{Part III}{What's Next?}
 
@@
 
-\sld{Thermodynamic Sampler}
+\sld{Adiabatic Sampling}
 \begin{itemize}
@@
 
-\sld{Marginal Maximum Likelihood}
-\begin{itemize}
-\item Marginalize out hierarchical parameters
-\item Gradient-based nested optimization algorithm
-\item Errors estimated via curvature as in MLE
+\sld{Maximum Marginal Likelihood}
+\begin{itemize}
+\item Fast, Approximate Inference
+\item Marginalize out lower-level parameters
+\item Optimize higher-level parameters and fix
+\item Optimize lower-level parameters given higher-level
+\item Errors estimated as in MLE
   \vfill
@@
 
-\sld{``Black Box'' VB and EP}
-\begin{itemize}
+\sld{``Black Box'' VB}
+\begin{itemize}
+\item Fast, Approximate Inference
 \item \myemph{Black box} so can run any model
@@
 \item Stochastic, data-streaming \myemph{variational Bayes} (VB)
+\item Optimize parameteric approximation to posterior to minimize KL
+  divergence 
+  \vfill
+\item Prototype stage (collaborating with Alp Kucukelbir, Dave Blei,
+  Rajesh Ranganath)
+\end{itemize}
+
+\sld{``Black Box'' EP}
+\begin{itemize}
+\item Fast, Approximate Inference
 \item Data-parallel \myemph{expectation propagation} (EP)
   \\ {\footnotesize (cavity distributions provide general shard combination)}
-\item Both use point estimates of parametric approximation to posterior
-\item Optimize parameters to minimize KL divergence
-\\
+\item Optimize parameteric approximation to posterior to minimize KL
+  divergence 
   {\footnotesize (VB, EP measure divergence in opposite directions)}
   \vfill
-\item Prototype stage
+  \vfill
+\item Design stage (collaborating with Nicolas Chopin, Christian
+  Robert, John Cunningham, Aki Vehtari, Pasi Jyl\"anki)
 \end{itemize}
@@
 \item Posterior proportional to likelihood times prior (i.e., joint)
+\end{itemize}
+
+\sld{``Naive Bayes'' Four Ways}
+\begin{itemize}
+%
+\item Joint Distribution \ $p(\pi,\phi,z,w)$, defined by:
+\vspace*{-3pt}
+\begin{itemize}\small
+\item $\pi \sim \mbox{\sf Dirichlet}(\alpha)$    \hfill (topic prevalence)
+\item $\phi_k \sim \mbox{\sf Dirichlet}(\beta)$  \hfill (word
+  prevalence in topic $k$)
+\item $z_d \sim \mbox{\sf Categorical}(\pi)$     \hfill (topic for doc $d$)
+\item $w_{d,n} \sim \mbox{\sf Categorical}(\phi_{z_d})$  \hfill (word
+  $n$ in doc $d$)
+\end{itemize}
+%
+\item Inference Problem \ $p(\tilde{z},\phi,\pi |  w, z, \tilde{w})$
+\vspace*{-3pt}
+\begin{itemize}\small
+\item fully supervised learning: \ $p(\pi,\phi | w, z)$
+\item semi-supervised learning: \ $p(\pi, \phi | w, z,
+  \tilde{w})$
+\item clustering: \ $p(\tilde{z} | \tilde{w})$
+\item prediction: \ $p(\tilde{z} | w, z, \tilde{w})$
+\end{itemize}
 \end{itemize}

### 2014-10-29 src/docs/talks/paris-dauphine-2014/stan-overview.tex
added intro slides back in

@@
 {\footnotesize \tt mc-stan.org}
+
+\sld{Why Stan?}
+%
+\begin{itemize}
+\item {\slshape\bfseries Application}: Fit rich Bayesian statistical models
+\end{itemize}
+%
+\vspace*{2pt}
+\begin{itemize}
+\item {\slshape Problem}: Gibbs and Metropolis too slow (diffusive) 
+\item {\slshape Solution}: Hamiltonian Monte Carlo (flow)
+%
+\vspace*{8pt}
+\item {\slshape Problem}:  Interpreters slow and unscalable
+\item {\slshape Solution}: Compiled to C++
+%
+\vspace*{8pt}
+\item {\slshape Problem}:  Need gradients of log posterior for HMC
+\item {\slshape Solution}: Reverse-mode algorithmic differentation
+\end{itemize}
+
+\sld{Why? (cont.)}
+%
+\begin{itemize}
+\item {\slshape Problem}:  Existing algo-diff slow, limited, unextensible
+\item {\slshape Solution}: Our own algo-diff
+%
+\vspace*{8pt}
+\item {\slshape Problem}:  Algo-diff requires functions templated on
+  all args
+\item {\slshape Solution}: Our own density library, Eigen linear
+ algebra
+%
+\vspace*{8pt}
+\item {\slshape Problem}:  Need unconstrained parameters for HMC
+\item {\slshape Solution}: Variable transforms w. Jacobian determinants
+%
+\end{itemize}
+
+\sld{Why? (cont.)}
+%
+\begin{itemize}
+\item {\slshape Problem}:  Need ease of use of BUGS
+\item {\slshape Solution}: Compile a domain-specific language
+%
+\vspace*{8pt}
+\item {\slshape Problem}:  Pure directed graphical language inflexible
+\item {\slshape Solution}: Imperative probabilistic programming language
+
+\vspace*{8pt}
+\item {\slshape Problem}:  Need to tune parameters for HMC
+\item {\slshape Solution}: Tune step size and estimate mass matrix
+  during warmup;  on-the-fly number of steps (NUTS)
+%
+\end{itemize}
+
+\sld{Why? (cont.)}
+\begin{itemize}
+%
+\vspace*{8pt}
+\item {\slshape Problem}:  Efficient up-to-proportion density calcs
+\item {\slshape Solution}: Density template metaprogramming 
+%
+\vspace*{8pt}
+\item {\slshape Problem}:  Limited error checking, recovery
+\item {\slshape Solution}: Static model typing, informative exceptions
+%
+\vspace*{8pt}
+\item {\slshape Problem}:  Poor boundary behavior
+\item {\slshape Solution}: Calculate limits (e.g. $\lim_{x \rightarrow
+    0} x \log x$)
+%
+\end{itemize}
+
+\sld{Why? (continued)}
+\begin{itemize}
+\item {\slshape Problem}:  Nobody knows everything
+\item {\slshape Solution}: Expand project team with specialists
+\vspace*{8pt}
+\item {\slshape Problem}:  Nobody knows everything
+\item {\slshape Solution}: Expand project team with specialists
+\vspace*{8pt}
+\item {\slshape Problem}:  Heterogeneous user base
+\item {\slshape Solution}: More interfaces (R, Python, MATLAB, Julia),
+  domain-specific examples, tutorials
+
+\end{itemize}
+
+\sld{Why? (continued)}
+\begin{itemize}
+\item {\slshape Problem}:  Expanding project team and interfaces
+\item {\slshape Solution}: GitHub process: branch, pull request, code
+  review, continuous integration
+%
+\vspace*{8pt}
+\item {\slshape Problem}:  Restrictive licensing (e.g., closed, GPL, etc.)
+\item {\slshape Solution}: Open-source, BSD license
+\end{itemize}
 

### 2014-11-08 src/docs/talks/paris-dauphine-2014/stan-overview.tex
updated talk to remove redundancy and move ode discussion up with funs

@@
 \item {\slshape Problem}:  Pure directed graphical language inflexible
-\item {\slshape Solution}: Imperative probabilistic programming language
-
+\item {\slshape Solution}: Imperative probabilistic programming
+  language
 \vspace*{8pt}
@@
 \vspace*{8pt}
-\item {\slshape Problem}:  Nobody knows everything
-\item {\slshape Solution}: Expand project team with specialists
+\item {\slshape Problem}:  Expanding code and project team
+\item {\slshape Solution}: GitHub: branch, pull 
+  request, code review
+\item {\slshape Solution}: Jenkins: continuous integration
+\item {\slshape Solution}: ongoing refactoring and code simplification
+%
+\end{itemize}
+
+\sld{Why? (continued)}
+\begin{itemize}
+\item {\slshape Problem}:  Heterogeneous user base
+\item {\slshape Solution}: More interfaces (R, Python, MATLAB, Julia)
+\item {\slshape Solution}:  domain-specific examples, tutorials
 \vspace*{8pt}
-\item {\slshape Problem}:  Heterogeneous user base
-\item {\slshape Solution}: More interfaces (R, Python, MATLAB, Julia),
-  domain-specific examples, tutorials
-
-\end{itemize}
-
-\sld{Why? (continued)}
-\begin{itemize}
-\item {\slshape Problem}:  Expanding project team and interfaces
-\item {\slshape Solution}: GitHub process: branch, pull request, code
-  review, continuous integration
-%
-\vspace*{8pt}
-\item {\slshape Problem}:  Restrictive licensing (e.g., closed, GPL, etc.)
-\item {\slshape Solution}: Open-source, BSD license
+\item {\slshape Problem}:  Restrictive licensing limits use
+\item {\slshape Solution}: Code and doc open source
+(BSD, CC-BY)
 \end{itemize}
@@
 \item \myemph{Specializations}: triangular, positive-definite, etc.
+\end{itemize}
+
+\sld{Differential Equation Solver}
+\begin{itemize}
+\item Auto-diff solutions w.r.t.\ parameters
+\item Integrate coupled system for solution with partials
+\item Auto-diff coupled Jacobian for stiff systems
+  \vfill
+\item C++ prototype integrated for large PK/PD models
+  \vspace*{-4pt}
+  \begin{itemize}\footnotesize
+  \item 
+    Project with Novartis: 
+    longitudinal clinical trial w.\ multiple drugs, dosings, placebo control,
+    hierarchical model of patient-level effects, meta-analysis
+  \item
+    Collaborators: Frederic Bois, Amy Racine, Sebastian Weber
+  \end{itemize}
 \end{itemize}
@@
 
-\sld{Differential Equation Solver}
-\begin{itemize}
-\item Auto-diff solutions w.r.t.\ parameters
-\item Integrate coupled system for solution with partials
-\item Auto-diff coupled Jacobian for stiff systems
-  \vfill
-\item C++ prototype integrated for large PK/PD models
-  \vspace*{-4pt}
-  \begin{itemize}\footnotesize
-  \item 
-    Project with Novartis: 
-    longitudinal clinical trial w.\ multiple drugs, dosings, placebo control,
-    hierarchical model of patient-level effects, meta-analysis
-  \item
-    Collaborators: Frederic Bois, Amy Racine, Sebastian Weber
-  \end{itemize}
-\end{itemize}
-
 \mypart{Part III}{What's Next?}

### 2014-12-01 src/docs/papers/stan-language/article.tex
updated article for reviewer comments

@@
 \renewcommand{\topfraction}{0.95}	% max fraction of floats at top
+
+\newenvironment{absolutelynopagebreak}
+  {\par\nobreak\vfil\penalty0\vfilneg
+   \vtop\bgroup}
+  {\par\xdef\tpd{\the\prevdepth}\egroup
+   \prevdepth=\tpd}
 
@@
   \proglang{R} using the \pkg{RStan} package, or through
-<  \proglang{Python} using the \pkg{PyStan} package.  All three
+  \proglang{Python} using the \pkg{PyStan} package.  All three
   interfaces support sampling or optimization-based inference and
   analysis, and \pkg{RStan} and \pkg{PyStan} also provide access to
-  log probabilities, gradients, Hessians, and I/O transforms.
+  log probabilities, gradients, Hessians, and data I/O.
 }
@@
 
-\section[Why Stan?]{Why \proglang{Stan}?}
-%% Note: If there is markup in \(sub)section, then it has to be escape as above.
-
-We did not set out to build \proglang{Stan} as it currently exists.%
-%
-\footnote{\proglang{Stan}'s home page, \url{http://mc-stan.org/},
-  links to the \proglang{Stan} manual, example models including translations of
-  the \proglang{BUGS} examples and the models from \citep{GelmanHill:2007},
-  getting started instructions and full documentation for \proglang{Stan}'s
-  interfaces for the command line shell (\pkg{CmdStan}), \proglang{Python}
-  (\pkg{PyStan}), and \proglang{R} (\pkg{RStan}), and the source code repositories
-  and issue trackers.}
-%
-Our original goal was to apply full Bayesian inference to the sort of multilevel
-generalized linear models discussed in Part~II of
-\citep{GelmanHill:2007}, which are structured with grouped and
-interacted predictors at multiple levels, hierarchical covariance
-priors, nonconjugate coefficient priors, latent effects as in
-item-response models, and varying output link functions and
-distributions.
-
-These models turned out to be a challenge for existing general purpose
-inference software.  A direct encoding in \proglang{BUGS}
-\citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
-\proglang{JAGS}~\citep{Plummer:2003} can grind these tools to a halt.
-We began by adding custom vectorized logistic regressions to
-\proglang{JAGS} using \proglang{C++} to overcome the cost of
-interpretation.  Although this is much faster than looping in
-\proglang{JAGS}, it quickly became clear that the root of the problem
-was the slow convergence of conditional sampling when parameters were
-highly correlated in the posterior, as for time-series models and
-hierarchical models with interacted predictors.  We finally realized
-we needed a better sampler, not a more efficient implementation of
-Gibbs sampling.
-
-We briefly considered trying to tune proposals for a random-walk
-Metropolis-Hastings sampler, but that seemed too problem specific and
-not even necessarily possible without some kind of adaptation rather
-than a global tuning of the proposals.
-
-We were hearing more about Hamiltonian Monte Carlo (HMC) sampling,
-which appeared promising but was also problematic in that the
-Hamiltonian dynamics simulation requires the gradient of the log
-posterior.  Although not difficult in theory for most functions,
-computing these gradients by hand on a model-by-model basis is very
-tedious and error prone.  That is when we discovered reverse-mode
-algorithmic differentiation, which, given a templated \proglang{C++}
-function for the log posterior, automatically computes its analytic
-gradient up to machine precision accuracy in only a few multiples of
-the time to evaluate the log probability function itself.%
-%
-\footnote{Hessian matrices (i.e., all second order derivatives) are
-  more expensive to calculate; each row corresponds to a parameter and
-  is filled by the gradient of the derivative of the log probability
-  function with respect to the parameter.}
-%
-We explored existing algorithmic differentiation packages with open
-licenses such as \pkg{RAD} \citep{Gay:2005} and its repackaging in the
-Sacado module of the \pkg{Trilinos} toolkit, and the \pkg{CppAD}
-package of the COIN-OR toolkit \citep{BellBurke:2008}.  Both packages,
-however, supported few special functions (e.g., probability functions,
-log gamma, inverse logit) or linear algebra operations (e.g., Cholesky
-decompositions, matrix division), and neither are easily or modularly
-extensible.
-
-Consequently we built our own reverse-mode algorithmic differentiation package.
-At that point, we ran into the problem that we could not just plug in
-the probability functions from a package like \pkg{Boost} because they
-weren't templated generally enough across all arguments.  Rather
-than pay the price of promoting floating point values to algorithmic
-differentiation variables, we wrote our own fully templated
-probability functions and other special functions.
-
-Next, we integrated the \proglang{C++} package \pkg{Eigen} for matrix
-operations and linear algebra functions.  \pkg{Eigen} makes extensive
-use of expression templates for lazy evaluation and the curiously
-recurring template pattern (CRTP) to implement concepts without
-virtual function calls; \cite{VandevoordeJosuttis:2002} provide a
-complete description of template metaprogramming techniques.
-Unfortunately, we ran into the same problem with Eigen as with the
-existing probability libraries---it doesn't support mixed operations
-of algorithmic differentiation variables and primitives like
-\code{double}.  Although we initially began by promoting
-floating-point vectors to algorithmic differentiation variables, we
-later completely rewrote mixed-mode operations like multiplying a data
-matrix by a parameter vector.  To compute derivatives of matrix
-operations such as division and inverse, we implemented specialized
-derivatives using the rules described by \cite{Giles:2008}.
-
-At this point, we could fit models coded directly in \proglang{C++} on
-top of the pre-release versions of the \proglang{Stan} application
-programming interface (API).  Seeing how well this all worked, we set
-our sights on the generality and ease of use of \proglang{BUGS} and
-designed a modeling language in which statisticians could write their
-models in familiar notation that could be transformed to efficient
-\proglang{C++} code before being compiled into an efficient executable
-program.  Although we started with the target of specifying directed
-graphical models, our first implementation turned out to be a more
-general imperative form of specifying log probability functions;
-\proglang{BUGS} models could be translated line for line, but the
-\proglang{Stan} language also supports much richer imperative
-constructs like conditionals, while loops, and local variables.  This
-paper is primarily about the \proglang{Stan} language and how it can
-be used to specify statistical models.
-
-The next problem we ran into when we started implementing richer
-models was variables with constrained support, such as positive variables, 
-simplexes, and covariance matrices.  
-Efficient implementation of these constraints
-required the introduction of typed variables which automatically transformed to
-unconstrained support with suitable adjustments to the log probability
-from the log absolute Jacobian determinant of the inverse transforms.
-
-Even with the prototype compiler generating models, we still faced a
-major hurdle for ease of use. The efficiency of HMC is very sensitive
-to two tuning parameters, the discretization interval (i.e., step
-size) and the total simulation time (i.e., number of steps).  The
-interval size parameter could be tuned during warm-up based on
-Metropolis rejection rates, but the number of steps proved difficult
-to tune without sacrificing the detailed balance of the sampler.  This
-led to the development of the No U-Turn (NUTS) sampler
-\citep{HoffmanGelman:2011}.  Roughly speaking, NUTS builds a tree of
-possible samples by randomly simulating Hamiltonian dynamics both
-forwards and backwards in time until the combined trajectory turns
-back on itself.  Once the trajectory has terminated, a new sample is
-drawn from the tree.  In its original version, NUTS also estimates a
-step size during warmup based on a target acceptance probability.
-
-We thought we were home free, but when we measured the
-speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
-very disappointed.  \proglang{BUGS}'s very first example model, Rats,
-ran more than an order of magnitude faster in \proglang{JAGS} than in
-\proglang{Stan}.  Rats is a tough test case because the conjugate
-priors and lack of posterior correlations make it an ideal candidate
-for efficient Gibbs sampling.  Realizing that we were doing redundant
-calculations, we wrote a vectorized form of the normal distribution
-for multiple variates with the same mean and scale, which sped things
-up a bit.  Performance was lacking until we finally figured out how
-to both eliminate redundant calculations and partially evaluate the
-gradients using a combination of expression templates and
-metaprogramming.
-
-When we attempted to fit a time-series model, we found that
-normalizing the data to unit sample mean and variance sped up the fits
-by an order of magnitude.  In hindsight this isn't surprising, as the
-performance of the numerical simulation in HMC scales with the
-variation of the parameter scales.  In \proglang{Stan} 1.0 we
-introduced an adaptive diagonal metric (mass matrix) into our HMC
-implementations that allowed the parameter scales to be standardized
-automatically; \proglang{Stan} 2.0 added an option to estimate a dense metric
-(mass matrix) and a full covariance standardization of the posterior.
-
-Because the diagonal and dense metrics perform only a global
-standardization, the performance of \proglang{Stan} in models where
-the relative parameter scales varies by location, such as hierarchical
-and latent models, can suffer (hierarchical modeling is covered in
-Section~\ref{hierarchical-modeling.section} and reparameterizations to speed it
-up in Section~\ref{centering.section}).  Riemannian manifold Hamiltonian Monte
-Carlo (RHMC) introduces a location-dependent metric that can overcome
-these final hurdles \citep{GirolamiCalderhead:2011}; \proglang{Stan} has a
-prototype implementation of RHMC based on the SoftAbs metric of
-\citep{Betancourt:2012}, using a generalization of NUTS to Riemannian
-manifolds \citep{Betancourt:2013}.
-
-\section{Overview}
+\section[Introduction]{Introduction}
+
+The goal of the \proglang{Stan} project is to provide a flexible
+probabilistic programming language for statistical modeling along with
+a suite of inference tools for fitting models that are robust,
+scalable, and efficient.
+
+\proglang{Stan} differs from \proglang{BUGS}
+\citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} and
+\proglang{JAGS}~\citep{Plummer:2003} in two primary ways.  First,
+\proglang{Stan} is based on a new imperative probabilistic programming
+language that is more flexible and expressive than the declarative
+graphical modeling languages underlying \proglang{BUGS} or
+\proglang{JAGS}, in ways such as declaring variables with types and
+supporting local variables and conditional statements.  Second,
+\proglang{Stan}'s Markov chain Monte Carlo (MCMC) techniques are based
+on Hamiltonian Monte Carlo (HMC), a more efficient and robust sampler
+than Gibbs sampling or Metropolis-Hastings for models with complex
+posteriors.%
+%
+\footnote{\cite{Neal:2011} analyzes the scaling benfit of HMC with
+  dimensionality.  \cite{Hoffman-Gelman:2014} provide practical
+  comparisions of \proglang{Stan}'s adaptive HMC algorithm with Gibbs,
+  Metropolis, and standard HMC samplers.}
+
+\proglang{Stan} has interfaces for the command-line shell
+(\pkg{CmdStan}), \proglang{Python} (\pkg{PyStan}), and \proglang{R}
+(\pkg{RStan}), and runs on Windows, Mac OS X, and Linux, and is
+open-source licensed.
+
+The next section provides an overview of how \proglang{Stan} works by
+way of an extended example, after which the details of \proglang{Stan}'s
+programming language and inference mechanisms are provided.
+
+\section{Core Functionality}\label{example.section}
 
@@
 %
+\begin{absolutelynopagebreak}
 \begin{Code}
@@
 \end{Code}
+\end{absolutelynopagebreak}
 %
@@
 Data for running \proglang{Stan} from the command line can be included
-in \proglang{R} dump format.%
-%
-\footnote{A \pkg{JSON} interface for structured data input and output is
-  currently under development.}
-%
-For example, 10 observations for the
-model in Figure~\ref{bernoulli-model.fig} could be encoded as%
+in \proglang{R} dump format.  All of the variables declared in the
+data block of the \proglang{Stan} program must be defined in the data
+file.  For example, 10 observations for the model in
+Figure~\ref{bernoulli-model.fig} could be encoded as%
 %
@@
 %
-For Windows, the \code{./} before the command should be removed.
+For Windows, the \code{./} before the command should be removed.  This
+call specifies that sampling should be performed with the model
+instantiated using the data in the specified file.
 
@@
 The sampler configuration parameters are echoed, here they are all
-default values other than the data file.  These parameters may be set
-on the command line.
+default values other than the data file.
+
+The command-line parameters marked \code{Default} may be explicitly
+set on the command line.  Each value is preceded by the full path to
+it in the hierarchy; for instance, to set the maximum depth for the
+no-U-turn sampler, the command would be the following, where backslash
+indicates a continued line.
+%
+\begin{CodeChunk}
+\begin{CodeInput}
+> ./bernoulli sample  \
+  algorithm=hmc engine=nuts max_depth=5  \
+  data max_depthfile=bernoulli.data.R
+\end{CodeInput}
+\end{CodeChunk}
+%
 
@@
 \proglang{Stan}'s behavior is fully specified by these configuration
-parameters.  By using the same version of \proglang{Stan} and these
-configuration parameters, exactly the same output file can be
-reproduced.  The pseudorandom numbers generated by the sampler are
-fully determined by the seed (here randomly generated based on the
-time of the run, with value \code{847896134}) and the identifier (here
-\code{0}).  The identifier is used to advance the underlying
-pseudorandom number generator a sufficient number of values that
-using multiple chains with the same seed and different identifiers
-will draw from different subsequences of the pseudorandom number
-stream determined by the seed.
+parameters, almost all of which have default values.  The
+\code{sample} configuration
+
+ By using the same version of
+\proglang{Stan} and these configuration parameters, exactly the same
+output file can be reproduced.  The pseudorandom numbers generated by
+the sampler are fully determined by the seed (here randomly generated
+based on the time of the run, with value \code{847896134}) and the
+identifier (here \code{0}).  The identifier is used to advance the
+underlying pseudorandom number generator a sufficient number of values
+that using multiple chains with the same seed and different
+identifiers will draw from different subsequences of the pseudorandom
+number stream determined by the seed.
 
@@
 %
-The values for \code{lp\_\_} indicate the log probability (up to an
-additive constant).  The column headed by \code{accept\_stat\_\_}
-provides the Metropolis/slice acceptance statistic for each
-iteration.%
+The values headed by \code{lp\_\_} are the log densities (up to an
+additive constant), \code{accept\_stat\_\_} are the Metropolis
+acceptance proababilities averaged over samples in the slice used by
+the no-U-turn sampler, \code{stepsize\_\_} is the leapfrog
+integrator's step size for simulating the Hamiltonian,
+\code{treedepth\_\_} is the depth of tree explored by the no-U-turn
+sampler, and \code{n\_divergent\_\_} is the number of iterations leading
+to a numerical instability during integration (e.g., numerical
+overflow or a positive-definiteness violation).
+
+ for each iteration.%
 %
@@
 
-\section{Models}
+\subsection{Roadmap for the Rest of the Paper}
+
+Now that the key functionality of \proglang{Stan} has been
+demonstrated, the remaining sections cover specific aspects of
+\proglang{Stan}'s architecture.  Section~\ref{data-types.section}
+covers variable data type declarations as well as expressions and type
+inference, Section~\ref{programming.section} describes the top-level
+blocks and execution of a \proglang{Stan} program,
+Section~\ref{statements.section} lays out the available statements,
+and Section~\ref{functions.section} the built-in math, matrix, and
+probability function library.  Section~\ref{inference-engines.section}
+lays out MCMC and optimization-based inference.  There are two
+appendices, Appedix~\ref{process.section} outlining the development
+process and Appendix~\ref{libraries.section} detailing the library
+dependencies.
+
+\section{Data types}\label{data-types.section}
+
+All expressions in \proglang{Stan} are statically typed, including
+variables.  This means their type is declared at compile time as part
+of the model, and does not change throughout the execution of the
+program.  This is the same behavior as is found in compiled
+programming languages such as \proglang{C(++)}, \proglang{Fortran},
+and \proglang{Java}, but is unlike the behavior of interpreted
+languages such as \proglang{BUGS}, \proglang{R}, and
+\proglang{Python}.  Statically typing the variables (as well as
+declaring them in appropriate blocks based on usage) makes
+\proglang{Stan} programs easier to read and easier to debug by making
+explicit the modeling decisions and expression types.
+
+\subsection{Primitive types}
+
+The primitive types of \proglang{Stan} are \code{real} and \code{int},
+which are used to represent continuous and integer values.  These
+values are represented directly in \proglang{C++} as types
+\code{double} and \code{int}.  Integer expressions can be used
+anywhere a real value is required, but not {\it vice-versa}.
+
+\subsection{Vector and matrix types}
+
+\proglang{Stan} supports vectors, row vectors, and matrices with the
+usual access operations.  Indexing for vector, matrix, and array types
+starts from one.
+
+Vectors are declared with their sizes and matrices with their number
+of rows and columns.  Vector, row vector, and matrix elements are
+accessed using bracket notation, as in \code{y[3]} for the third
+element of a vector or row vector and \code{a[2,3]} for the element in
+the third column of the second row of a matrix.  Indexing begins from
+1.  The notation \code{a[2]} accesses the second row of matrix \code{a}.
+
+All vector and matrix types contain real values and may not be
+declared to contain integers.  Collections of integers are represented
+using arrays.
+
+\subsection{Array types}
+
+An array may have entries of any other type.  For example, arrays of
+integers and reals are allowed, as are arrays of vectors or arrays of
+matrices.
+
+Higher-dimensional arrays are intrinsically arrays of arrays.  An
+entry in a two-dimensional array \code{y} may be accessed as
+\code{y[1,2]}.  The expression \code{y[1]} by itself denotes the
+one-dimensional array whose values correspond to the first row of
+\code{y}.  Thus \code{y[1][2]} has the same value as \code{y[1,2]}.%
+%
+\footnote{Arrays are stored in row-major order and matrices in
+  column-major order.}
+
+Unlike integers, which may be used where real values are required,
+arrays of integers may not be used where real arrays are required.%
+%
+\footnote{In the language of type theory, \proglang{Stan} arrays are
+  not covariant.  This follows the behavior of both arrays and
+  standard library containers in \proglang{C++}.}
+
+The manual contains a chapter discussing the efficiency tradeoffs and
+motivations for separating arrays and matrices.
+
+\subsection{Constrained variable types}
+
+Variables may be declared with constraints.  The constraints have
+different effects depending on the block in which the variable is
+declared.
+
+Integer and real types may be provided with lower bounds, upper
+bounds, or both.  This includes the types used in arrays, and the real
+types used in vectors and matrices.
+
+Vector types may be constrained to be unit simplexes (all entries
+non-negative and summing to 1), unit length vectors (sum of squares is 1), or
+ordered (entries are in ascending order), positive ordered (entries in
+ascending order, all non-negative), using the types \code{simplex[K]},
+\code{unit_vector[K]}, \code{ordered[K]}, or
+\code{positive_ordered[K]}, where \code{K} is the size of the vector.
+
+Matrices may be constrained to be covariance matrices (symmetric,
+positive definite) or correlation matrices (symmetric, positive
+definite, unit diagonal), using the types \code{cov_matrix[K]} and
+\code{corr_matrix[K]}.
+
+\subsection{Expressions}\label{expressions.section}
+
+The syntax of \code{Stan} is defined in terms of expressions and
+statements.  Expressions denote values of a particular type.  
+Statements represent operations such as assignment and sampling as
+well as control structures such as for loops and conditionals.
+
+\proglang{Stan} provides the usual kinds of expressions found in
+programming languages.  This includes variables, literals denoting
+integers, real values or strings, binary and unary operators over
+expressions, and function application.
+
+\subsubsection{Type inference}
+
+The type of a numeric literal is determined by whether or not it
+contains a period or scientific notation; for example, \code{20} has
+type \code{int} whereas \code{20.0} and \code{2e+1} have type
+\code{real}.
+
+The type of applying an operator or a function to one or more
+expressions is determined by the available signatures for the
+function.  For example, the multiplication operator (\code{*}) has a
+signature that maps two \code{int} arguments to an \code{int} and two
+\code{real} arguments to a \code{real} result.  Another signature for
+the same operator maps a \code{row\_vector} and a \code{vector} to a
+\code{real} result.
+
+\subsubsection{Type promotion}
+
+If necessary, an integer type will be promoted to a \code{real} value.
+For example, multiplying an \code{int} by a \code{real} produces a
+\code{real} result by promoting the \code{int} argument to a
+\code{real}.
+
+\section{Top-Level Blocks and Program Execution}\label{programming.section}\label{hierarchical-modeling.section}
 
@@
 
-\subsection{Example: hierarchical model, with inference}\label{hierarchical-modeling.section}
-
-\cite[Section 5.1]{GelmanEtAl:2013} define a hierarchical model of the
-incidence of tumors in rats in control groups across trials; a very
-similiar model is defined for mortality rates in pediatric surgeries
-across hospitals in \citep[Examples, Volume
+We begin with an example that will be used throughout the rest of this
+section.  \cite[Section 5.1]{GelmanEtAl:2013} define a hierarchical
+model of the incidence of tumors in rats in control groups across
+trials; a very similiar model is defined for mortality rates in
+pediatric surgeries across hospitals in \citep[Examples, Volume
 1]{LunnEtAl:2000,LunnEtAl:2009}.
@@
 variables.  For example, the following transformed data block declares
-a vector \code{x\_std}, then defines it to be the standardization of \code{x}.
+a vector \code{x\_std}, then defines it to be the standardization of \code{x}:
 %
@@
 \code{lambda} are constrained to fall between zero and one, whereas
-\code{lambda} is constrained to be greater than or equal to 0.1 to
+\code{kappa} is constrained to be greater than or equal to 0.1 to
 match the support of the Pareto hyperprior it receives in the model block.
@@
 evaluated.  This may be multiple times per iteration of a sampling or
-optimization algorithm.  Furthermore, different samplers and
-optimizers use different instantiations of the log probability
-function depending on the form and order of the derivative information
-they require; see Section~\ref{inference-engines.section} for
-details.
+optimization algorithm.
 
@@
 
-The transformed parameter block is executed after the parameter
-block.  Constraints are validated after all of the statements defining the
+The transformed parameter block is executed after the parameter block.
+Constraints are validated after all of the statements defining the
 transformed parameters have executed.  Failure to validate a
@@
 execution of the log probability function.  The log probability
-function can be defined to return negative infinity or a special
-not-a-number value.
+function can be defined to return negative infinity or the special
+not-a-number value, both of which are available through built-in
+functions and may be passed to the \code{increment\_log\_prob}
+function (see below).
 
 If transformed parameters are used on the left-hand side of a sampling
-statement, it is up to the user to add the appropriate log Jacobian
-adjustment to the log probability accumulator.  For instance, a
-lognormal variate could be generated as follows without the built-in
-\code{lognormal} density function  using the normal density as
+statement, it is up to the user to add the appropriate log absolute
+Jacobian determinant adjustment to the log probability accumulator.
+For instance, a lognormal variate could be generated as follows
+without the built-in \code{lognormal} density function using the
+normal density as
 %
@@
   v <- log(u);
-  increment_log_prob(u);      // log absolute Jacobian adjustment
+  // log absolute Jacobian determinant adjustment
+  increment_log_prob(u);      
 }
@@
 The transorm is $f(u) = \log u$, the inverse transform is $f^{-1}(v) =
-\exp v$, so the absolute log Jacobian is $|\frac{d}{dv} \exp v| =
-\exp v = u$.  Whenever a transformed parameter is used on the left
-side of a sampling statement, a warning is printed to remind the user
-of the need for a Jacobian adjustment for the change of variables.
+\exp v$, so the absolute log Jacobian determinant is $|\frac{d}{dv}
+\exp v| = \exp v = u$.  Whenever a transformed parameter is used on
+the left side of a sampling statement, a warning is printed to remind
+the user of the need for a Jacobian adjustment for the change of
+variables.
+
+The \code{increment\_log\_prob} statement is used to add a term to the
+total log probability function defined by the model block and the log
+absolute Jacobian determinants of the transforms.  The variable
+\code{lp\_\_}, representing the currently accumulated total log
+density, may not be assigned to directly.
 
@@
 
-The model block defines the log probability function on the
-constrained parameter space.  The example in
+The purpose of the model block is to define the log probability
+function on the constrained parameter space.  The example in
 Figure~\ref{hier-binom.fig} has a simple model containing four
@@
 $\kappa^{-5/2}$.  Note that the hierarchical prior on \code{theta} is
-vectorized: each element of \code{theta} is drawn independently from
-a beta distribution with prior success count \code{alpha} and prior
+vectorized: each element of \code{theta} is drawn independently from a
+beta distribution with prior success count \code{alpha} and prior
 failure count \code{beta}.  Both \code{alpha} and \code{beta} are
@@
 
+\subsubsection{Implicit uniform priors}\label{implicit-prior.section}
+
+The default distribution for a variable is uniform over its
+declared (constrained) support.  For instance, a variable declared with a lower
+bound of 0 and an upper bound of 1 implicitly receives a $\mbox{\sf
+  Uniform}(0,1)$ distribution.  These implicit uniform priors are
+improper if the variable has unbounded support.  For instance, the
+uniform distributions over real values with upper and lower bounds,
+simplexes and correlation matrices is proper, but the uniform
+distribution over unconstrained or one-side constrained reals, ordered
+vectors or covariance matrices are not proper.
+
+\proglang{Stan} does not require proper priors, but if the posterior
+is improper, \proglang{Stan} will halt with an error message.%
+%
+\footnote{Improper posteriors are diagnosed automatically when
+  parameters overflow to infinity during simulation.}
+
 \subsection{Generated quantities block}
@@
 used to calculate predictive inferences as well as to carry out
-forward simulation for predictive posterior checks; see
+forward simulation for posterior predictive checks; see
 \citep{GelmanEtAl:2013} for examples.
@@
 \code{highest[j]} to 1 if hospital \code{j} has the highest estimated
-mortality rate.  Applying a hierarchical model then considering
-posterior inference appropriately adjusts for mulitple comparisons;
-see \citep{GelmanEtAl:2012, Efron:2010} for discussion.
+mortality rate (for a discussion of multiple comparisions and
+hierarchical models, see \citep{GelmanEtAl:2012, Efron:2010}).
 
@@
 default initialization is to randomly generate values uniformly on
-$[-2,2]$ by default or another interval by specification.  This
-supplies fairly diffuse starting points when transformed back to the
-constrained scale, and thus help with convergence diagnostics as
-discussed in \citep{GelmanEtAl:2013}.  Models with more data or more
-elaborate structure require narrower intervals for initialization to
-ensure the sampler is able to quickly converge to a stationary
-distribution in the high mass region of the posterior.
-
-\subsection{Variable definition and block execution summary}
-
-A table summarizing the point at which variables are read, written,
-or defined is provided in Figure~\ref{block-actions.fig}. 
-%
-\begin{figure}
-\begin{center}
-\begin{tabular}{l|c|l|l}
-{ Block} & { Statements?} & { Action} & { Evaluated}
-\\\hline\hline
-\code{user initialization} & n/a & transform & chain
-\\[3pt]
-\code{random initialization} & n/a & randomize & chain 
-\\\hline\hline
-\code{data} & no & read & chain  
-\\
-\code{transformed data} & yes & evaluate & chain  
-\\ \hline
-\code{parameters} & no & inv.\ transform, Jacobian & leapfrog  \\
-& & inv.\ transform, write & sample 
-\\[3pt]
-\code{transformed parameters} & yes & evaluate & leapfrog \\
-& & write & sample 
-\\\hline
-\code{model} & yes & evaluate & leapfrog
-\\\hline
-\code{generated quantities} & yes & evaluate & sample \\
-& & write & sample
-\end{tabular}
-\end{center}
-\caption{\it Each \proglang{Stan} program block admits certain actions that are evaluated
-  at specific times during sampling.  For example, the parameter initialization in
-  the \code{data} block requires a read operation once per chain.}
-\label{block-actions.fig}
-\end{figure}
-%
-This table is defined assuming HMC or NUTS samplers, which require a
-log probability and gradient calculation for one or more leapfrog
-steps per iteration; see \citep{HoffmanGelman:2011}.
-%
-\begin{figure}
-\begin{center}
-\begin{tabular}{l|l}
-{ Variable Categorization} & { Declaration Block}
-\\ \hline\hline
-% constants & \code{transformed data}
-% \\ \hline
-unmodeled data & \code{data}, \code{transformed data}
-\\ 
-modeled data & \code{data}, \code{transformed data}
-\\ \hline
-missing data & \code{parameters}, \code{transformed parameters}
-\\
-modeled parameters & \code{parameters}, \code{transformed parameters}
-\\
-unmodeled parameters & \code{data}, \code{transformed data}
-\\[2pt] \hline
-generated quantities & \code{transformed data}, \code{transformed parameters}, 
-\\ 
-& \code{generated quantities}
-\\ \hline\hline
-loop indices & any loop statement
-\\ \hline
-local variables & any statement block
-\\ 
-\end{tabular}
-\end{center}
-\caption{\it Variables of each categorization must be declared in
-  specific blocks.  Data may also be expressed using numeric
-  literals.}\label{variable-kinds.fig}
-\end{figure}
-%
-\citet[p.~366]{GelmanHill:2007} provide a taxonomy of the kinds of
-variables used in Bayesian models.  Figure~\ref{variable-kinds.fig}
-contains Gelman and Hill's taxonomy aligned with the corresponding
-locations of declarations and definitions in \proglang{Stan}.
-Unmodeled data variables includes size constants and regression
-predictors.  Modeled data variables include known outcomes or
-measurements.  A data variable or constant literal is modeled in a
-\proglang{Stan} program if it (or a variable that depends on it)
-occurs on the left-hand side of a sampling statement.  Unmodeled
-parameters are provided as data, and are either known or fixed to some
-value for convenience, a typical use being the parameters of a weakly
-informative prior for a parameter.
-
-A modeled parameter is given a distribution by the model (usually
-dependent on data and correlated with other parameters).  This can be
-done either by placing it (or a variable that depends on it) on the
-left-hand side of a sampling statement or by declaring the variable
-with a constraint (see Section~\ref{implicit-prior.section}).
-
-Any variable that occurs in a (transformed) data block can also be
-provided instead as a constant.  So a user may decide to write \code{y
-~ normal(0,1)} or to declare \code{mu} and \code{sigma} as data, and
-write \code{y ~ normal(mu,sigma)}.  The latter choice allows the
-parameters to be changed without recompiling the model, but requires
-them to be specified as part of the input data.
-
-Missing data is a new variable type included here.  In order to
-perform inference on missing data, it must be declared as a parameter
-and modeled; see \citep{GelmanEtAl:2013} for a discussion of
-statistical models of missing data.  Unlike \proglang{BUGS}, data may
-not contain a mixture of missing (i.e., \code{NA}) values and numeric
-values.  \proglang{Stan} uses the built-in \proglang{C++}
-floating-point (and integer) arithmetic, for which there is no
-equivalent of \code{NA}.  See \citep{Stan:2013} for strategies for
-coding missing data problems in \proglang{Stan}.
-
-Generated quantities include simple transforms of data required for
-printing.  For example, the variable of interest might be a standard
-deviation, resulting from transforming a precision parameter $\tau$ to
-a scale parameter $\sigma = \tau^{-1/2}$.  A non-linearly transformed
-variable will have different effective sample sizes and $\hat{R}$
-statistics than the variable it is derived from, so it is convenient
-to define variables of interest in the generated quantities block to
-calculate these statistics automatically.  As seen in the example in
-Figure~\ref{hier-binom.fig}, generated quantities may also be used for
-event probability estimates.
-
-The generated quantities block may also be used for forward
-simulations, generating values to make predictions or to perform
-posterior predictive checks; see \citep{GelmanEtAl:2013} for more
-information.  The generated quantities block is the only location in
-\proglang{Stan} in which random-number generators may be applied
-explicitly (they are implicit in parameters).
-
-Calculations in the generated quantities block do not impact the
-estimates of the parameters.  In this way, generated quantities can be
-used for predictions with a behavior not unlike that of cut in
-\proglang{BUGS} \citep{LunnEtAl:2012}.  Nevertheless, we recommend
-full Bayesian inference in general, and only use the generated
-quantities block for efficiency and clarity of code.  For example, in
-many models, generating predictions for unseen data or replicating
-data for posterior predictive checks does not affect the estimation of
-the parameters of interest.
-
-The list in Figure~\ref{variable-kinds.fig} is completed with two
-types of local variables, loop indices and traditional local
-variables.  Unlike \proglang{BUGS}, \proglang{Stan} allows local
-variables to be declared and assigned.  For example, it is possible to
-compute the sum of the squares of entries in an array or vector
-\code{y} as follows.%
-%
-\footnote{The built-in squared norm function is the right way to
-  implement sums of squares in \proglang{Stan}.}
-%
-\begin{Code}
-{ 
-  real sum_of_squares;
-  sum_of_squares <- 0;
-  for (n in 1:N)
-    sum_of_squares <- sum_of_squares + y[n] * y[n];
-}  
-\end{Code}
-%
-Local variables may not be declared with constraints, because there is
-no location at which it makes sense to test that they satisfy the
-constraints.
-
-\subsection{Implicit uniform priors}\label{implicit-prior.section}
-
-The default distribution for a variable is uniform over its support.
-For instance, a variable declared with a lower bound of 0 and an upper
-bound of 1 implicitly receives a $\mbox{\sf Uniform}(0,1)$
-distribution.  These implicit uniform priors are improper if the
-variable has unbounded support.  For instance, the uniform
-distributions over real values with upper and lower bounds, simplexes
-and correlation matrices is proper, but the uniform distribution over
-unconstrained or one-side constrained reals, ordered vectors or
-covariance matrices are not proper.
-
-\proglang{Stan} does not require proper priors, but if the posterior
-is improper, \proglang{Stan} will halt with an error message.%
-%
-\footnote{Improper posteriors are diagnosed automatically when
-  parameters overflow to infinity during simulation.}
-
-\subsection{The non-centered parameterization for hierarchical models}\label{centering.section}
-
-Consider the following hierarchical logistic regression model fragment.
-%
-\begin{Code}
-data {
-  int<lower=1,upper=K> group[N];  // group of data item n
-}  
-parameters {
-  ...
-  vector mu;             // mean coeff
-  real<lower=0> sigma;   // coeff scale
-  vector[K] beta;        // coeff for group k
-}
-model {
-  beta ~ normal(mu,sigma);
-  for (n in 1:N)
-    y[n] ~ bernoulli_logit(beta[group[n]] * x[n]);
-  ...
-}
-\end{Code}
-%
-With this parameterization, the \code{beta[k]} are centered around
-\code{mu}.  As a result, a change in \code{sigma} is amplified by the
-lower-level parameters and introduces a large change in density across
-the posterior.  Unfortunately the expected density variation of an HMC
-transition is limited%
-%
-\footnote{The limited density variation is ultimately a consequence of
-  a constant metric.  An additional benefit of RHMC are transitions
-  that can cover much larger variations in density, making it uniquely
-  suited to these models; see \citep{Neal:2011}.}
-%
-and many transitions are required to explore the full posterior.  The
-resulting sampler devolves into a random walk with high
-autocorrelations.  \cite{BetancourtGirolami:2013} provide a detailed
-analysis of the benefits of centered vs.\ non-centered
-parameterizations of hierarchical models in Euclidean and Riemannian
-HMC, with the conclusion that non-centered are best when data are
-sparse and centered when the data strongly identifies coefficients.
-
-The following reparameterization employs a non-centered
-parameterization for comparison.
-%
-\begin{Code}
-parameters {
-  ...
-  vector[K] beta_raw;  // replaces beta as parameter  
-}
-transformed parameters {
-  vector[K] beta;      // beta now a transformed parameter
-  beta <- mu + sigma * beta_raw;
-}
-model {
-  beta_raw ~ normal(0,1);  // beta now sampled from unit normal
-  ...
-}
-\end{Code}
-%
-This reparameterization removes the particular correlations amongst the
-hierarchical variables that would otherwise limit the effectiveness
-of the samplers.  We recommend starting with the more natural centered
-parameterization and moving to the non-centered parameterization if
-sampling of the coefficients does not mix well.%
-%
-\footnote{As the saying goes in computer science, it is easier to
-  optimize a correct program than debug an optimized program.}
-
-\section{Data types}
-
-All expressions in \proglang{Stan} are statically typed, including
-variables.  This means their type is declared at compile time as part
-of the model, and does not change throughout the execution of the
-program.  This is the same behavior as is found in compiled
-programming languages such as \proglang{C(++)}, \proglang{Fortran},
-and \proglang{Java}, but is unlike the behavior of interpreted
-languages such as \proglang{BUGS}, \proglang{R}, and
-\proglang{Python}.  Statically typing the variables (as well as
-declaring them in appropriate blocks based on usage) makes
-\proglang{Stan} programs easier to read and easier to debug by making
-explicit the modeling decisions and expression types.
-
-\subsection{Primitive types}
-
-The primitive types of \proglang{Stan} are \code{real} and \code{int},
-which are used to represent continuous and integer values.  These
-values are represented directly in \proglang{C++} as types
-\code{double} and \code{int}.  Integer expressions can be used
-anywhere a real value is required, but not {\it vice-versa}.
-
-\subsection{Vector and matrix types}
-
-\proglang{Stan} supports vectors, row vectors, and matrices with the
-usual access operations.  Indexing for vector, matrix, and array types
-starts from one.
-
-Vectors are declared with their sizes and matrices with their number
-of rows and columns.  Vector, row vector, and matrix elements are
-accessed using bracket notation, as in \code{y[3]} for the third
-element of a vector or row vector and \code{a[2,3]} for the element in
-the third column of the second row of a matrix.  Indexing begins from
-1.
-
-The notation \code{a[2]} accesses the second row of matrix \code{a}.
-Rows, columns and subblocks of matrices can be accessed using
-\code{row}, \code{col}, and \code{block} functions.  Slices of arrays
-can be accessed using the \code{head}, \code{tail}, and \code{segment}
-functions.  There are also special functions for creating a diagonal
-matrix from a vector and accessing the diagonal of a vector.
-
-All vector and matrix types contain real values and may not be
-declared to contain integers.  Collections of integers are represented
-using arrays.
-
-\subsection{Array types}
-
-An array may have entries of any other type.  For example, arrays of
-integers and reals are allowed, as are arrays of vectors or arrays of
-matrices.
-
-Higher-dimensional arrays are intrinsically arrays of arrays.  An
-entry in a two-dimensional array \code{y} may be accessed as
-\code{y[1,2]}.  The expression \code{y[1]} by itself denotes the
-one-dimensional array whose values correspond to the first row of
-\code{y}.  Thus \code{y[1][2]} has the same value as \code{y[1,2]}.%
-%
-\footnote{Arrays are stored in row-major order and matrices in
-  column-major order.}
-
-Unlike integers, which may be used where real values are required,
-arrays of integers may not be used where real arrays are required.%
-%
-\footnote{In the language of type theory, \proglang{Stan} arrays are
-  not covariant.  This follows the behavior of both arrays and
-  standard library containers in \proglang{C++}.}
-
-The manual contains a chapter discussing the efficiency tradeoffs and
-motivations for separating arrays and matrices.
-
-\subsection{Constrained variable types}
-
-Variables may be declared with constraints.  The constraints have
-different effects depending on the block in which the variable is
-declared.
-
-Integer and real types may be provided with lower bounds, upper
-bounds, or both.  This includes the types used in arrays, and the real
-types used in vectors and matrices.
-
-Vector types may be constrained to be unit simplexes (all entries
-non-negative and summing to 1), unit length vectors (sum of squares is 1), or
-ordered (entries are in ascending order), positive ordered (entries in
-ascending order, all non-negative), using the types \code{simplex[K]},
-\code{unit_vector[K]}, \code{ordered[K]}, or
-\code{positive_ordered[K]}, where \code{K} is the size of the vector.
-
-Matrices may be constrained to be covariance matrices (symmetric,
-positive definite) or correlation matrices (symmetric, positive
-definite, unit diagonal), using the types \code{cov_matrix[K]} and
-\code{corr_matrix[K]}.
-
-\section{Expressions and type inference}
-
-The syntax of \code{Stan} is defined in terms of expressions and
-statements.  Expressions denote values of a particular type.  
-Statements represent operations such as assignment and sampling as
-well as control structures such as for loops and conditionals.
-
-\subsection{Expressions}
-
-\proglang{Stan} provides the usual kinds of expressions found in
-programming languages.  This includes variables, literals denoting
-integers, real values or strings, binary and unary operators over
-expressions, and function application.
-
-\subsubsection{Type inference}
-
-The type of each variable is declared statically and cannot change.
-
-The type of a numeric literal is determined by whether or not it
-contains a period or scientific notation; for example, \code{20} has
-type \code{int} whereas \code{20.0} and \code{2e+1} have type
-\code{real}.
-
-The type of applying an operator or a function to one or more
-expressions is determined by the available signatures for the
-function.  For example, the multiplication operator (\code{*}) has a
-signature that maps two \code{int} arguments to an \code{int} and two
-\code{real} arguments to a \code{real} result.  Another signature for
-the same operator maps a \code{row\_vector} and a \code{vector} to a
-\code{real} result.
-
-\subsubsection{Type promotion}
-
-If necessary, an integer type will be promoted to a \code{real} value.
-For example, multiplying an \code{int} by a \code{real} produces a
-\code{real} result by promoting the \code{int} argument to a
-\code{real}.
-
-\section{Statements}
+$[-2,2]$; another interval may be specified with \code{init=x} for
+some non-negative floating-point value \code{x}.  This supplies fairly
+diffuse starting points when transformed back to the constrained
+scale, and thus help with convergence diagnostics as discussed in
+\citep{GelmanEtAl:2013}.  Models with more data or more elaborate
+structure require narrower intervals for initialization to ensure the
+sampler is able to quickly converge to a stationary distribution in
+the high mass region of the posterior.
+
+Although \proglang{Stan} is quite effective at converging from diffuse
+random initializations, the user may supply their own initial values
+for sampling, optimization, or diagnosis.  The top-level command-line
+option is \code{init=path}, where \code{path} is a path to a file
+specifying values for all parameters in \proglang{R} dump format.
+
+\section{Statements}\label{statements.section}
 
@@
 
-Because computation is only up to a proportion, this sampling
-statement in turn has the same effect as the direct implementation in
-terms of basic arithmetic,
+Because computation is only up to a proportionality constant (an
+additive constant on the log scale), this sampling statement in turn
+has the same effect as the direct implementation in terms of basic
+arithmetic,
 %
@@
 
-\section{Function and distribution library}
+As an example, suppose a user's program raises an error at run time
+because a covariance matrix defined in the transformed parameters
+block fails its symmetry constraint.  
+%
+\begin{Code}
+transformed parameters {
+  cov_matrix[K] Sigma;
+  for (m in 1:M)
+    for (n in m:M) 
+      Sigma[m,n] <- Omega[m,n] * sigma[m] * sigma[n];
+
+  print("Sigma=", Sigma);
+}
+\end{Code}
+%
+The print statement added at the last line will dump out the values in
+the matrix.
+
+\section{Function and distribution library}\label{functions.section}
 
@@
 functions to be efficient in computing both values and derivatives,
-they need to be vectorized so that shared computations can be reused.
+they need to operate directly on vectors of arguments so that shared
+computations can be reused.  For example, if \code{y} is a vector and
+\code{sigma} is a scalar, the logarithm of \code{sigma} need only be
+evaluated once in order to compute the normal density for every member
+of \code{y} in
+%
+\begin{Code}
+y ~ normal(mu,sigma);
+\end{Code}
 
@@
 
-Following the usual convention in mathematics, matrix and array
-indexing uses the usual square brackets (\code{[ ]}) operator, and
-begins indexing from 1.  For example, if \code{Sigma} is a matrix,
-then \code{Sigma[m,n]} is the entry at row \code{m} and column
-\code{n}.  \proglang{Stan} also allows slices of data structures to be
-returned, so that \code{Sigma[m]} is row \code{m} of the matrix
-\code{Sigma}.
+Rows, columns and subblocks of matrices can be accessed using
+\code{row}, \code{col}, and \code{block} functions.  Slices of arrays
+can be accessed using the \code{head}, \code{tail}, and \code{segment}
+functions.  There are also special functions for creating a diagonal
+matrix from a vector and accessing the diagonal of a vector.
 
@@
 and \code{mu} are vectors and \code{Sigma} is a square matrix, all of
-the same dimensionality, then \code{y - mu} is a vector, \code{(y -
-  mu)'} is a row vector, \code{(y - mu)' * Sigma} is a row vector, and
-\code{(y - mu)' * Sigma * (y - mu)} will be a real value.  Matrix
-division is provided, which is much more arithmetically stable than
-inversion, e.g., \code{(y - mu)' / Sigma} computes the same function
-as \code{(y - mu)' * inverse(Sigma)}.  \proglang{Stan} also supports
-elementwise multiplication (\code{.*}) and division (\code{./}).
+the same dimensionality, then \code{y~-~mu} is a vector,
+\code{(y~-~mu)'} is a row vector, \code{(y~-~mu)'~*~Sigma} is a row
+vector, and \code{(y~-~mu)'~*~Sigma~*~(y~-~mu)} will be a real value.
+Matrix division is provided, which is much more arithmetically stable
+than inversion, e.g., \code{(y~-~mu)'~/~Sigma} computes the same
+function as \code{(y~-~mu)'~*~inverse(Sigma)}.  \proglang{Stan} also
+supports elementwise multiplication (\code{.*}) and division
+(\code{./}).
 
@@
 parameters or constants.  Various functions are specialized for speed,
-such as quadratic products, diagonal specializations, multiply by self
-transposed, e.g., the previous example could be written as
-\code{quad\_form(Sigma, y - mu)}.
+such as quadratic products, diagonal specializations, and multiply by
+self transposed; e.g., the previous example
+\code{(y~-~mu)'~*~Sigma~*~(y~-~mu)} could be coded as
+as \code{quad\_form(Sigma,~y~-~mu)}.
 
@@
 All probability functions check that their arguments are within the
-appropriate constrained support and may be configured to throw
-exceptions or return $-\infty$ or a special not-a-number value
-(\code{NaN}) for out-of-domain arguments (the behavior of positive and
-negative infinity and not-a-number values are built into
-floating-point arithmetic).  For example,
-\code{normal\_log(y,} \code{mu,} \code{sigma)} requires the scale parameter
-\code{sigma} to be non-negative.
+appropriate constrained support and are configured to throw exceptions
+and print error messages for out-of-domain arguments (the behavior of
+positive and negative infinity and not-a-number values are built into
+floating-point arithmetic).  For example, \code{normal\_log(y,}
+\code{mu,} \code{sigma)} requires the scale parameter \code{sigma} to
+be non-negative.  Exceptions that are raised by functions will be
+caught by the sampler, optimizer or diagnostic, and their warning
+messages will be printed for the user.  Log density evaluations in
+which exceptions are raised are treated as if they had evaluated to
+negative infinity, and are thus rejected by the sampler or optimizer.
 
@@
 %
-\footnote{Both vectorization and dropping constant terms are
+\footnote{Both vector arguments and dropping constant terms are
   implemented in \proglang{C++} through template metaprograms that
@@
 
-\subsubsection{Vectorization}
-
-All of the univariate%
-%
-\footnote{We are in the process of vectorizing the multivariate
-  probability functions, but they are not all available in
-  \proglang{Stan} 2.0.  We are also in the process of vectorizing
-  all of the special functions.}
-%
-probability functions in \proglang{Stan} are vectorized so that they
-accept arrays or vectors of arguments.  For example, although the
-basic signature of the probability function
+\subsubsection{Vector arguments and shared computations}
+
+All of the univariate probability functions in \proglang{Stan} are
+implemented so that they accept arrays or vectors of arguments.  For
+example, although the basic signature of the probability function
 \code{normal\_log(y,mu,sigma)} involves real \code{y}, \code{mu} and
@@
 \code{mu} and \code{sigma} contain more than one element. A typical
-use case would be for linear regression, such as \code{y ~ normal(X
-  * beta,sigma)}, where \code{y} is a vector of observed data,
-\code{X} is a predictor matrix, \code{beta} is a coefficient vector,
-and \code{sigma} is a real value for the noise scale.
-
-The advantage of using vectorization is twofold.  First, the models
-are more concise and closer to mathematical notation.  Second, the
-vectorized versions are much faster.  They reduce the number of times
-expensive operations need to be evaluated and also reduce the number
-of virtual function calls required in the compiled \proglang{C++}
-executable for calculating gradients and higher-order derivatives.
-In the example above, \code{y ~ normal(X *
-  beta,sigma)}, the logarithm of \code{sigma} need only be computed
-once; if \code{y} is an $N$-vector, it also reduces the number of
-virtual function calls from $N$ to 1.
+use case would be for linear regression, such as \code{y ~ normal(X *
+  beta,sigma)}, where \code{y} is a vector of observed data, \code{X}
+is a predictor matrix, \code{beta} is a coefficient vector, and
+\code{sigma} is a real value for the noise scale.
+
+The advantage of using vectors is twofold.  First, the models are more
+concise and closer to mathematical notation.  Second, the vectorized
+versions are much faster.  They reduce the number of times expensive
+operations need to be evaluated and also reduce the number of virtual
+function calls required in the compiled \proglang{C++} executable for
+calculating gradients and higher-order derivatives.  For example if
+\code{sigma} is a parameter, then in evaluating {y ~
+  normal(X~*~beta,~sigma)}, the logarithm of \code{sigma} need only be
+computed once; if either \code{y} or \code{beta} is an $N$-vector, it
+also reduces the number of virtual function calls from $N$ to 1.
 
@@
 Monte Carlo (EHMC, or just HMC) \citep{DuaneEtAl:1987, Neal:1994,
-  Neal:2011} and the No-U-Turn sampler (NUTS)
+  Neal:2011} and the no-U-turn sampler (NUTS)
 \citep{HoffmanGelman:2011}.  Both the basic and NUTS versions of HMC
@@
 
-\proglang{Stan} will soon have an implementation of Riemannian
-manifold Hamiltonian Monte Carlo (RHMC)
-\citep{GirolamiCalderhead:2011}, using the SoftAbs metric to robustly
-incorporate posterior curvature into the Hamiltonian simulation
-\cite{Betancourt:2012}.  \proglang{Stan}'s MCMC framework also makes
-it straightforward to incorporate other metrics.  This implementation
-will also generalize NUTS to Riemannian manifolds
-\citep{Betancourt:2013}.  Both support adapting step sizes during
-warmup.  RHMC with SoftAbs uses first, second, and third order
-derivatives to adapt to the local curvature of the posterior.
-
-\subsubsection{Metropolis-Hastings}
-
-Random-walk Metropolis-Hastings samplers with multivariate normal
-proposals have been prototyped, but not yet integrated into the
-released versions of \proglang{Stan}.  The primary purpose of
-implementing Metropolis-Hastings is to provide baselines against which
-to measure more efficient approaches to sampling.  The covariance
-matrix used can be adapted during warmup, assuming either equal
-diagonal covariance, diagonal covariance, or a full covariance matrix.
-Metropolis-Hastings does not require any derivatives of the log
-probability function, but the lack of gradient information induces
-random walk behavior, leading to slow mixing.
-
-\subsubsection{Ensemble samplers}
-
-Two flavors of ensemble samplers are also in the works: an affine invariant
-ensemble sampler \citep{GoodmanWeare:2010} and a differential
-evolution sampler \citep{TerBraak:2006}.  Ensemble samplers do not
-require derivative information, but typically require at least as many
-chains to be run as the number of dimensions in order to have full
-rank search of the posterior.  They also require slightly different
-management of output due to a single sampler instance producing
-multiple correlated chains of draws.
-
 \subsection{Optimizers}
@@
 
-\section{Random number generation}\label{rng.section}
-
-Random number generation for \proglang{Stan} is done on a per-chain
-basis; ensemble samplers form a single joint chain over the ensemble.  By
-specifying the chain being used, the random number generator can be
-skipped ahead sufficiently to avoid replication of subsequences of
-random numbers across chains.
-
-The generated model code and underlying \proglang{C++} algorithms
-provide template parameters for a class implementing the \pkg{Boost}
-random number generator concept.
-
-By default, \proglang{Stan} uses linear congruential generators
-\citep{LEcuyer:1988}.  This generator supports efficient skip-ahead.
-
-\subsection{Replicability}
-
-The \proglang{Stan} interfaces all allow random-number generator seeds
-to be specified explicitly.  Execution uses a single base random
-number generator instance.  Therefore, by specifying a seed,
-\proglang{Stan}'s behavior is deterministic.  This is very useful for
-debugging purposes.  Seeds can be generated randomly based on
-properties of the system time, but when they are, the seed used is
-printed as part of the output to allow it to be fully replicated.
-
-\section{Library dependencies}
-
-\proglang{Stan}'s modeling language is only dependent on two external
-libraries.
-
-\subsection{Boost}
-
-\proglang{Stan} depends on several of the \pkg{Boost} \proglang{C++}
-libraries \citep{Boost:2011}.  \proglang{Stan} makes extensive use of
-\pkg{Boost}'s template metaprogramming facilities including the
-\pkg{Enable if} package, the \pkg{Type Traits} library, and the
-\pkg{Lexical Cast} library.  The \proglang{Stan} language is parsed
-using \pkg{Boost}'s \pkg{Spirit} parser, which itself depends on
-the binding libraries \pkg{Phoenix}, \pkg{Bind}, and \pkg{Lambda},
-the variant type library \pkg{Variant}, and the container library
-\pkg{Fusion}.  Exceptions are largely handled and configured through
-the error checking facilities in the \pkg{Math} and
-\pkg{Exception} packages.  Output formatting and ad-hoc input parsing
-for various formats is facilitated with the \pkg{Format} library.
-\proglang{Stan} relies heavily on the special functions defined in the
-\pkg{Math} subpackages \pkg{Special Functions} and \pkg{Statistical
-  Distributions}.  Random number generation is carried out using the
-\pkg{Random} package.  The posterior analysis framework and some
-built-in functions depend on the \pkg{Accumulators} package.
-
-\subsection{Eigen}
-
-\proglang{Stan}'s handling of matrices and linear algebra is
-implemented through the \pkg{Eigen} \proglang{C++} template library
-\citep{Eigen:2012}.  Eigen uses template metaprogramming to achieve
-state-of-the-art performance for matrix and linear algebra operations
-with a great deal of flexiblity with respect to input types.
-Unfortunately, many of the expression templates that Eigen uses for
-efficient static anaysis and lazy evaluation are short-circuited
-because of \proglang{Stan}'s need to have mixed type operations (i.e.,
-multiplying a constant predictor matrix of double values by a
-parameter vector of algorithmic differentiation values).  To make up for
-this in some important cases, \proglang{Stan} has provided compound
-functions such as the quadratic form, which allow speedups of both the
-matrix operations and their derivatives compared to a direct
-implementation using \proglang{Stan}'s built-in operators.
-
-\section{Developer process}
-
-\subsection{Version control and source repository}
-
-\proglang{Stan}'s source code is hosted on GitHub and managed using
-the \pkg{Git} version control system \citep{Chacon:2009}.  To manage
-the workflow with so many developers working at any given time, the
-project follows the GitFlow process \citep{Driessen:2010}.  All
-developer submissions are managed through pull requests and we have
-gratefully received patches from numerous sources outside the core
-development team.
-
-\subsection{Continuous integration}
-
-\proglang{Stan} uses continuous integration, meaning that the entire
-program and set of tests are run automatically as code is pushed to
-the \pkg{Git} repository.  Each pull request is tested for
-compatibility with the development branch, and the development branch
-itself is tested for stability.  \proglang{Stan} uses Jenkins
-\citep{Smart:2011}, an open-source continuous integration server.
-
-\subsection{Testing framework}
-
-\proglang{Stan} includes extensive unit tests for low-level \proglang{C++} code.
-Unit tests are implemented using the \pkg{googletest} framework
-\citep{GoogleTest:2011}.  The probability functions and command-line
-invocations are complex enough that programs are used to automatically
-generate test code for \pkg{googletest}.
-
-These unit tests evaluate every function for both appropriate values
-and appropriate derivatives.  This requires an extensive meta-testing
-framework for the probability distributions due to their high degree
-of configurability as to argument types.  The testing portion of the
-makefile also runs tests of all of the built-in models, including
-almost all of the \proglang{BUGS} sample models.  Models are tested
-for both convergence and posterior mean estimation to within MCMC
-standard error.
-
-\subsection{Builds}
-
-The build process for \proglang{Stan} is highly automated through a
-cross-platform series of \code{make} files.  The top-level makefile
-builds the \proglang{Stan}-to-\proglang{C++} translator command
-\code{bin/stanc} and posterior analysis command \code{bin/print}.  It
-also builds the library archive \code{bin/libstan.a}.  Great care was
-taken to avoid complicated platform-dependent configuration
-requirements that place a high burden on user system knowledge for
-installation.  All that is needed is a relatively recent
-\proglang{C++} compiler and version of \code{make}.
-
-As exemplified in the introduction, the makefile is automated enough
-to build an executable form of a \proglang{Stan} model in a single
-command.  All libraries and other executables will be built as a side
-effect.
-
-The top-level makefile also supplies targets to build all of the
-documentation \proglang{C++} API documentation is generated using the
-\pkg{doxygen} package \citep{Doxygen:2011}.  The \proglang{Stan}
-manual \citep{Stan:2013} is typeset using the \LaTeX\ package
-\citep{MittelbachEtAl:2004}.
-
-The makefile also has targets for all of the unit and functional
-testing, for building the source-level distribution, and for cleaning
-any temporary files that it creates.
+% \section{Random number generation}\label{rng.section}
+
+% Random number generation for \proglang{Stan} is done on a per-chain
+% basis; ensemble samplers form a single joint chain over the ensemble.  By
+% specifying the chain being used, the random number generator can be
+% skipped ahead sufficiently to avoid replication of subsequences of
+% random numbers across chains.
+
+% The generated model code and underlying \proglang{C++} algorithms
+% provide template parameters for a class implementing the \pkg{Boost}
+% random number generator concept.
+
+% By default, \proglang{Stan} uses linear congruential generators
+% \citep{LEcuyer:1988}.  This generator supports efficient skip-ahead.
+
+% \subsection{Replicability}
+
+% The \proglang{Stan} interfaces all allow random-number generator seeds
+% to be specified explicitly.  Execution uses a single base random
+% number generator instance.  Therefore, by specifying a seed,
+% \proglang{Stan}'s behavior is deterministic.  This is very useful for
+% debugging purposes.  Seeds can be generated randomly based on
+% properties of the system time, but when they are, the seed used is
+% printed as part of the output to allow it to be fully replicated.
 
@@
 
-Finally, we would like to thank all those who contributed
-documentation corrections and code patches, Jeffrey Arnold, David
-R. Blair, Eric C. Brown, Eric N. Brown, Devin Caughey, Wayne Folta,
-Andrew Hunter, Marco Inacio, Louis Luangkesorn, Jeffrey Oldham, Mike
-Ross, Terrance Savitsky, Yajuan Si, Dan Stowell, Zhenming Su, and
-Dougal Sutherland.
-
-And a special thanks to those who have contributed code for new
-features, Jeffrey Arnold, Yuanjun Gao, and Marco Inacio.
+We would like to think those who have contributed code for new
+features, Jeffrey Arnold, Yuanjun Gao, and Marco Inacio, as well as
+those who contributed documentation corrections and code patches,
+Jeffrey Arnold, David R. Blair, Eric C. Brown, Eric N. Brown, Devin
+Caughey, Wayne Folta, Andrew Hunter, Marco Inacio, Louis Luangkesorn,
+Jeffrey Oldham, Mike Ross, Terrance Savitsky, Yajuan Si, Dan Stowell,
+Zhenming Su, and Dougal Sutherland.
+
+Finally, we would like to thank two anonymous referees.
 
@@
 
+\appendix
+
+\section{Developer process}\label{process.section}
+
+\subsection{Version control and source repository}
+
+\proglang{Stan}'s source code is hosted on GitHub and managed using
+the \pkg{Git} version control system \citep{Chacon:2009}.  To manage
+the workflow with so many developers working at any given time, the
+project follows the GitFlow process \citep{Driessen:2010}.  All
+developer submissions are managed through pull requests and we have
+gratefully received patches from numerous sources outside the core
+development team.
+
+\subsection{Continuous integration}
+
+\proglang{Stan} uses continuous integration, meaning that the entire
+program and set of tests are run automatically as code is pushed to
+the \pkg{Git} repository.  Each pull request is tested for
+compatibility with the development branch, and the development branch
+itself is tested for stability.  \proglang{Stan} uses Jenkins
+\citep{Smart:2011}, an open-source continuous integration server.
+
+\subsection{Testing framework}
+
+\proglang{Stan} includes extensive unit tests for low-level \proglang{C++} code.
+Unit tests are implemented using the \pkg{googletest} framework
+\citep{GoogleTest:2011}.  The probability functions and command-line
+invocations are complex enough that programs are used to automatically
+generate test code for \pkg{googletest}.
+
+These unit tests evaluate every function for both appropriate values
+and appropriate derivatives.  This requires an extensive meta-testing
+framework for the probability distributions due to their high degree
+of configurability as to argument types.  The testing portion of the
+makefile also runs tests of all of the built-in models, including
+almost all of the \proglang{BUGS} sample models.  Models are tested
+for both convergence and posterior mean estimation to within MCMC
+standard error.
+
+\subsection{Builds}
+
+The build process for \proglang{Stan} is highly automated through a
+cross-platform series of \code{make} files.  The top-level makefile
+builds the \proglang{Stan}-to-\proglang{C++} translator command
+\code{bin/stanc} and posterior analysis command \code{bin/print}.  It
+also builds the library archive \code{bin/libstan.a}.  Great care was
+taken to avoid complicated platform-dependent configuration
+requirements that place a high burden on user system knowledge for
+installation.  All that is needed is a relatively recent
+\proglang{C++} compiler and version of \code{make}.
+
+As exemplified in the introduction, the makefile is automated enough
+to build an executable form of a \proglang{Stan} model in a single
+command.  All libraries and other executables will be built as a side
+effect.
+
+The top-level makefile also supplies targets to build all of the
+documentation \proglang{C++} API documentation is generated using the
+\pkg{doxygen} package \citep{Doxygen:2011}.  The \proglang{Stan}
+manual \citep{Stan:2013} is typeset using the \LaTeX\ package
+\citep{MittelbachEtAl:2004}.
+
+The makefile also has targets for all of the unit and functional
+testing, for building the source-level distribution, and for cleaning
+any temporary files that it creates.
+
+\section{Library dependencies}\label{libraries.section}
+
+\proglang{Stan}'s modeling language is only dependent on two external
+libraries.
+
+\subsection{Boost}
+
+\proglang{Stan} depends on several of the \pkg{Boost} \proglang{C++}
+libraries \citep{Boost:2011}.  \proglang{Stan} makes extensive use of
+\pkg{Boost}'s template metaprogramming facilities including the
+\pkg{Enable if} package, the \pkg{Type Traits} library, and the
+\pkg{Lexical Cast} library.  The \proglang{Stan} language is parsed
+using \pkg{Boost}'s \pkg{Spirit} parser, which itself depends on
+the binding libraries \pkg{Phoenix}, \pkg{Bind}, and \pkg{Lambda},
+the variant type library \pkg{Variant}, and the container library
+\pkg{Fusion}.  Exceptions are largely handled and configured through
+the error checking facilities in the \pkg{Math} and
+\pkg{Exception} packages.  Output formatting and ad-hoc input parsing
+for various formats is facilitated with the \pkg{Format} library.
+\proglang{Stan} relies heavily on the special functions defined in the
+\pkg{Math} subpackages \pkg{Special Functions} and \pkg{Statistical
+  Distributions}.  Random number generation is carried out using the
+\pkg{Random} package.  The posterior analysis framework and some
+built-in functions depend on the \pkg{Accumulators} package.
+
+\subsection{Eigen}
+
+\proglang{Stan}'s handling of matrices and linear algebra is
+implemented through the \pkg{Eigen} \proglang{C++} template library
+\citep{Eigen:2012}.  Eigen uses template metaprogramming to achieve
+state-of-the-art performance for matrix and linear algebra operations
+with a great deal of flexiblity with respect to input types.
+Unfortunately, many of the expression templates that Eigen uses for
+efficient static anaysis and lazy evaluation are short-circuited
+because of \proglang{Stan}'s need to have mixed type operations (i.e.,
+multiplying a constant predictor matrix of double values by a
+parameter vector of algorithmic differentiation values).  To make up for
+this in some important cases, \proglang{Stan} has provided compound
+functions such as the quadratic form, which allow speedups of both the
+matrix operations and their derivatives compared to a direct
+implementation using \proglang{Stan}'s built-in operators.
+
 % \nocite{WainwrightJordan:2008}

### 2014-12-01 src/docs/papers/stan-language/cut.tex
updated article for reviewer comments

\section[Why Stan?]{Why \proglang{Stan}?}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

We did not set out to build \proglang{Stan} as it currently exists.%
%
\footnote{\proglang{Stan}'s home page, \url{http://mc-stan.org/},
  links to the \proglang{Stan} manual, example models including translations of
  the \proglang{BUGS} examples and the models from \citep{GelmanHill:2007},
  getting started instructions and full documentation for \proglang{Stan}'s
  interfaces for the command line shell (\pkg{CmdStan}), \proglang{Python}
  (\pkg{PyStan}), and \proglang{R} (\pkg{RStan}), and the source code repositories
  and issue trackers.}
%
Our original goal was to apply full Bayesian inference to the sort of multilevel
generalized linear models discussed in Part~II of
\citep{GelmanHill:2007}, which are structured with grouped and
interacted predictors at multiple levels, hierarchical covariance
priors, nonconjugate coefficient priors, latent effects as in
item-response models, and varying output link functions and
distributions.

These models turned out to be a challenge for existing general purpose
inference software.  A direct encoding in \proglang{BUGS}
\citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
\proglang{JAGS}~\citep{Plummer:2003} can grind these tools to a halt.
We began by adding custom vectorized logistic regressions to
\proglang{JAGS} using \proglang{C++} to overcome the cost of
interpretation.  Although this is much faster than looping in
\proglang{JAGS}, it quickly became clear that the root of the problem
was the slow convergence of conditional sampling when parameters were
highly correlated in the posterior, as for time-series models and
hierarchical models with interacted predictors.  We finally realized
we needed a better sampler, not a more efficient implementation of
Gibbs sampling.

We briefly considered trying to tune proposals for a random-walk
Metropolis-Hastings sampler, but that seemed too problem specific and
not even necessarily possible without some kind of adaptation rather
than a global tuning of the proposals.

We were hearing more about Hamiltonian Monte Carlo (HMC) sampling,
which appeared promising but was also problematic in that the
Hamiltonian dynamics simulation requires the gradient of the log
posterior.  Although not difficult in theory for most functions,
computing these gradients by hand on a model-by-model basis is very
tedious and error prone.  That is when we discovered reverse-mode
algorithmic differentiation, which, given a templated \proglang{C++}
function for the log posterior, automatically computes its analytic
gradient up to machine precision accuracy in only a few multiples of
the time to evaluate the log probability function itself.%
%
\footnote{Hessian matrices (i.e., all second order derivatives) are
  more expensive to calculate; each row corresponds to a parameter and
  is filled by the gradient of the derivative of the log probability
  function with respect to the parameter.}
%
We explored existing algorithmic differentiation packages with open
licenses such as \pkg{RAD} \citep{Gay:2005} and its repackaging in the
Sacado module of the \pkg{Trilinos} toolkit, and the \pkg{CppAD}
package of the COIN-OR toolkit \citep{BellBurke:2008}.  Both packages,
however, supported few special functions (e.g., probability functions,
log gamma, inverse logit) or linear algebra operations (e.g., Cholesky
decompositions, matrix division), and neither are easily or modularly
extensible.

Consequently we built our own reverse-mode algorithmic differentiation package.
At that point, we ran into the problem that we could not just plug in
the probability functions from a package like \pkg{Boost} because they
weren't templated generally enough across all arguments.  Rather
than pay the price of promoting floating point values to algorithmic
differentiation variables, we wrote our own fully templated
probability functions and other special functions.

Next, we integrated the \proglang{C++} package \pkg{Eigen} for matrix
operations and linear algebra functions.  \pkg{Eigen} makes extensive
use of expression templates for lazy evaluation and the curiously
recurring template pattern (CRTP) to implement concepts without
virtual function calls; \cite{VandevoordeJosuttis:2002} provide a
complete description of template metaprogramming techniques.
Unfortunately, we ran into the same problem with Eigen as with the
existing probability libraries---it doesn't support mixed operations
of algorithmic differentiation variables and primitives like
\code{double}.  Although we initially began by promoting
floating-point vectors to algorithmic differentiation variables, we
later completely rewrote mixed-mode operations like multiplying a data
matrix by a parameter vector.  To compute derivatives of matrix
operations such as division and inverse, we implemented specialized
derivatives using the rules described by \cite{Giles:2008}.

At this point, we could fit models coded directly in \proglang{C++} on
top of the pre-release versions of the \proglang{Stan} application
programming interface (API).  Seeing how well this all worked, we set
our sights on the generality and ease of use of \proglang{BUGS} and
designed a modeling language in which statisticians could write their
models in familiar notation that could be transformed to efficient
\proglang{C++} code before being compiled into an efficient executable
program.  Although we started with the target of specifying directed
graphical models, our first implementation turned out to be a more
general imperative form of specifying log probability functions;
\proglang{BUGS} models could be translated line for line, but the
\proglang{Stan} language also supports much richer imperative
constructs like conditionals, while loops, and local variables.  This
paper is primarily about the \proglang{Stan} language and how it can
be used to specify statistical models.

The next problem we ran into when we started implementing richer
models was variables with constrained support, such as positive variables, 
simplexes, and covariance matrices.  
Efficient implementation of these constraints
required the introduction of typed variables which automatically transformed to
unconstrained support with suitable adjustments to the log probability
from the log absolute Jacobian determinant of the inverse transforms.

Even with the prototype compiler generating models, we still faced a
major hurdle for ease of use. The efficiency of HMC is very sensitive
to two tuning parameters, the discretization interval (i.e., step
size) and the total simulation time (i.e., number of steps).  The
interval size parameter could be tuned during warm-up based on
Metropolis rejection rates, but the number of steps proved difficult
to tune without sacrificing the detailed balance of the sampler.  This
led to the development of the No U-Turn (NUTS) sampler
\citep{HoffmanGelman:2011}.  Roughly speaking, NUTS builds a tree of
possible samples by randomly simulating Hamiltonian dynamics both
forwards and backwards in time until the combined trajectory turns
back on itself.  Once the trajectory has terminated, a new sample is
drawn from the tree.  In its original version, NUTS also estimates a
step size during warmup based on a target acceptance probability.

We thought we were home free, but when we measured the
speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
very disappointed.  \proglang{BUGS}'s very first example model, Rats,
ran more than an order of magnitude faster in \proglang{JAGS} than in
\proglang{Stan}.  Rats is a tough test case because the conjugate
priors and lack of posterior correlations make it an ideal candidate
for efficient Gibbs sampling.  Realizing that we were doing redundant
calculations, we wrote a vectorized form of the normal distribution
for multiple variates with the same mean and scale, which sped things
up a bit.  Performance was lacking until we finally figured out how
to both eliminate redundant calculations and partially evaluate the
gradients using a combination of expression templates and
metaprogramming.

When we attempted to fit a time-series model, we found that
normalizing the data to unit sample mean and variance sped up the fits
by an order of magnitude.  In hindsight this isn't surprising, as the
performance of the numerical simulation in HMC scales with the
variation of the parameter scales.  In \proglang{Stan} 1.0 we
introduced an adaptive diagonal metric (mass matrix) into our HMC
implementations that allowed the parameter scales to be standardized
automatically; \proglang{Stan} 2.0 added an option to estimate a dense metric
(mass matrix) and a full covariance standardization of the posterior.

Because the diagonal and dense metrics perform only a global
standardization, the performance of \proglang{Stan} in models where
the relative parameter scales varies by location, such as hierarchical
and latent models, can suffer (hierarchical modeling is covered in
Section~\ref{hierarchical-modeling.section} and reparameterizations to speed it
up in Section~\ref{centering.section}).  Riemannian manifold Hamiltonian Monte
Carlo (RMHMC) introduces a location-dependent metric that can overcome
these final hurdles \citep{GirolamiCalderhead:2011}; \proglang{Stan} has a
prototype implementation of RMHMC based on the SoftAbs metric of
\citep{Betancourt:2012}, using a generalization of NUTS to Riemannian
manifolds \citep{Betancourt:2013}.

=============================================

\subsection{Variable definition and block execution summary}

A table summarizing the point at which variables are read, written,
or defined is provided in Figure~\ref{block-actions.fig}. 
%
\begin{figure}
\begin{center}
\begin{tabular}{l|c|l|l}
{ Block} & { Statements?} & { Action} & { Evaluated}
\\\hline\hline
\code{user initialization} & n/a & transform & chain
\\[3pt]
\code{random initialization} & n/a & randomize & chain 
\\\hline\hline
\code{data} & no & read & chain  
\\
\code{transformed data} & yes & evaluate & chain  
\\ \hline
\code{parameters} & no & inv.\ transform, Jacobian & leapfrog  \\
& & inv.\ transform, write & sample 
\\[3pt]
\code{transformed parameters} & yes & evaluate & leapfrog \\
& & write & sample 
\\\hline
\code{model} & yes & evaluate & leapfrog
\\\hline
\code{generated quantities} & yes & evaluate & sample \\
& & write & sample
\end{tabular}
\end{center}
\caption{\it Each \proglang{Stan} program block admits certain actions that are evaluated
  at specific times during sampling.  For example, the parameter initialization in
  the \code{data} block requires a read operation once per chain.}
\label{block-actions.fig}
\end{figure}
%
This table is defined assuming HMC or NUTS, both of which apply the
leapfrog integrator to simulate the Hamiltonian dynamics
\citep{HoffmanGelman:2014,Neal:2011} of a particle representing a
parameter.  Each leapfrog step requires an evaluation of the log
density and its gradient.  NUTS automatically adapts the integration
time (and hence number of leapfrog steps) each iteration and uses slice
sampling for improved acceptance, whereas basic HMC sets integration
time statically in the command-line parameters using
\code{engine=static int\_time=t}, with \code{t} set to some positive
floating point value.
%
\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Variable Categorization} & { Declaration Block}
\\ \hline\hline
% constants & \code{transformed data}
% \\ \hline
unmodeled data & \code{data}, \code{transformed data}
\\ 
modeled data & \code{data}, \code{transformed data}
\\ \hline
missing data & \code{parameters}, \code{transformed parameters}
\\
modeled parameters & \code{parameters}, \code{transformed parameters}
\\
unmodeled parameters & \code{data}, \code{transformed data}
\\[2pt] \hline
generated quantities & \code{transformed data}, \code{transformed parameters}, 
\\ 
& \code{generated quantities}
\\ \hline\hline
loop indices & any loop statement
\\ \hline
local variables & any statement block
\\ 
\end{tabular}
\end{center}
\caption{\it Variables of each categorization must be declared in
  specific blocks.  Data may also be expressed using numeric
  literals.}\label{variable-kinds.fig}
\end{figure}
%
\citet[p.~366]{GelmanHill:2007} provide a taxonomy of the kinds of
variables used in Bayesian models.  Figure~\ref{variable-kinds.fig}
contains Gelman and Hill's taxonomy aligned with the corresponding
locations of declarations and definitions in \proglang{Stan}.
Unmodeled data variables includes size constants and regression
predictors.  Modeled data variables include known outcomes or
measurements.  A data variable or constant literal is modeled in a
\proglang{Stan} program if it (or a variable that depends on it)
occurs on the left-hand side of a sampling statement.  Unmodeled
parameters are provided as data, and are either known or fixed to some
value for convenience, a typical use being the parameters of a weakly
informative prior for a parameter.

A modeled parameter is given a distribution by the model (usually
dependent on data and correlated with other parameters).  This can be
done either by placing it (or a variable that depends on it) on the
left-hand side of a sampling statement or by declaring the variable
with a constraint (see Section~\ref{implicit-prior.section}).

Any variable that occurs in a (transformed) data block can also be
provided instead as a constant.  So a user may decide to write \code{y
~ normal(0,1)} or to declare \code{mu} and \code{sigma} as data, and
write \code{y ~ normal(mu,sigma)}.  The latter choice allows the
parameters to be changed without recompiling the model, but requires
them to be specified as part of the input data.

Missing data is a new variable type included here.  In order to
perform inference on missing data, it must be declared as a parameter
and modeled; see \citep{GelmanEtAl:2013} for a discussion of
statistical models of missing data.  Unlike \proglang{BUGS}, data may
not contain a mixture of missing (i.e., \code{NA}) values and numeric
values.  \proglang{Stan} uses the built-in \proglang{C++}
floating-point (and integer) arithmetic, for which there is no
equivalent of \code{NA}.  See \citep{Stan:2013} for strategies for
coding missing data problems in \proglang{Stan}.

Generated quantities include simple transforms of data required for
printing.  For example, the variable of interest might be a standard
deviation, resulting from transforming a precision parameter $\tau$ to
a scale parameter $\sigma = \tau^{-1/2}$.  A non-linearly transformed
variable will have different effective sample sizes and $\hat{R}$
statistics than the variable it is derived from, so it is convenient
to define variables of interest in the generated quantities block to
calculate these statistics automatically.  As seen in the example in
Figure~\ref{hier-binom.fig}, generated quantities may also be used for
event probability estimates.

The generated quantities block may also be used for forward
simulations, generating values to make predictions or to perform
posterior predictive checks; see \citep{GelmanEtAl:2013} for more
information.  The generated quantities block is the only location in
\proglang{Stan} in which random-number generators may be applied
explicitly (they are implicit in parameters).

Calculations in the generated quantities block do not impact the
estimates of the parameters.  In this way, generated quantities can be
used for predictions with a behavior not unlike that of cut in
\proglang{BUGS} \citep{LunnEtAl:2012}.  Nevertheless, we recommend
full Bayesian inference in general, and only use the generated
quantities block for efficiency and clarity of code.  For example, in
many models, generating predictions for unseen data or replicating
data for posterior predictive checks does not affect the estimation of
the parameters of interest.

The list in Figure~\ref{variable-kinds.fig} is completed with two
types of local variables, loop indices and traditional local
variables.  Unlike \proglang{BUGS}, \proglang{Stan} allows local
variables to be declared and assigned.  For example, it is possible to
compute the sum of the squares of entries in an array or vector
\code{y} as follows.%
%
\footnote{The built-in squared norm function is the right way to
  implement sums of squares in \proglang{Stan}.}
%
\begin{Code}
{ 
  real sum_of_squares;
  sum_of_squares <- 0;
  for (n in 1:N)
    sum_of_squares <- sum_of_squares + y[n] * y[n];
}  
\end{Code}
%
Local variables may not be declared with constraints, because there is
no location at which it makes sense to test that they satisfy the
constraints.

===================================================

\subsection{The non-centered parameterization for hierarchical models}\label{centering.section}

Consider the following hierarchical logistic regression model fragment.
%
\begin{Code}
data {
  int<lower=1,upper=K> group[N];  // group of data item n
}  
parameters {
  ...
  vector mu;             // mean coeff
  real<lower=0> sigma;   // coeff scale
  vector[K] beta;        // coeff for group k
}
model {
  beta ~ normal(mu,sigma);
  for (n in 1:N)
    y[n] ~ bernoulli_logit(beta[group[n]] * x[n]);
  ...
}
\end{Code}
%
With this parameterization, the \code{beta[k]} are centered around
\code{mu}.  As a result, a change in \code{sigma} is amplified by the
lower-level parameters and introduces a large change in density across
the posterior.  Unfortunately the expected density variation of an HMC
transition is limited%
%
\footnote{The limited density variation is ultimately a consequence of
  a constant metric.  An additional benefit of RMHMC are transitions
  that can cover much larger variations in density, making it uniquely
  suited to these models; see \citep{Neal:2011}.}
%
and many transitions are required to explore the full posterior.  The
resulting sampler devolves into a random walk with high
autocorrelations.  \cite{BetancourtGirolami:2013} provide a detailed
analysis of the benefits of centered vs.\ non-centered
parameterizations of hierarchical models in Euclidean and Riemannian
HMC, with the conclusion that non-centered are best when data are
sparse and centered when the data strongly identifies coefficients.

The following reparameterization employs a non-centered
parameterization for comparison.
%
\begin{Code}
parameters {
  ...
  vector[K] beta_raw;  // replaces beta as parameter  
}
transformed parameters {
  vector[K] beta;      // beta now a transformed parameter
  beta <- mu + sigma * beta_raw;
}
model {
  beta_raw ~ normal(0,1);  // beta now sampled from unit normal
  ...
}
\end{Code}
%
This reparameterization removes the particular correlations amongst the
hierarchical variables that would otherwise limit the effectiveness
of the samplers.  We recommend starting with the more natural centered
parameterization and moving to the non-centered parameterization if
sampling of the coefficients does not mix well.%
%
\footnote{As the saying goes in computer science, it is easier to
  optimize a correct program than debug an optimized program.}

=============================================

\subsubsection{Metropolis-Hastings}

Random-walk Metropolis-Hastings samplers with multivariate normal
proposals have been prototyped, but not yet integrated into the
released versions of \proglang{Stan}.  The primary purpose of
implementing Metropolis-Hastings is to provide baselines against which
to measure more efficient approaches to sampling.  The covariance
matrix used can be adapted during warmup, assuming either equal
diagonal covariance, diagonal covariance, or a full covariance matrix.
Metropolis-Hastings does not require any derivatives of the log
probability function, but the lack of gradient information induces
random walk behavior, leading to slow mixing.

\subsubsection{Ensemble samplers}

Two flavors of ensemble samplers are also in the works: an affine invariant
ensemble sampler \citep{GoodmanWeare:2010} and a differential
evolution sampler \citep{TerBraak:2006}.  Ensemble samplers do not
require derivative information, but typically require at least as many
chains to be run as the number of dimensions in order to have full
rank search of the posterior.  They also require slightly different
management of output due to a single sampler instance producing
multiple correlated chains of draws.

\proglang{Stan} will soon have an implementation of Riemannian
manifold Hamiltonian Monte Carlo (RMHMC)
\citep{GirolamiCalderhead:2011}, using the SoftAbs metric to robustly
incorporate posterior curvature into the Hamiltonian simulation
\cite{Betancourt:2012}.  \proglang{Stan}'s MCMC framework also makes
it straightforward to incorporate other metrics.  This implementation
will also generalize NUTS to Riemannian manifolds
\citep{Betancourt:2013}.  Both support adapting step sizes during
warmup.  RMHMC with SoftAbs uses first, second, and third order
derivatives to adapt to the local curvature of the posterior.
